---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Density estimation {#density}

This chapter is about nonparametric density estimation. A classical
nonparametric estimator of a density is the histogram, which provides
discontinuous and piecewise constant estimates. In this chapter, however, we
instead focus on some of the alternatives that provide continuous or even
smooth estimates.

*Kernel methods* form an important class of smooth density estimators and are
implemented in the R function `density()`. These estimators are essentially 
just locally weighted averages and their computation is relatively 
straightforward in theory. In practice, however, different approaches to implementing
the computations can lead to considerable variations in actual runtime.

The implementation of kernel density estimators will illustrate three points:

* If possible, choose vectorized implementations in R.
* If a small loss in accuracy is acceptable, an approximate solution 
  can be orders of magnitudes faster than a literal implementation.
* The time it takes to numerically evaluate different 
  [elementary functions](https://en.wikipedia.org/wiki/Elementary_function)
  may depend strongly on the function and how you implement the computation.

The first point is emphasized because it results in implementations that 
are short, expressive and easier to understand just as much as it typically results 
in computationally more efficient implementations. Note also that not every
computation can be vectorized in a beneficial way, and one should never go 
through hoops to vectorize a computation. 

Kernel methods rely on one or more *regularization parameters* that must be 
selected to achieve the right balance of adapting 
to data without adapting too much to the random variation in the data. 
Choosing the right amount of regularization is just as important as choosing 
the method to use in the first place. It may, in fact, be more important. 
We actually do not have a complete implementation of a nonparametric estimator 
until we have implemented a data driven and automatic way of choosing the 
amount of regularization. Implementing only the computations for 
evaluating a kernel estimator, say, and leaving it completely 
to the user to choose the bandwidth is a job half done. Methods and implementations
for choosing the bandwidth are therefore treated in some detail in this chapter.

In the final section a likelihood analysis is carried out. 
This is done to further clarify why regularized estimators are needed 
to avoid overfitting to the data, and why there is in general no nonparametric 
maximum likelihood estimator of a density. Regularization of the likelihood 
can be achieved by constraining the density estimates to belong to a family 
of increasingly flexible parametric densities that are fitted to data. This is 
known as the *method of sieves*. Another approach is based on basis expansions,
but in either case, automatic selection of the amount of regularization is just 
as important as for kernel methods. 

Section \@ref(bandwidth) deals with choosing the bandwidth 
parameter appropriately. In Section \@ref(likelihood) we see how likelihood
methods can be combined with the use of flexible parametric families of 
densities. Section \@ref(cv) introduces cross-validation as a general method 
for choosing tuning parameters, such as the bandwidth parameter or parameters 
that control the complexity of a parametrized model.

## Kernel methods {#kernel-density}

Recall the data on [$\phi$- and $\psi$-angles](#intro-angles) in 
polypeptide backbone structures, as considered in Section \@ref(intro-angles).
Figure \@ref(fig:angleDens2) shows histograms of the data.
We will in this chapter treat methods for density 
estimation for univariate data such as data on either the $\phi$- or the 
$\psi$-angle.

(ref:angleDenscap2) Histograms equipped with a rug plot of the distribution of $\phi$-angles (left) and $\psi$-angles (right) of the peptide planes in the protein human protein 1HMP. 

<!-- [1HMP](https://www.rcsb.org/structure/1HMP). -->

```{r angleDens2, fig.cap="(ref:angleDenscap2)", fig.show="hold", out.width="49%", fig.pos="h", echo=FALSE}
hist(angle$phi, prob = TRUE)
rug(angle$phi)
hist(angle$psi, prob = TRUE)
rug(angle$psi)
```

We let $f_0$ denote the unknown density that we want to estimate. 
That is, we imagine that the data points $x_1, \ldots, x_N$ are 
all observations drawn from the probability measure with 
density $f_0$ w.r.t. Lebesgue measure on $\mathbb{R}$. 

In nonparametric density estimating we generally want to estimate the target
density, $f_0$, without assuming that it belongs to a particular parametrized
family of densities. In this section we develop kernel density
estimation, whose usage via the `density()` function was illustrated in Section
\@ref(intro-smooth). 

A starting point for nonparametric density estimation is the approximation

\[
\P(X \in (x-h, x+h)) = \int_{x-h}^{x+h} f_0(z) \ dz \approx f_0(x) 2h,
\]

which is valid for any continuous density $f_0$. Inverting this approximation
and using the law of large numbers, 

\begin{align*}
  f_0(x) & \approx \frac{1}{2h}\P(X \in (x-h, x+h)) \\
         & \approx \frac{1}{2hN} \sum_{j=1}^N 1_{(x-h, x+h)}(x_j) \\
         & = \underbrace{\frac{1}{2hN} \sum_{j=1}^N 1_{(-h, h)}(x - x_j)}_{\hat{f}_h(x)}
\end{align*}

for i.i.d. observations $x_1, \ldots, x_N$ from the distribution with density 
$f_0$ w.r.t. Lebesgue measure. 
The function $\hat{f}_h$ as defined above is an example of a kernel 
density estimator using the *uniform* or *rectangular kernel*\index{rectangular kernel}

\[
K(x) = \frac{1}{2} 1_{(-1, 1)}(x)
\]

and bandwidth\index{bandwidth} $h > 0$. The general *kernel estimator*,
\index{kernel estimator} is defined as

\begin{equation}
  \hat{f}_h(x) = \frac{1}{hN} \sum_{j=1}^N K\left(\frac{x - x_j}{h}\right)
  (\#eq:kernel-def)
\end{equation}

for a kernel\index{kernel}
$K : \mathbb{R} \to \mathbb{R}$ and a bandwidth parameter $h > 0$. 

As an illustration of how kernel density estimation works, we consider an
example where we have five observations: -1, 3, 5, 6, and 9, and want to
compute a kernel density estimate.

```{r}
#| echo: false
#| fig-width: 4
#| fig-cap: An illustrative example of a kernel density estimate using
#|   the Gaussian kernel. The observations are denoted by the rug plot
#|   marks at the bottom of the plot. For each of these, we add a
#|   Gaussian kernel centered at that point. Then, the
#|   final estimate is obtained by summing (averaging) all of the
#|   kernels.
x <- c(-1, 3, 5, 6, 9)
n <- length(x)

l <- lapply(x, function(xi) {
  a <- seq(-4, 4, length.out = 100)
  d <- dnorm(a) / n
  data.frame(
    x = xi,
    a = a + xi,
    d = d
  )
})

res <- do.call(rbind, l)

p1 <- ggplot() +
  xlim(c(-5, 15)) +
  ylim(c(0, 0.15)) +
  labs(x = "x", y = "Density")

p2 <- p1 +
  geom_rug(aes(x), data = tibble(x = x), linewidth = 1, col = "navy")

p3 <- p2 +
  geom_line(aes(a, d, group = x), col = "dark orange", data = res)

p4 <- p3 +
  geom_density(aes(x), bw = 1, data = tibble(x = x))

p4
```

We often use kernels $K$ that are non-negative and normalized so 
that $\int K(x) \mathrm{d}x = 1$, in which case $K$ is a probability 
density function. A widely used kernel is the *Gaussian kernel*\index{Gaussian kernel}

\[
  K(x) = \frac{1}{\sqrt{2\pi}} e^{-\frac{x^2}{2}}.
\]

One direct benefit of considering other kernels than the rectangular is that 
$\hat{f}_h$ inherits all smoothness properties from $K$. Whereas the rectangular 
kernel is not even continuous, the Gaussian kernel is $C^{\infty}$ and so is 
the resulting kernel density estimate. 
Note also that when $K$ is a probability density function, 
$\hat{f}_h$ is likewise a probability density function.
Kernel density estimation is 
sometimes referred to as the Parzen or Parzen–Rosenblatt window 
method with reference to their seminal contributions 
[@Rosenblatt:1956; @Parzen:1962].

Two natural questions arise when we want to use kernel density estimation 
in practice. How should we choose the kernel and how should we choose the 
bandwidth? We discuss below some qualitative effects of these choices on the 
estimate. In practice the kernel is chosen independently of data 
based on either computational, statistical or mathematical properties of 
the kernel---or maybe just based on convenience and availability. 
The bandwidth is, on the contrary, chosen in a data dependent way, typically 
to optimize some performance criterion of the estimator. The bandwidth 
is an example of a *tuning parameter*\index{tuning parameter}, 
which has to be adjusted appropriately and in a data dependent way for 
optimal performance of the estimator. 

For the rectangular kernel it is fairly easy to understand how the 
choice of bandwidth qualitatively affects the kernel estimator.
We see directly from the definition that $\hat{f}_h(x)$ is an average over the
window $(x-h, x+h)$, and as we move this window over the data points by moving
$x$, $\hat{f}_h(x)$ will jump up or down depending upon whether a data point is
entering or exiting the window. Each jump will be of size $(2hN)^{-1}$ (when all
data points are distinct). For large $h$, the window is large, the jumps are
small and $\hat{f}_h$ will be flat and close to a constant. If $h$ is small, the
window is small and $\hat{f}_h$ will make larger jumps and be more wiggly. A too
large $h$ will *oversmooth*\index{oversmooth} the data relative to $f_0$ by
effectively ignoring the data, while a too small $h$ will
*undersmooth*\index{undersmooth} the data relative to $f_0$ by allowing individual
data points to have large local effects.

The effect of the kernel on the estimator is less obvious. While the 
bandwidth has a global effect on the density estimate, the kernel's effect 
is more local. We noted the most obvious local effect above; that 
$\hat{f}_h(x)$ inherits smoothness properties from $K$. If we consider 
the rectangular kernel, we may also note that all data points in $(x-h, x+h)$
count equally much. For estimating the density in $x$ it seems natural that 
data points close to $x \pm h$ should count less than points close to $x$. 
The rectangular kernel makes a sharp cut at $x \pm h$; either a data point counts or it 
does not. Other (smoother) kernels more gradually downweight points further 
away from $x$.

Figure \@ref(fig:psiDens) illustrates how the choice of bandwidth and kernel
affects the resulting kernel density estimate. The choice of kernel has
relatively minor effects in practice, and we will in this chapter mostly use the
Gaussian kernel, but see Exercise \@ref(exr:epanechnikov) for an alternative. The
choice of bandwidth affects the estimate a lot, and Section \@ref(bandwidth) on
[bandwidth selection](#bandwidth) pursues the theory behind automatic and data
driven bandwidth selection in more detail. Before going into the details 
of optimal bandwidth selection we will in the next section treat implementations and 
computational aspects of kernel density estimation. 

### Implementation {#dens-implement}

What does an implementation of a kernel density estimator entail? What 
should actually be computed? The definition just specifies how to evaluate 
$\hat{f}_h$ in any given point $x$, but there is really nothing to compute
until we need to evaluate $\hat{f}_h.$ There are two different answers to the 
question of what to implement. Either we implement $\hat{f}_h$ as a function 
in R, which can compute and return evaluations of $\hat{f}_h(x)$ for different 
$x$-s. Or we implement an algorithm
that evaluates and returns the values of the density estimate in a finite number 
of points. We will pursue the latter type of implementation, which is 
similar to how `density()` is implemented. The former type of implementation 
is discussed in Section \@ref(likelihood) and Appendix \@ref(vectorization).

Our first implementation, `kern_dens_loop()`, is a fairly low-level
implementation that returns the evaluation of the density estimate in a number
of equidistant points. The function mimics some of the defaults of `density()`
so that it actually evaluates the estimate in the same points as `density()`.

```{r kern_dens_loop}
kern_dens_loop <- function(x, h, n = 512) {
  rg <- range(x)
  # xx is equivalent to grid points in 'density()'
  xx <- seq(rg[1] - 3 * h, rg[2] + 3 * h, length.out = n)
  y <- numeric(n)
  # The computation is done using nested for-loops. The outer loop
  # is over the grid points, and the inner loop is over the data points.
  for (i in seq_along(xx)) {
    for (j in seq_along(x)) {
      y[i] <- y[i] + exp(- (xx[i] - x[j])^2 / (2 * h^2))
    }
  }
  y <- y / (sqrt(2 * pi) * h * length(x))
  list(x = xx, y = y)
}
```


The function `kern_dens_loop()` has three formal arguments: 
`x` is the numeric vector of data points, `h` is the bandwidth 
and `n` is the number of grid points. The default value of `n`
is 512 matching the default of `density()`. 

Note that `kern_dens_loop()` returns a list containing the grid points (`x`) where 
the density estimate is evaluated as well as the estimated density 
evaluations (`y`). 

We test if the implementation works as expected---in this 
case by comparing it to `density()`, which is thus used as a reference 
implementation. 

```{r kern-dens-fig, dependson=c("kern-dens"), fig.cap = "Kernel density estimates with the Gaussian kernel (left) using R's implementation (black) and our implementation (red) together with differences of the estimates (right).", fig.show='hold', out.width="49%"}
#| fig-width: 3
library(CSwR) # for the 'angle' dataset

f_hat <- kern_dens_loop(angle$psi, 0.2)
f_hat_dens <- density(angle$psi, 0.2)

hist(angle$psi, prob = TRUE, border = NA)
lines(f_hat, lwd = 4, ylab = "Density")
lines(f_hat_dens, col = "red", lwd = 2)

plot(
  f_hat$x, f_hat$y - f_hat_dens$y,
  type = "l",
  lwd = 2,
  xlab = "psi",
  ylab = "Difference"
)
```

Figure \@ref(fig:kern-dens-fig) suggests that the estimates computed by our 
implementation and by `density()` are the same when we just visually compare the
plotted densities. However, if we look at the differences instead, 
we see that the differences can be as large as $4 \times 10^{-4}$ in absolute value. This is 
way above what we should expect from rounding errors alone when using 
double precision arithmetic. Thus the two implementations only compute 
*approximately* the same. The reason for these differences is that 
`density()` actually relies on certain approximations for runtime 
efficiency.

In R we can often beneficially implement computations in a vectorized way 
instead of using an explicit loop. It is fairly easy to change the implementation 
to be more vectorized by computing each evaluation in one single line using 
the `sum()` function and the fact that `exp()` and squaring are vectorized.

```{r kern-dens-vec}
kern_dens_vec <- function(x, h, n = 512) {
  rg <- range(x)
  xx <- seq(rg[1] - 3 * h, rg[2] + 3 * h, length.out = n)
  y <- numeric(n)
  # The inner loop from 'kern_dens_loop()' has been vectorized, and 
  # only the outer loop over the grid points remains. 
  const <- (sqrt(2 * pi) * h * length(x))
  for (i in seq_along(xx)) {
    y[i] <- sum(exp(-(xx[i] - x)^2 / (2 * h^2))) / const
  }
  list(x = xx, y = y)
}
```

We test this new implementation by comparing it to our previous 
implementation.

```{r kern-dens-vec-test, dependson=c( "kern-dens-vec")}
range(kern_dens_loop(angle$psi, 0.2)$y - kern_dens_vec(angle$psi, 0.2)$y)
```

The magnitude of the differences are of order at most $10^{-16}$, which is what 
can be expected due to rounding errors. Thus we conclude that up to rounding 
errors, `kern_dens_loop()` and `kern_dens_vec()` return the same on this dataset. This is,
of course, not a comprehensive test, but it is an example of one among a 
number of tests that should be considered.

There are several ways to get completely rid of the explicit loops and write 
an entirely vectorized implementation in R. One of the solutions will use the 
`vapply()` function, which belongs to the family of `*apply()` functions
in base R that apply a function to each element in a vector or a list. In the 
parlance of functional programming the `*apply()` functions are variations 
of the functional, or higher-order-function, known as [map](https://en.wikipedia.org/wiki/Map_(higher-order_function)).

```{r kern-dens-apply}
kern_dens_apply <- function(x, h, n = 512) {
  rg <- range(x)
  xx <- seq(rg[1] - 3 * h, rg[2] + 3 * h, length.out = n)
  const <- sqrt(2 * pi) * h * length(x)
  kern <- function(z) sum(exp(-(z - x)^2 / (2 * h^2))) / const
  y <- vapply(xx, kern, FUN.VALUE = double(1))
  list(x = xx, y = y)
}
```

The `vapply()` call\index{vapply} applies the function `kern()`, 
defined in the line before, to every element in the vector `xx`. The 
`FUN.VALUE = double(1)` specifies that the return value of `kern()` 
is a double and that `vapply()` should return a vector of doubles. 
An alternative to `vapply()` is `lapply()`, which always returns a list, 
or `sapply()`, which is a simple wrapper around `lapply()` that attempts 
to "simplify" the result from a list to an array. For programming, 
it is safest to use `vapply()` or `lapply()` since these functions always 
return an object of the same type, while `sapply()` can return a list as 
well as a vector or an array depending on the circumstances. Going beyond 
base R, the [purrr package offers a different implementation of maps and  
functionals](https://adv-r.hadley.nz/functionals.html).

An alternative, and also completely vectorized, solution is based on 
the functions `outer()` and `rowMeans()`.

```{r kern-dens-outer}
kern_dens_outer <- function(x, h, n = 512) {
  rg <- range(x)
  xx <- seq(rg[1] - 3 * h, rg[2] + 3 * h, length.out = n)
  y <- outer(xx, x, function(zz, z) exp(- (zz - z)^2 / (2 * h^2)))
  y <- rowMeans(y) / (sqrt(2 * pi) * h)
  list(x = xx, y = y)
}
```

The `outer()` function evaluates the kernel in all combinations of the grid and
data points and returns a matrix of dimensions $N \times N$. The function
`rowMeans()` computes the means of each row and returns a vector of length $N$.
Note how the third argument to `outer()` is an *anonymous function*. That is, it
is a function that does not get a name and exists only during the evaluation of
the `outer()` call. It is immaterial for the result and efficiency of the
computation whether a function gets a name, like `kern()` in the implementation
of `kern_dens_apply()`, or if it is anonymous as in the implementation of
`kern_dens_outer()`. But it does affect the readability of the code. It is fine 
to use simple anonymous functions, but more complicated functions spanning multiple 
lines of code should preferably be separated out and given a name. 

We should, of course, remember to test the two last implementations, e.g., 
by comparing with `kern_dens_loop()`. 

```{r kern-dens-outer-test, dependson=c( "kern-dens", "kern-dens-vec", "kern_dens_loop-apply", "kern_dens_loop-outer")}
range(kern_dens_loop(angle$psi, 0.2)$y - kern_dens_apply(angle$psi, 0.2)$y)
range(kern_dens_loop(angle$psi, 0.2)$y - kern_dens_outer(angle$psi, 0.2)$y)
```

The natural question is how to choose between the different implementations?
Besides being correct it is important that the code is easy to read 
and understand. Which of the four implementations above that is best in 
this respect may depend a lot on the background of the reader.
All four implementations are arguably quite readable, but
`kern_dens_loop()` with the double loop might appeal a bit more to people 
with a background in imperative programming, while `kern_dens_apply()` might 
appeal more to people with a preference for functional programming. 
This functional and vectorized solution is also 
a bit closer to the mathematical notation with, 
e.g., the sum sign $\Sigma$ being mapped directly to the `sum()` function
instead of the incremental addition in the for-loop. For 
these specific implementations the differences are mostly
aesthetic nuances and preferences may be more subjective than substantial. 

To make a qualified choice between the implementations we should
investigate if they differ in terms of runtime and memory consumption.

### Benchmarking

Benchmarking is about measuring and comparing performance. For software
this often means measuring runtime and memory usage, though there are clearly 
many other aspects of software that should be benchmarked in general. This 
includes user experience, energy consumption and implementation and maintenance 
time. In this section we focus on benchmarking runtime and memory usage.

Base R provides a simple way of measuring runtime via the function `system.time()`.
Throughout this book we will use the tools from the bench package to benchmark 
code, see also Appendix \@ref(performance). It includes the `bench_time()` 
function as an alternative to `system.time()` for measuring 
the time it takes to evaluate a single expression. 

```{r kern-dens-time, echo=c(7, 9, 11, 13), dependson=c( "kern-dens", "kern-dens-vec", "kern-dens-apply", "kern-dens-outer"), results='hold'}
# Evaluations to trigger JIT compilation
invisible(kern_dens_loop(angle$psi, 0.2))
invisible(kern_dens_vec(angle$psi, 0.2))
invisible(kern_dens_apply(angle$psi, 0.2))
invisible(kern_dens_outer(angle$psi, 0.2))
cat("kern_dens_loop:\n")
bench::bench_time(kern_dens_loop(angle$psi, 0.2))
cat("kern_dens_vec:\n")
bench::bench_time(kern_dens_vec(angle$psi, 0.2))
cat("kern_dens_apply:\n")
bench::bench_time(kern_dens_apply(angle$psi, 0.2))
cat("kern_dens_outer:\n")
bench::bench_time(kern_dens_outer(angle$psi, 0.2))
```

The bench package does high precision benchmarking and offers a range of
convenient tools, including good default formatting options so that all timings
are printed in a human readable way with correct units. The 
`bench_time()` function prints out two measurements. 
The "real" time is the total runtime, also known as the "wallclock time", 
while the "process" time is the time that the CPU actually spent on the 
expression evaluation. Curiously, for fast evaluations as above, the 
wallclock time can be smaller than the process time if the two are not 
measured with the same precision (and this is the case on macOS).

From this simple benchmark, `kern_dens_loop()` is clearly substantially slower
than the three other implementations. For more systematic benchmarking of 
runtime we use the `mark()` function from the bench package, which evaluates 
expressions repeatedly to give a more complete picture of the distribution 
of runtime and memory usage, see also Appendix \@ref(performance).

```{r kern-bench, echo=2, dependson=c( "kern-dens", "kern-dens-vec", "kern-dens-apply", "kern-dens-outer")}
set.seed(2345)
kern_bench <- bench::mark(
  loop = kern_dens_loop(angle$psi, 0.2),
  vec = kern_dens_vec(angle$psi, 0.2),
  apply = kern_dens_apply(angle$psi, 0.2),
  outer = kern_dens_outer(angle$psi, 0.2)
)
```

Note that `mark()` by default checks that the results from all benchmarked
expressions are (nearly) equal by `all.equal()`, thus the above benchmarking
also implicitly tests consistency of the implementations. The result stored in
`kern_bench` is a tabular data structure with one row for each expression
benchmarked and a good default print method for showing summary statistics 
from the benchmark. 

```{r kern-bench-hidden, echo=FALSE, dependson="kern-bench"}
old_options <- options(digits = 3)
kern_bench
options(digits = old_options$digits)
```

The median runtimes are similar to the measurements done using 
`bench_time()`. The implementation of the kernel density estimator
using a double loop takes 15--16 ms, while 
the three vectorized implementations take about 1.3 ms and are thus more than a 
factor 10 faster. The memory allocation is, however, larger for all vectorized 
implementations and largest for the implementation using `outer()`. 

To visualize the complete benchmark data behind the summary statistics we
can use the plot method for benchmark data. It is based on ggplot2 and thus 
easy to modify, and we change a few defaults here for a better presentation 
of the results.

```{r kern-bench-autoplot, dependson="kern-bench-hidden", message=FALSE}
#| fig-cap: Benchmarks of four different implementations of
#|   kernel density estimation, using either a double loop, vectorization of
#|   the inner loop, the `apply()` family of functions or the `outer()` function.
library(bench)
plot(kern_bench) +
    scale_y_discrete(limits = c("outer", "apply", "vec", "loop")) +
    bench::scale_x_bench_time(breaks = c(1, 3, 10, 30, 100) * 10^{-3})
```

The more refined benchmark study using `mark()` (\@ref(fig:kern-bench-fig))
does not change our initial impression from using `bench_time()` substantially.
The function `kern_dens_loop()` is notably slower than the three vectorized
implementations, but we are now able to more clearly see the minor differences
among them. It appears that `kern_dens_apply()` and `kern_dens_outer()` have
somewhat similar runtime distributions and that `kern_dens_vec()` is marginally
faster than both of these. Other minor differences in these empirical runtime
distributions change if the experiment is repeated. 

Note that the observations in Figure \@ref(fig:kern-bench-fig) are colored
according to their level of garbage collection ("gc" in the legend). Garbage
collection is an automatic memory management tool used in R (and many other
programming languages) that intermittently frees up memory that is no longer in
use. Garbage collection is necessary to mitigate eventual exhaustion of all
your computer's available memory, yet, as you can see from the figure, it comes
with a cost in the form of increased runtime. And although it is impossible to
avoid completely, you can sometimes reduce its impact by writing more
memory-efficient code and avoiding short-lived intermediate objects. We will
see later, when we introduce C++ integration via Rcpp, that this problem can be
automatically alleviated by relying on the built-in optimizations of compiled
code.

### Runtime complexity

In many cases when we benchmark runtime it is of interest 
to investigate how runtime depends on various parameters. This is so 
for kernel density estimation, where we want to understand how changes in 
the number of data points, $N$, and the number of grid points, $n$, affect
runtime. The function `press()` from the bench package can be used to 
run benchmarks across a grid of parameters, and it will "press" the 
results together in a tabular data structure.

```{r kern-bench-grid, dependson=c("kern-dens", "kern.dens-vec", "kern-dens-apply", "kern-dens-outer"), echo = -1, results='hide', message=FALSE}
set.seed(1234)
x <- rnorm(2^11)
kern_benchmarks <- bench::press(
  N = 2^(5:11),
  n = 2^c(5, 7, 9, 11),
  {
    bench::mark(
      loop = kern_dens_loop(x[1:N], h = 0.2, n = n),
      vec = kern_dens_vec(x[1:N], h = 0.2, n = n),
      apply = kern_dens_apply(x[1:N], h = 0.2, n = n),
      outer = kern_dens_outer(x[1:N], h = 0.2, n = n)
    )
  }
)
```

```{r kern-bench-grid-show, dependson="kern-bench-grid"}
head(kern_benchmarks)
```


```{r kern-bench-fig, dependson="kern-bench-grid", fig.cap="Median runtime for the four different implementations of kernel density estimation. The dashed gray line is a reference line with slope 1.", echo=FALSE, warning=FALSE, message=FALSE, out.width="100%"}
dplyr::mutate(kern_benchmarks, expression = factor(
    attr(expression, "description"),
    levels = c("loop", "vec", "apply", "outer")
  )) %>% 
  ggplot(aes(x = N, y = as.numeric(median) * 1000, color = expression)) + 
  geom_abline(intercept = -6, slope = 1, color = "gray", linetype = 2) +
  geom_point() +
  geom_line() +
  facet_wrap(~ n) + 
  scale_x_continuous(trans = "log2") + 
  scale_y_continuous("Time (ms)", trans = "log2",
    breaks = c(1e-4, 1e-3, 1e-2, 1e-1) * 1000) +
  scale_color_discrete("Function:") + 
  theme(legend.position = "top")
```

Figure \@ref(fig:kern-bench-fig) shows median runtimes for an experiment with 28
combinations of parameters for each of the four different implementations
yielding a total of 112 different R expressions being benchmarked. The results
confirm that `kern_dens_loop()` is substantially slower than the vectorized
implementations for all combinations of $N$ and $n$. However, Figure
\@ref(fig:kern-bench-fig) also reveals a new pattern: `kern_dens_outer()`
appears to scale with $N$ in a slightly different way than the two other
vectorized implementations for small $N$. It is comparable to or even a bit
faster than `kern_dens_vec()` and `kern_dens_apply()` for very small datasets,
while it becomes slower for the larger datasets.

Runtime for many algorithms have, to good approximation, a dominating 
power law behavior as a function of typical size parameters, that is, the 
runtime will scale approximately like $N \mapsto C N^a$ for constants $C$ and $a$
and with $N$ denoting a size parameter (number of data points, say). Therefore it is 
beneficial to plot runtime using log-log scales and to design benchmark studies
with size parameters being equidistant on a log-scale. With approximate power 
law scaling, the log runtime behaves like

\[
\log(C) + a \log(N),
\]

that is, on a log-log scale we see approximate straight lines. The slope 
reveals the exponent $a$, and two different algorithms for solving the 
same problem might have different exponents and thus different slopes 
on the log-log-scale. Two different implementations of the same 
algorithm should have approximately the same slope 
but may differ in the constant $C$ depending upon how efficient the 
particular implementation is in the particular programming language used.
Differences in $C$ correspond to vertical translations on the log-log scale. 

In practice, we will see some deviations from straight lines on the log-log plot
for a number of reasons. Writing the runtime as $C N^a + R(N)$, 
the residual term $R(N)$ will often be noticeable or even dominating and positive 
for small $N$. It is only for large enough $N$ 
that the power law term, $C N^a$, will dominate. In addition, runtime 
can be affected by hardware constraints such as cache and memory sizes, which 
can cause abrupt jumps in runtime.  

## Bandwidth selection {#bandwidth}

We show in this section that optimal bandwidth selection can be seen as a
bias-variance tradeoff. To quantify how good a specific estimate $\hat{f}_h$ is, 
we introduce the *integrated squared error*,\index{integrated squared error}

\[
\mathrm{ISE}(\hat{f}_h) = \int (\hat{f}_h(x) - f_0(x))^2 \ \mathrm{d}x = \|\hat{f}_h - f_0\|_2^2.
\]

The quality of the estimation procedure producing $\hat{f}_h$ from data
can then be quantified by taking the mean ISE,

\[
  \mathrm{MISE}(h) = \E(\mathrm{ISE}(\hat{f}_h)),
\]

where the expectation in an integral over the data distribution. Using Tonelli's theorem 
we may interchange the expectation and the integration over $x$ to 
get 

\[
  \mathrm{MISE}(h) = \int \mathrm{MSE}_h(x) \ \mathrm{d}x
\]

where 

\[
  \mathrm{MSE}_h(x) = \E(\hat{f}_h(x) - f_0(x))^2
\]

is the mean squared error in $x$. Introducing the mean of the 
kernel density estimator $\hat{f}_h$ as

\begin{equation} \label{eq:dens-mean}
  f_h(x) = \E(\hat{f}_h(x)),
\end{equation}

we can then also write 

\begin{align*}
  \mathrm{MSE}_h(x)  & = \E(\hat{f}_h(x) - f_h(x))^2 + (f_h(x) - f_0(x))^2 \\
                     & = \V(\hat{f}_h(x)) + \mathrm{bias}(\hat{f}_h(x))^2.
\end{align*}

We thus have 

\[
  \mathrm{MISE}(h) = \int \V(\hat{f}_h(x)) \ \mathrm{d}x + 
  \int \mathrm{bias}(\hat{f}_h(x))^2 \ \mathrm{d}x. 
\]

Choosing $h$ by minimizing $\mathrm{MISE}(h)$ can thus be seen as 
finding a suitable tradeoff between the integrated variance and integrated 
squared bias. Before proceeding with the general theory, we focus again 
on the rectangular kernel where some computations are simpler. 

### Revisiting the rectangular kernel {#rectangular}

For the rectangular kernel, and with $p(x, h) = \P(X \in (x-h, x+h))$, we 
find that 

\begin{align} 
  f_h(x)           & =  p(x, h) / (2h) (\#eq:meanRect) \\
  \V(\hat{f}_h(x)) & = \frac{p(x, h) (1 - p(x, h))}{4h^2 N} \approx f_h(x) \frac{1}{2hN},
  (\#eq:varRect)
\end{align}

with the approximation valid for $h$ small. 
We see from these computations that for $\hat{f}_h(x)$ to be approximately unbiased for any $x$
we need $h$ to be small -- ideally letting $h \to 0$ since then $f_h(x) \to f_0(x)$. 
However, this will make the variance blow up, and to minimize variance we should 
instead choose $h$ as large as possible. An appropriate choice of $h$
strikes a balance between the bias and the variance as a function 
of $h$ by minimizing the mean squared error of $\hat{f}_h(x)$.

To carry out a more rigorous analysis we assume that $f_0$ is sufficiently 
differentiable. For the derivation it may be helpful to think of $N$ as 
large and $h$ as small. Indeed, 
we will eventually choose $h = h_N$ as a function of $N$ such 
that as $N \to \infty$ we have $h_N \to 0$. We also note the $f_h(x)$
is a density, thus $\int f_h(x) \mathrm{d}x = 1$. 

We use a Taylor expansion of the distribution function $F_0$ to get that

\begin{align*}
f_h(x) & = \frac{1}{2h}\left(F_0(x + h) - F_0(x - h)\right) \\
& = \frac{1}{2h}\left(2h f_0(x) + \frac{h^3}{3} f_0''(x) + R_0(x,h) \right) \\
& = f_0(x) + \frac{h^2}{6} f_0''(x) + R_1(x,h) 
\end{align*}

where $R_1(x, h) = o(h^2)$. One should note how the quadratic terms in $h$ 
in the Taylor expansion canceled. This gives the following formula for 
the squared bias of $\hat{f}_h$.

\begin{align*}
\mathrm{bias}(\hat{f}_h(x))^2 & = (f_h(x) - f_0(x))^2 \\
& = \left(\frac{h^2}{6} f_0''(x) + R_1(x,h) \right)^2 \\
& = \frac{h^4}{36} f_0''(x)^2 + R(x,h)
\end{align*}

where $R(x,h) = o(h^4)$. For the variance we see from \@ref(eq:varRect)
that 

\[
\V(\hat{f}_h(x)) = f_h(x)\frac{1}{2Nh} - f_h(x)^2 \frac{1}{N}.
\]

Integrating the sum of the bias and the variance over $x$ gives the 
integrated mean squared error

\begin{align*}
\mathrm{MISE}(h) & = \int \mathrm{bias}(\hat{f}_h(x))^2 + \V(\hat{f}_h(x)) \mathrm{d}x \\
& = \frac{h^4}{36} \|f_0''\|^2_2 + \frac{1}{2hN} + \int R(x,h) \mathrm{d} x - 
\frac{1}{N} \int f_h(x)^2 \mathrm{d} x.
\end{align*}

If $f_h(x) \leq C$ (which happens if $f_0$ is bounded),

\[
\int f_h(x)^2 \mathrm{d} x \leq C \int f_h(x) \mathrm{d} x = C,
\]

and the last term is $o((Nh)^{-1})$. The second last term is $o(h^4)$ 
if we can interchange the limit and integration order. It is conceivable
that we can do so under suitable assumptions on $f_0$, but we will not pursue 
those at this place. The sum of the two remaining and asymptotically dominating terms in 
the formula for MISE is

\[
  \mathrm{AMISE}(h) = \frac{h^4}{36} \|f_0''\|^2_2 + \frac{1}{2N h},
\]

which is known as the *asymptotic* mean integrated squared error. Clearly, 
for this to be a useful formula, we must assume $\|f_0''\|_2^2 < \infty$. 
In this case the formula for AMISE can be used to find the asymptotic 
optimal tradeoff between integrated bias and variance. Differentiating w.r.t. $h$ we 
find that

\[
  \mathrm{AMISE}'(h) = \frac{h^3}{9} \|f_0''\|^2_2 - \frac{1}{2N h^2},
\]

and solving for $\mathrm{AMISE}'(h) = 0$ yields 

\[
  h_N = \left(\frac{9}{2 \|f_0''\|_2^2}\right)^{1/5} N^{-1/5}.
\]

When AMISE is regarded as a function of $h$ we observe that 
it tends to $\infty$ for $h \to 0$ as well as for $h \to \infty$, thus the unique 
stationary point $h_N$ is a unique global minimizer. Choosing the bandwidth to 
be $h_N$ will therefore minimize the asympototic mean integrated squared error, and it 
is in this sense an optimal choice of bandwidth.

We see how "wiggliness" of $f_0$ enters into the formula for the optimal 
bandwidth $h_N$ via $\|f_0''\|_2$. This norm of the second derivative 
is precisely a quantification of how much $f_0$ oscillates. A large value, 
indicating a wiggly $f_0$, will drive the optimal bandwidth down whereas
a small value will drive the optimal bandwidth up. 

We should also observe that if we plug the optimal bandwidth into the formula
for AMISE, we get 

\begin{align*}
\mathrm{AMISE}(h_N) & = \frac{h_N^4}{36} \|f_0''\|^2_2 + \frac{1}{2N h_N } \\
& = C N^{-4/5},
\end{align*}

which indicates that in terms of integrated mean squared error the rate 
at which we can nonparametrically estimate 
$f_0$ is $n^{-4/5}$. This should be contrasted to the common parametric
rate of $n^{-1}$ for mean squared error. 

From a practical viewpoint there is one major problem with the optimal 
bandwidth $h_N$; it depends via $\|f_0''\|^2_2$ upon the unknown $f_0$ 
that we are trying to estimate. We therefore refer to $h_N$ as an *oracle* 
bandwidth---it is the bandwidth that an oracle that knows $f_0$ would 
tell us to use. In practice, we will have to come up with an estimate 
of $\|f_0''\|^2_2$ and plug that estimate into the formula for $h_N$.
We pursue a couple of different options for doing so for general kernel 
density estimators below together with methods that do not rely on the 
AMISE formula. 

### The oracle bandwidth and its estimation

We can generalize the computations for the rectangular kernel in the section 
above to any kernel, subject to some regularity conditions. To this end we 
define the *asymptotic mean integrated squared error* as \index{AMISE}
\index{asymptotic mean integrated squared error}

\[
  \mathrm{AMISE}(h) = \frac{\|K\|_2^2}{nh} + \frac{h^4 \sigma^4_K \|f_0''\|_2^2}{4}
\]

with 

\[
  \|g\|_2^2 = \int g(x)^2 \ \mathrm{d}x  \quad (\mathrm{squared } \ L_2\mathrm{-norm}).
\]

and $\sigma_K^2 = \int x^2 K(x) \ \mathrm{d}x$.

For $\mathrm{AMISE}(h)$ to be well defined and finite we need the density 
$f_0$ to be twice differentiable, we need both $f_0''$ and the kernel, $K$, 
to be square integrable, and we need $\sigma_K^2$ to be finite. 
When the kernel is a density, these conditions
turn out to also be sufficient to express the asymptotic behavior of 
$\mathrm{MISE}(h)$ in terms of $\mathrm{AMISE}(h)$.

::: {.theorem #amise} 
Suppose the kernel $K$ is a square integrable density for a probability 
distribution on $\mathbb{R}$. If 

\[
 \sigma_K^2 = \int x^2 K(x) \ \mathrm{d}x \in (0, \infty),
\]

and if $f_0$ is twice differentiable with if $f_0''$ square integrable,

\[
  \mathrm{MISE}(h) = \mathrm{AMISE}(h) + o((Nh)^{-1} + h^4).
\]
:::

The proof of Theorem \@ref(thm:amise) can be based on the same kind of Taylor 
expansion argument as in Section \@ref(rectangular). See Proposition A.1 in 
@Tsybakov:2009 for a rigorous proof, where remainder terms are handled carefully.


By minimizing $\mathrm{AMISE}(h)$ we derive the asymptotically optimal *oracle bandwidth*
\index{oracle bandwidth}

\begin{equation}
(\#eq:oracle)
h_N = \left( \frac{\|K\|_2^2}{ \|f_0''\|^2_2  \sigma_K^4} \right)^{1/5} n^{-1/5}.
\end{equation}

If we plug this formula into the formula for AMISE we arrive at the asymptotic 
error rate $\mathrm{AMISE}(h_N) = C N^{-4/5}$ with a constant $C$ depending on 
$f_0''$ and the kernel. 

It is noteworthy that the asymptotic analysis can be carried 
out even if $K$ is allowed to take negative values. Then $K$ is not a density and 
the resulting estimate may not be a valid density as it is. @Tsybakov:2009 
demonstrates how to improve on 
the rate $N^{-4/5}$ by allowing for kernels whose moments of order two or above vanish. 
Necessarily, such kernels must take negative values and cannot be densities.

We observe that for the rectangular kernel, 

\[
  \sigma_K^4 = \left(\frac{1}{2} \int_{-1}^1 z^2 \ \mathrm{d} z\right)^2 = \frac{1}{9}
\]

and 

\[
  \|K\|_2^2 = \frac{1}{2^2} \int_{-1}^1 \ \mathrm{d} z = \frac{1}{2}.
\]

Plugging these numbers into \@ref(eq:oracle) we find the oracle bandwidth for 
the rectangular kernel as derived in Section \@ref(rectangular). For the Gaussian
kernel we find that $\sigma_K^4 = 1$, while 

\[
  \|K\|_2^2 = \frac{1}{2 \pi} \int e^{-x^2} \ \mathrm{d} x = \frac{1}{2 \sqrt{\pi}}.
\]

The major practical problem with the oracle bandwidth is that it depends on $\|f_0''\|^2_2$.
Thus to compute the asymptotically optimal 
bandwidth we have to know the second derivative of $f_0$ that we are trying 
to estimate in the first place. 
To resolve the circularity we will make a first guess of what $f_0$ is and plug that 
guess into the formula for the oracle bandwidth. 

Our first guess is that $f_0$ is Gaussian with mean 0 and variance $\sigma^2$. 
Then 

\begin{align*}
\|f_0''\|^2_2 & = \frac{1}{2 \pi \sigma^2} \int \left(\frac{x^2}{\sigma^4} -  \frac{1}{\sigma^2}\right)^2 e^{-x^2/\sigma^2} \ \mathrm{d}x \\
& = \frac{1}{2 \sigma^9 \sqrt{\pi}} \frac{1}{\sqrt{\pi \sigma^2}} \int (x^4 - 2 \sigma^2 x^2 + \sigma^4) e^{-x^2/\sigma^2} \ \mathrm{d}x \\ 
& = \frac{1}{2 \sigma^9 \sqrt{\pi}} (\frac{3}{4} \sigma^4 - \sigma^4 + \sigma^4) \\
& = \frac{3}{8 \sigma^5 \sqrt{\pi}}.
\end{align*}

Plugging this expression for the squared 2-norm of the second derivative of the density 
into the formula for the oracle bandwidth gives 

\begin{equation}
(\#eq:oracle-silverman)
h_N = \left( \frac{8 \sqrt{\pi} \|K\|_2^2}{3 \sigma_K^4} \right)^{1/5} \sigma n^{-1/5},
\end{equation}

with $\sigma$ the only quantity depending on the unknown density $f_0$. We can now
simply estimate $\sigma$, using, e.g., the empirical standard 
deviation $\hat{\sigma}$, and plug this estimate into the formula above to get an 
estimate of the oracle bandwidth. 

It is well known that the empirical standard deviation is sensitive to outliers,
and if $\sigma$ is overestimated for that reason, the bandwidth will be too large
and the resulting estimate will be oversmoothed. To
get more robust bandwidth selection, @Silverman:1986 suggested using 
the interquartile range to estimate $\sigma$. In fact, he suggested 
estimating $\sigma$ by

\[
  \tilde{\sigma} = \min\{\hat{\sigma}, \mathrm{IQR} / 1.34\}.
\]

In this estimator, IQR denotes the empirical interquartile range, and
$1.34$ is approximately the interquartile range, $\Phi^{-1}(0.75) - \Phi^{-1}(0.25)$, 
of the standard Gaussian distribution. Curiously, the interquartile range for 
the standard Gaussian distribution is $1.35$ to two decimals accuracy, but the 
use of $1.34$ in the estimator $\tilde{\sigma}$
has prevailed. Silverman, moreover, suggested to reduce the kernel-dependent 
constant in the formula \@ref(eq:oracle-silverman) for $h_N$ to further reduce
oversmoothing. 

If we specialize to the Gaussian kernel, formula \@ref(eq:oracle-silverman) 
simplifies to 

\[ 
  \hat{h}_N = \left(\frac{4}{3}\right)^{1/5} \tilde{\sigma} n^{-1/5},
\]

with $\tilde{\sigma}$ plugged in as a robust estimate of $\sigma$. @Silverman:1986 
made the ad hoc suggestion to reduce the factor $(4/3)^{1/5} \approx 1.06$ to $0.9$.
This results in the bandwidth estimate 

\[
  \hat{h}_N = 0.9 \tilde{\sigma} n^{-1/5},
\]

which has become known as *Silverman's rule of thumb*.\index{Silverman's rule of thumb}  

Silverman's rule of thumb is the default bandwidth estimator for `density()`
as implemented by the function `bw.nrd0()`. The `bw.nrd()` function implements 
bandwidth selection using the factor $1.06$ instead of $0.9$. Though the theoretical 
derivations behind these implementations assume that $f_0$ 
is Gaussian, either will give reasonable bandwidth selection for a range of 
unimodal distributions. If $f_0$ is multimodal, Silverman's 
rule of thumb is known to oversmooth the density estimate.  

Instead of computing $\|f_0''\|^2_2$ assuming that the distribution is Gaussian,
we can compute the norm for a pilot estimate, $\tilde{f}$, and plug the result 
into the formula for $h_N$. If the pilot estimate is a kernel estimate with 
kernel $H$ and bandwidth $r$ we get 

\[
  \|\tilde{f}''\|^2_2 = \frac{1}{n^2r^6} \sum_{i = 1}^N \sum_{j=1}^N 
\int H''\left( \frac{x - x_i}{r} \right) H''\left( \frac{x - x_j}{r} \right) \mathrm{d} x.
\]

The problem is, of course, that now we have to choose the pilot bandwidth
$r$. But doing so using a simple method like Silverman's rule of thumb at this 
stage is typically not too bad an idea. Thus we arrive at the following 
plug-in procedure using the Gaussian kernel for the pilot estimate:

* Compute an estimate, $\hat{r}$, of the pilot bandwidth using Silverman's 
rule of thumb.
* Compute $\|\tilde{f}''\|^2_2$ using the Gaussian kernel as pilot kernel $H$ and 
using the estimated pilot bandwidth $\hat{r}$.
* Plug $\|\tilde{f}''\|^2_2$ into the oracle bandwidth formula \@ref(eq:oracle)
to compute $\hat{h}_N$ for the kernel $K$. 

Note that to use a pilot kernel different from the Gaussian we have 
to adjust the constant 0.9 (or 1.06) in Silverman's rule of thumb accordingly 
by computing $\|H\|_2^2$ and $\sigma_H^4$ and using \@ref(eq:oracle-silverman). 

@Sheather:1991 took these plug-in ideas a step further and analyzed in detail 
how to choose the pilot bandwidth in a good and data adaptive way. The resulting
method is somewhat complicated but implementable. We skip the details but simply 
observe that their method is implemented in R in the function `bw.SJ()`, and
it can be selected when using `density()` by setting the argument `bw = "SJ"`. 
This plug-in method is regarded as a solid default that performs well for 
many different data generating densities $f_0$. 

(ref:kern-dens-bw) Kernel density estimates with the Gaussian kernel using R's bandwidth selection functions `bw.nrd()` (blue), `bw.nrd0()` (red) and `bw.SJ()` (purple).

```{r kern-dens-fig-bw, dependson=c( "kern-dens"), fig.cap = '(ref:kern-dens-bw)'}
f_hat_nrd <- kern_dens_vec(angle$psi, bw.nrd(angle$psi))
f_hat_nrd0 <- kern_dens_vec(angle$psi, bw.nrd0(angle$psi))
f_hat_SJ <- kern_dens_vec(angle$psi, bw.SJ(angle$psi))
hist(angle$psi, prob = TRUE, border = NA)
lines(f_hat_nrd, col = "blue", lwd = 2)
lines(f_hat_nrd0, col = "red", lwd = 2)
lines(f_hat_SJ, col = "purple", lwd = 2)
```

## Parametric models {#likelihood}

As an alternative to the *ad hoc* nonparametric kernel density estimator, we
could also approximate the density $f_0$ by a density from a parametrized
statistical model $(f_{\theta})_{\theta \in \Theta}$, where $f_{\theta}$ is a
density w.r.t. Lebesgue measure on $\mathbb{R}$ for each parameter $\theta \in
\Theta$. If $\hat{\theta}$ is an estimate of the parameter, $f_{\hat{\theta}}$
is an estimate of the unknown density $f_0$. For a parametric family we can
always try to use the maximum-likelihood estimator

\[
\hat{\theta} = \argmax_{\theta} \sum_{j=1}^N \log f_{\theta}(x_j)
\]

as an estimate of $\theta$. If the parametrized family of distributions is 
the family of Gaussian distributions, we can estimate the mean and 
variance parameters by the empirical mean and variance 
for the data and plug those numbers into the density for the Gaussian
distribution. In this way obtain a Gaussian density estimate of $f_0$.

```{r psi-gauss, fig.cap="Gaussian density (red) fitted to the $\\psi$-angles."}
psi_mean <- mean(angle$psi)
psi_sd <- sd(angle$psi)
hist(angle$psi, prob = TRUE)
rug(angle$psi)
curve(dnorm(x, psi_mean, psi_sd), add = TRUE, col = "red")
```

As Figure \@ref(fig:psi-gauss) shows, if we fit a Gaussian distribution to the
$\psi$-angle data we get a density estimate that clearly does not match the
histogram. The Gaussian density matches the data on the first and second
moments, but the histogram shows a bimodality that the Gaussian
density can never match. Thus we need a more flexible model than
the family of Gaussian distributions. Could we go all-in and simply maximize the
likelihood over all possible densities?

The log-likelihood

\[
  \ell(f) = \sum_{j=1}^N \log f(x_j)
\]

is well defined as a function of the density $f$---even when $f$ is not restricted 
to belong to a model parametrized by a finite-dimensional parameter.
To investigate if a nonparametric maximum-likelihood estimator is meaningful we consider how the likelihood
behaves for the densities

\[
  \overline{f}_h(x) = \frac{1}{Nh \sqrt{2 \pi}} \sum_{j=1}^N e^{- \frac{(x - x_j)^2}{2 h^2} }
\]

for different choices of $h$. Note that $\overline{f}_h$ is simply the 
kernel density estimator with the Gaussian kernel and bandwidth $h$. 
Contrary to the implementations in Section \@ref(kernel-density), 
that evaluated the estimator in a grid, we implement the kernel density estimator 
as a function. We do this in a way similarly to the implementations 
in Appendix \@ref(functions), in particular, we explicitly vectorize the 
computations over the argument `x`, see Appendix \@ref(vectorization).


```{r gaus-dens}
gauss <- function(x, h) {
  exp(- x^2 / (2 * h^2)) / (h * sqrt(2 * pi))
}
f_bar <- Vectorize(
  function(x, h) mean(gauss(x - angle$psi, h))
)
```

Figure \@ref(fig:gausKern) shows what some of 
these densities look like compared to the histogram of the $\psi$-angle data. 
Clearly, for large $h$ these densities are smooth and slowly oscillating, while 
as $h$ gets smaller the densities  become more and more wiggly. As $h \to 0$
the density becomes increasingly adapted to the individual data points. 

(ref:gausKern) The densities $\overline{f}_h$ for different choices of $h$. 

```{r gausKern, fig.cap="(ref:gausKern)", fig.show="hold", out.width="49%", dependson=c("gaus-dens"), fig.pos="h", echo=FALSE}
hist(angle$psi, prob = TRUE, main = "h = 1")
rug(angle$psi)
curve(f_bar(x, 1), add = TRUE, col = "red")
hist(angle$psi, prob = TRUE, main = "h = 0.5")
rug(angle$psi)
curve(f_bar(x, 0.5), add = TRUE, col = "red", n = 1001)
hist(angle$psi, prob = TRUE, main = "h = 0.1")
rug(angle$psi)
curve(f_bar(x, 0.1), add = TRUE, col = "red", n = 1001)
hist(angle$psi, prob = TRUE, main = "h = 0.05")
rug(angle$psi)
curve(f_bar(x, 0.05), add = TRUE, col = "red", n = 10001)
```

The way that these densities adapt to the data points is reflected in the 
log-likelihood as well. 

```{r LogLike, dependson=c( "gausKern"), fig.cap="Log-likelihood, $\\ell(\\overline{f}_h)$, for the densities $\\overline{f}_h$ as a function of $h$. Note the log-scale on the right.", out.width="49%", fig.show='hold'}
hseq <- seq(1, 0.001, -0.001)
ll <- sapply(hseq, function(h) sum(log(f_bar(angle$psi, h))))
ggplot(mapping = aes(hseq, ll)) + geom_line() + 
  xlab("h") + ylab("Log-likelihood")
ggplot(mapping = aes(hseq, ll)) + geom_line() + 
  scale_x_log10("h (log-scale)") + ylab("")
```

From Figure \@ref(fig:LogLike) it is clear that the likelihood is decreasing in 
$h$, and it appears that it is unbounded as $h \to 0$. This is most clearly 
seen on the figure when $h$ is plotted on the log-scale because then it appears
that the log-likelihood approximately behaves as $-\log(h)$ for $h \to 0$. 

We can show that that is, indeed, the case. If $x_i \neq x_j$ when $i \neq j$

\begin{align*}
\ell(\overline{f}_h) & = \sum_{i} \log\left(1 + \sum_{j \neq i} e^{-(x_i - x_j)^2 / (2 h^2)} \right) - 
n \log(nh\sqrt{2 \pi}) \\
& \sim - n \log(nh\sqrt{2 \pi}) = - n \log(h) - n \log(n\sqrt{2 \pi})
\end{align*}

for $h \to 0$. Hence, $\ell(\overline{f}_h) \to \infty$ for $h \to 0$. This demonstrates that the maximum-likelihood estimator 
of the density does not exist in the set of all distributions with densities.

In the sense of [weak convergence](https://en.wikipedia.org/wiki/Convergence_of_measures#Weak_convergence_of_measures) it actually holds that 

\[
  \overline{f}_h \cdot m \overset{\mathrm{wk}}{\longrightarrow} 
P_N = \frac{1}{N} \sum_{j=1}^N \delta_{x_j}
\]

for $h \to 0$. The *empirical measure* $P_N$\index{empirical measure} can sensibly be regarded as 
a nonparametric maximum-likelihood estimator of the distribution, but the empirical measure does not have 
a density. We conclude that we cannot directly define a sensible 
density estimator as a maximum-likelihood estimator.  

### Basis expansions {#basis-density}

We suppose in this section that the data points are all contained in the
interval $[a,b]$ for $a, b \in \mathbb{R}$. This is true for the angle 
data with $a = -\pi$ and $b = \pi$ no matter the size of the dataset,
but if $f_0$ does not have a bounded support it may be necessary to let $a$
and $b$ change with the data. However, for any fixed dataset we can choose 
some sufficiently large $a$ and $b$. 

Suppose that we have chosen $p$ continuous 
functions $\varphi_1, \ldots, \varphi_p : [a,b] \to \mathbb{R}$. These will be 
called *basis functions*. We then define the density

\[
f_{\boldsymbol{\beta}}(x) =  \frac{1}{c(\boldsymbol{\beta})}
\exp\left(\sum_{k=1}^p \beta_k \varphi_k(x)\right) 
\]

for the parameter $\boldsymbol{\beta} = (\beta_1, \ldots, \beta_h)^T \in \mathbb{R}^p$.
The normalization constant is 

\[
c(\boldsymbol{\beta}) = \int_a^b \exp\left(\sum_{k=1}^p \beta_k \varphi_k(x)\right) \mathrm{d} x.
\]

The maximum-likelihood estimator is then 

\begin{align*}
\hat{\boldsymbol{\beta}} & = 
\argmax_{\boldsymbol{\beta} \in \mathbb{R}^p} \ \sum_{j=1}^N \left(\sum_{k=1}^p \beta_k \varphi_k(x_j) \right) - \log c(\boldsymbol{\beta}) \\
& = \argmax_{\boldsymbol{\beta} \in \mathbb{R}^p} \ \ \mathbf{1}^T \boldsymbol{\Phi} \boldsymbol{\beta} - 
N \log c(\boldsymbol{\beta}).
\end{align*}

Here $\boldsymbol{\Phi}$ is the $N \times p$ matrix with $\boldsymbol{\Phi}_{jk} = \varphi_k(x_j)$. 
By differentiation, the maximum-likelihood estimator is seen to be a solution to the
non-linear equation 

\begin{equation}
\mathbf{1}^T \boldsymbol{\Phi} = N \tau(\boldsymbol{\beta})
\end{equation}

where $\tau : \mathbb{R}^p \to \mathbb{R}^p$ is the map with coordinates 

\[
\tau_l(\boldsymbol{\beta}) = \frac{\partial_l c(\boldsymbol{\beta})}{c(\boldsymbol{\beta})}
=
\frac{1}{c(\boldsymbol{\beta})}
\int_a^b \varphi_l(x) \exp\left(\sum_{k=1}^p \beta_k \varphi_k(x)\right) \mathrm{d} x
\]

for $l \in \{1, \ldots, p\}$. It may not be entirely straightforward how to compute 
$c(\boldsymbol{\beta})$, but it is possible to implement the computation of 
$\tau$ using numerical integration, and the equation above can then be solved 
numerically as well. We will, however, pursue a different estimation technique that 
does not require numerical evaluation of integrals or solving non-linear equations. 

### Score matching

Suppose that the basis functions are twice differentiable and define the $p \times p$ 
matrix $\mathbf{B}$ and the $p$-vector $\mathbf{b}$ by 

\begin{align*}
\mathbf{B}_{kl} & = \sum_{j=1}^N \phi_k'(x_j) \phi_l'(x_j) \\
\mathbf{b}_k & = \sum_{j=1}^N \phi_k''(x_j).
\end{align*}

If $\mathbf{B}$ is of full rank $p$, the *score matching*\index{score matching} estimator 
equals 

\[
\tilde{\boldsymbol{\beta}} = \mathbf{B}^{-1} \mathbf{b}.
\]

The score matching estimator is simple to define and compute by the expression 
above, which involves only the evaluations of the two first derivatives of the 
basis functions in the data points and a single solution of a linear equation 
in $p$ unknowns. The justification of the score matching estimator is a little 
more involved, and it involves a boundary assumption, which we will return to. 





Assuming the basis functions and $f_0$ are continuously differentiable 
we define the *Fisher divergence*\index{Fisher divergence} as 

\[
  J(\boldsymbol{\beta}) = \int_a^b 
  \left( \frac{f_0'(x)}{f_0(x)} - \frac{f_{\boldsymbol{\beta}}'(x)}{f_{\boldsymbol{\beta}}(x)}\right)^2 
  f_0(x) \mathrm{d}x.
\]

The best approximation of $f_0$ in terms of the Fisher divergence is found by 
minimizing $J(\boldsymbol{\beta})$. To derive an estimator we rewrite $J(\boldsymbol{\beta})$
as follows

\begin{align*}
J(\boldsymbol{\beta}) & =  \int_a^b  \left( \frac{f_0'(x)}{f_0(x)} \right)^2 f_0(x) \mathrm{d}x 
+
\int_a^b  \left( \frac{f_{\boldsymbol{\beta}}'(x)}{f_{\boldsymbol{\beta}}(x)} \right)^2 f_0(x) \mathrm{d}x 
+ \\
& \qquad 2  \int_a^b 
\left(\frac{f_{\boldsymbol{\beta}}'(x)}{f_{\boldsymbol{\beta}}(x)} \right) 
\left( \frac{f_0'(x)}{f_0(x)} \right) f_0(x) \mathrm{d}x \\
& = C(f_0) + 
\int_a^b  \left( \sum_{k=1}^p \beta_k \phi_k'(x) \right)^2 f_0(x) \mathrm{d}x 
 + 
 2  \int_a^b \left( \sum_{k=1}^p \beta_k \phi_k'(x)  \right) f_0'(x) \mathrm{d}x.
\end{align*}

The first term above does not depend on the parameter $\boldsymbol{\beta}$,
and the second term is an expectation, which can be estimated by 

\[
\frac{1}{N} \sum_{j=1}^N  \left(\sum_{k=1}^p \beta_k \phi_k'(x_j) \right)^2 
= \frac{1}{N} \boldsymbol{\beta}^T \mathbf{B} \boldsymbol{\beta}
\]

where $\mathbf{B}$ is 
a symmetric and positive semi-definite matrix. To 
estimate the third term above we use integration-by-parts, which gives

\[
   \int_a^b \left( \sum_{k=1}^p \beta_k \phi_k'(x)  \right) f_0'(x) \mathrm{d}x
 = \left( \sum_{k=1}^p \beta_k \phi_k'(x)  \right) f_0(x) \Bigg|_a^b   
 -  \int_a^b \left( \sum_{k=1}^p \beta_k \phi_k''(x)  \right) f_0(x) \mathrm{d}x
\]

assuming that the basis functions are twice continuously differentiable. 
If

* either $f_0(a) = f_0(b) = 0$ 
* or $f_0(a) = f_0(b)$ together with $\phi_k''(a) = \phi_k''(b)$ for $k = 1, \ldots, p$

the first term above 
vanishes, and we can estimate the second term as an expectation by  

\[
\frac{1}{N} \sum_{j=1}^N  \sum_{k=1}^p \beta_k \phi_k''(x_j) 
= \frac{1}{N}  \mathbf{b} \boldsymbol{\beta}
\]

where

\[
  \tilde{\boldsymbol{\beta}} = \argmin_{\boldsymbol{\beta}} \tilde{J}(\boldsymbol{\beta})
\]




where the empirical objective function we minimize is the quadratic function

\[
  \tilde{J}(\boldsymbol{\beta})  = \boldsymbol{\beta}^T \mathbf{B} \boldsymbol{\beta}
 - 2 \mathbf{b} \boldsymbol{\beta}
\]

When $\mathbf{B}$ has full rank, the unique minimizer and score matching 
estimator is given explicitly as



Many basis functions are possible. One possible choice is the 
trigonometric functions

\[ 
  \varphi_1(x) = \cos(x), \varphi_2(x) = \sin(x), \ldots, 
  \varphi_{2p-1}(x) = \cos(hx),  \varphi_{2p}(x) = \sin(hx)
\]

on the interval $[-\pi, \pi]$. In Section \@ref(vM) a simple special case 
was actually treated corresponding to $p = 2$, where the normalization constant 
was identified in terms of a modified Bessel function. 

John von Neumann is famously quoted for saying that "With four parameters I can 
fit an elephant, and with five I can make him wiggle his trunk." An example 
of a four-parameter family of distributions is the [normal-inverse Gaussian](https://en.wikipedia.org/wiki/Normal-inverse_Gaussian_distribution)
distributions with the [generalized hyperbolic distributions](https://en.wikipedia.org/wiki/Generalised_hyperbolic_distribution) an extension with five
parameters. These are flexible distributions, with the parameters controlling 
scale and location of the distribution, and additional shape properties 
such as tail decay and (a)symmetry. However, von Neumann was probably 
thinking more in terms of expansions with four or five suitably 
chosen basis functions, such as the trigonometric functions above, 
or using polynomials or splines as will be treated in more detail in 
Sections \@ref(poly) and \@ref(splines). What his quote says is that 
even with relatively few 
parameters we can adapt a basis expansion so that approximates many 
functions quite well. With too many parameters, we easily end up 
overfitting the model to data. That is, 
we adapt the model too much to the individual data points in the dataset, 
in which case it is not a good representation of the underlying distribution.

The quote is not a mathematical statement but an empirical observation. 
With a handful of parameters we get a quite flexible class of 
densities that will fit many real datasets well. But remember 
that a reasonably good fit does not mean
that we have found the "true" data generating model. Though data is 
in some situations not as scarce a resource today as when von Neumann made
elephants wiggle their trunks, the quote suggests that $h$ should grow
rather slowly with $N$ to avoid overfitting.



### Method of sieves {#sieves}

A sieve is a family of models, $\Theta_{h}$, indexed by
a real valued parameter, $h \in \mathbb{R}$, such that $\Theta_{h_1} \subseteq \Theta_{h_2}$ for 
$h_1 \leq h_2$. In this chapter $\Theta_{h}$ will denote a set of probability densities. 
If the increasing family of models is chosen in a sensible way, we 
may be able to compute the maximum-likelihood estimator

\[
\hat{f}_h = \argmax_{f \in \Theta_h} \ell(f),
\]

and we may even be able to choose $h = h_N$ as a function of the sample size $n$
such that $\hat{f}_{h_N}$ becomes a consistent estimator of $f_0$. 

It is possible to take 

\[
\Theta_h = \{ \overline{f}_{h'} \mid h' \leq h \}
\]

with $\overline{f}_{h'}$ as defined above, in which case $\hat{f}_h = \overline{f}_h$. 
We will see in the following section that this is simply a kernel estimator. 

A more interesting example is obtained by letting 

\[
\Theta_h = \left\{ x \mapsto  \frac{1}{h \sqrt{2 \pi}} \int  e^{- \frac{(x - z)^2}{2 h^2} } \mathrm{d}\mu(z) \Biggm| \mu \textrm{ a probability measure}   \right\},
\]

which is known as the convolution sieve. We note that $\overline{f}_h \in \Theta_h$ by 
taking $\mu = \P_N$, but generally $\hat{f}_h$ will be 
different from $\overline{f}_h$. 

We will not pursue the general theory of sieve estimators, but refer 
to the paper [Nonparametric Maximum Likelihood Estimation by the Method of Sieves](https://projecteuclid.org/euclid.aos/1176345782) by Geman and Hwang. 
In the following section we will work out some more practical details 
for a particular sieve estimator based on a basis expansion of the 
log-density.

## Cross-validation {#cv}

An alternative to relying on the asymptotic optimality arguments for 
integrated mean squared error and the corresponding plug-in estimates 
of the bandwidth is known as *cross-validation*. The method mimics 
the idea of setting aside a subset of the dataset, which is then *not*
used for computing an estimate but only for validating the estimator's 
performance. The benefit of cross-validation over simply setting 
aside a validation dataset is that we do not "waste" any of the data points
on validation only. All data points are used for the ultimate computation
of the estimate. The deficit is that cross-validation is usually 
computationally more demanding. 

Suppose that $I_1, \ldots, I_k$ form a partition of the index set $\{1, \ldots, N\}$
and define 

\[
  I^{-i} = \bigcup_{l: i \not \in I_l} I_l.
\]

That is, $I^{-i}$ contains all indices but those that belong to the set $I_l$ 
containing $i$. In particular, $i \not \in I^{-i}$. Define also $N_i = |I^{-i}|$
and 

\[
  \hat{f}^{-i}_h = \frac{1}{h N_i} \sum_{j \in I^{-i}} K\left(\frac{x_i - x_j}{h}\right).
\]

That is, $\hat{f}^{-i}_h$ is the kernel density estimate based on data with 
indices in $I^{-i}$ and evaluated in $x_i$. Since the density 
estimate evaluated in $x_i$ is not based on $x_i$, the quantity $\hat{f}^{-i}_h$ 
can be used to assess how well the density estimate computed using a bandwidth
$h$ concur with the data point $x_i$. This can be summarized using the 
log-likelihood 

\[
  \ell_{\mathrm{CV}}(h) = \sum_{i=1}^N \log (\hat{f}^{-i}_h),
\]

that we will refer to as the cross-validated log-likelihood, and we define 
the bandwidth estimate as

\[
  \hat{h}_{\mathrm{CV}} = \argmax_h \ \ \ell_{\mathrm{CV}}(h).
\]

This cross-validation based bandwidth can then be used for computing kernel 
density estimates using the entire dataset. 

If the partition of indices consists of $k$ subsets we usually talk about 
$k$-fold cross-validation. If $k = n$ so that all subsets consist of just a single 
index we talk about leave-one-out cross-validation. For leave-one-out 
cross-validation there is only one possible partition, while for $k < n$ 
there are many possible partitions. Which should be chosen then? In practice, 
we choose the partition by sampling indices randomly without replacement 
into $k$ sets of size roughly $n / k$. 

It is also possible to use cross-validation in combination with MISE. Rewriting 
we find that 

\[
  \mathrm{MISE}(h) = E (\| \hat{f}_h\|_2^2) - 2 E (\hat{f}_h(X)) + \E(\|f_0^2\|_2^2)
\]

for $X$ a random variable independent of the data and with distribution having 
density $f_0$. The last term does not depend upon $h$ and we can ignore it from
the point of view of minimizing MISE. For the first term we have an unbiased 
estimate in $\| \hat{f}_h\|_2^2$. The middle term can be estimated
without bias by 

\[ 
  \frac{2}{N} \sum_{i=1}^N \hat{f}^{-i}_h,
\]

and this leads to the statistic

\[
  \mathrm{UCV}(h) = \| \hat{f}_h\|_2^2 - \frac{2}{N} \sum_{i=1}^N \hat{f}^{-i}_h
\]

known as the unbiased cross-validation criterion. The corresponding 
bandwidth estimate is 

\[
  \hat{h}_{\mathrm{UCV}} = \textrm{arg min}_h \ \ \mathrm{UCV}(h).
\]

Contrary to the log-likelihood based criterion, this criterion requires 
the computation of $\| \hat{f}_h\|_2^2$. Bandwidth selection using UCV 
is implemented in R in the function `bw.ucv()` and can be 
used with `density()` by setting the argument `bw = "ucv"`. 

## Exercises {#density:ex}

::: {.exercise #epanechnikov}
The Epanechnikov kernel is given by 

\[
K(x) = \frac{3}{4}(1 - x^2)
\]

for $x \in [-1, 1]$ and 0 elsewhere. Show that this is a probability density
with mean zero and compute $\sigma_K^2$ as well as $\|K\|_2^2$. 
:::


For the following exercises use the log(F12) variable as considered in 
Exercise \@ref(exr:plotHist).

::: {.exercise #epanechnikov-density}
Use `density()` to compute the kernel density estimate with the Epanechnikov kernel
to the log(F12) data. Try different bandwidths. 
:::

::: {.exercise #epanechnikov-implement}
Implement kernel density estimation yourself using the Epanechnikov kernel. 
Test your implementation by comparing it to `density()` using the log(F12) data.
:::

::: {.exercise #benchmark-epanechnikov}
Construct the following vector
:::

```{r xsim}
x <- rnorm(2^13)
```

Then use `bench::mark()` to benchmark the computation of 

```{r density-xsim, eval = FALSE}
density(x[1:k], 0.2)
```

for `k` ranging from $2^5$ to $2^{13}$. Summarize the benchmarking results. 

::: {.exercise #benchmark-own-epanechnikov}
Benchmark your own implementation of kernel density estimation using the Epanechnikov
kernel. Compare the results to those obtained for `density()`. 
:::

::: {.exercise #benchmark-kernel}
Experiment with different implementations of kernel evaluation in R using 
the Gaussian kernel and the Epanechnikov kernel. Use `bench::mark()` to compare 
the different implementations. 
:::
