---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Random number generation

This chapter deals with algorithms for generating random numbers from a
probability distribution on $\mathbb{R}$ or a subset thereof. Two applications
of random number generation within statistics are: simulation studies and Monte
Carlo integration. A simulation study investigates distributional properties of
statistical procedures, and such studies have become an indispensable
experimental part of the development of statistical methodology. Monte Carlo
integration is a more specific application of random number generation as a way
to numerically compute a probability or an integral.

Random number generators are also used when implementing randomization protocols
in randomized controlled trials, for resampling and subsampling based
statistical methods, such as bootstrapping, and in other randomized algorithms,
such as stochastic optimization algorithms. There are likewise important
applications of random number generators outside of statistics, e.g., within
cryptography. We will not pursue all these different applications, nor will we
deal with the design and implementation of general simulation studies.

In this chapter the focus is on the generation of a single random number or an
i.i.d. sequence of random numbers. In particular, how pseudorandom number
generators are implemented and used to approximate sampling of i.i.d. random
variables that are uniformly distributed on $(0, 1)$. Section \@ref(transform)
illustrates how simulation from other distributions than the uniform can be
obtained from pseudorandom number generators via transformations.

Chapter \@ref(reject-samp) treats rejection sampling, in particular for
distributions on $\mathbb{R}$. Chapter \@ref(mci) deals with Monte Carlo
integration in greater detail, in particular how to assess the precision of
computations based on random number generation. Chapters \@ref(stochopt) and
\@ref(stochem) rely on (pseudo)random numbers as an integral part of any
stochastic optimization algorithm.

## Pseudorandom number generators

Most simulation algorithms are based on algorithms for generating 
*pseudorandom* uniformly distributed variables in $(0, 1)$. Such pseudorandom
variables all arise from deterministic sequences of numbers initiated by a *seed*. A classical
example of a pseudorandom integer generator is the 
linear congruential generator. A sequence of numbers from 
this generator is computed iteratively by

\[
x_{n+1} = (a x_n + c)  \text{ mod m}
\]

for integer parameters $a$, $c$ and $m$. The seed $x_1$ is a 
number between $0$ and $m - 1$, and the resulting 
sequence is in the set $\{0, \ldots, m - 1\}$. The ANSI C standard specifies 
the choices $m = 2^{31}$, $a = 1,103,515,245$ and $c = 12,345$. The generator
is simple to understand and implement but has been superseded by much better 
generators. 

Pseudorandom number generators are generally defined in terms of a finite state space 
$\mathcal{Z}$ and a one-to-one map $f : \mathcal{Z} \to \mathcal{Z}$. The 
generator produces a sequence in $\mathcal{Z}$ iteratively from the 
seed $\mathbf{z}_1 \in \mathcal{Z}$ by

\[
\mathbf{z}_n = f(\mathbf{z}_{n-1}).
\]

Pseudorandom integers are typically obtained as 

\[
x_n = h(\mathbf{z}_n)
\]

for a map $h : \mathcal{Z} \mapsto \mathbb{Z}$. If the 
image of $h$ is in the set $\{0, 1, \ldots, 2^{w} - 1\}$ of $w$-bit integers, 
pseudorandom numbers in $[0, 1)$ are typically obtained as 

\[
x_n = 2^{-w} h(\mathbf{z}_n).
\]

In R, the default pseudorandom number generator is the 32-bit 
*Mersenne Twister*, which generates integers in the range 

\[
\{0, 1, \ldots, 2^{32} -1\}.
\]

The state space is 

\[
\mathcal{Z} =  \{0, 1, \ldots, 2^{32} -1\}^{624},
\]

that is, a state is a 624 dimensional vector of 32-bit integers. 
The function $f$ is of the form 

\[
f(\mathbf{z}) = (z_2, z_3, \ldots, z_{623}, f_{624}(z_1, z_2, z_{m + 1})),
\]

for $1 \leq m < 624$, and $h$ is a function of $z_{624}$ only. 
The standard choice $m = 397$ is used in the R implementation.
The function $f_{624}$ is a bit complicated. It includes 
what is known as the *twist transformation*, and it requires 
additional parameters. The period of the generator 
is the astronomical number 

\[
2^{32 \times 624 - 31} - 1 = 2^{19937} - 1,
\]

which is a Mersenne prime. 
Moreover, all combinations of consecutive integers 
up to dimension 623 occur equally often in a period, and empirical tests of the 
generator demonstrate that it has good statistical properties, though it 
is known to fail some tests.

In R you can set the seed using the function `set.seed()` that takes an 
integer argument and produces an element in the state space. The argument 
given to `set.seed()` is not the actual seed, and `set.seed()` computes a 
valid seed for any pseudorandom number generator that R is using, 
whether it is the Mersenne Twister or not. Thus the use of `set.seed()` is 
the safe and recommended way of setting a seed. 

The actual seed (together with some additional information) can be accessed 
via the vector `.Random.seed`. Its first entry, `.Random.seed[1]`, encodes the 
pseudorandom number generator used as well as the generator for 
Gaussian variables and discrete uniform variables. This 
information is decoded by `RNGkind()`. 

```{r}
RNGkind()
```


For the Mersenne Twister, 
`.Random.seed[3:626]` contains the vector in the state space, while 
`.Random.seed[2]` contains the "current position" in the state vector. 
The implementation needs a position variable because it does 624 updates of the 
state vector at a time and then runs through those values sequentially 
before the next update. This is equivalent to but more efficient than
implementing the position shifts explicitly as in the definition of $f$ above. 


```{r seed}
set.seed(27112015)           # Computes a new seed from an integer
oldseed <- .Random.seed[-1]  # The actual seed
.Random.seed[1]              # Encoding of generators used (stays fixed)
.Random.seed[2]              # Start position after new seed is 624
tmp <- runif(1)
tmp
```

Every time a random number is generated, by `runif()` as above or by any other 
R function relying on the default pseudorandom number generator, the same underlying 
sequence of pseudorandom numbers is used, and the state vector stored in 
`.Random.seed` is updated accordingly.

```{r changeseed, dependson="seed"}
head(oldseed, 5)
head(.Random.seed[-1], 5)     # State vector and position updated
c(tmp, runif(1))
head(.Random.seed[-1], 5)     # State vector unchanged, position updated
```

Resetting the seed will restart the pseudorandom number generator with the 
same seed and result in the same sequence of random numbers.

```{r newseed, dependson=c("changeseed", "seed")}
set.seed(27112015)
head(.Random.seed[-1], 5)
head(oldseed, 5)             # Same as current .Random.seed
runif(1)                     # Same as tmp
```

Note that when using any of the standard R generators, any value of $0$ or $1$
returned by the underlying pseudorandom uniform generator is 
adjusted to be in $(0,1)$. Uniform random variables are thus 
guaranteed to be in $(0, 1)$. 

Some of the random number generators implemented in R use more than one 
pseudorandom number per variable. This is, for instance, the case when we simulate 
gamma distributed random variables. 

```{r gammaseed}
set.seed(27112015)
rgamma(1, 1)                 # A single gamma distributed random number
head(.Random.seed[-1], 5)    # Position changed to 2
rgamma(1, 1)                 # A single gamma distributed random number
head(.Random.seed[-1], 5)    # Position changed to 5
```

In the example above, the first gamma variable required two pseudorandom 
numbers, while the second required three pseudorandom numbers. The detailed 
explanation is given in Section \@ref(reject-samp), where it is shown 
how to generate random variables from the gamma distribution via 
rejection sampling. This requires as a minimum two pseudorandom numbers 
for every gamma variable generated. 

The development of high quality pseudorandom number generators is a research
field in itself. This is particularly true if one needs theoretical guarantees
for randomized algorithms or cryptographically secure generators. For scientific
computations and simulations, reproducibility and speed are more important than
cryptographic security---if the generator has acceptable statistical properties.
It is, nevertheless, not easy to invent a good generator, and the field is still
developing. For a generator to be seriously considered, its mathematical
properties should be well understood, and it should pass (most) tests in
standardized test suites such as
[TestU01](http://simul.iro.umontreal.ca/testu01/tu01.html), see @LEcuyer:2007.

R provides a couple of alternatives to the Mersenne Twister, 
see `?RNG`, but there is no compelling reason to switch to any of those for ordinary 
use. They are mostly available for historical reasons. 
One exception is the L'Ecuyer-CMRG generator, which is useful when
independent pseudorandom sequences are needed for parallel computations. 

Though the Mersenne Twister is a widely used pseudorandom number generator, 
it has some [known shortcomings](http://www.pcg-random.org/other-rngs.html)
[@Oneill:2018], see also [@Oneill:pcg2014]. 
In the following two sections we discuss alternatives to the standard 
pseudorandom number generators available in R. 

### Implementing a pseudorandom number generator

This section dives into low-level aspects of an efficient implementation 
of a pseudorandom number generator via Rcpp. This involves turning the 
generator into the default generator used by R. The implementation illustrates 
bit-level C/C++ programming techniques, some aspects of R's interface to 
compiled code, and how users can make compiled code available. However, 
the section can safely be skipped in a first reading, and other parts 
of the book do not rely explicitly on material in this section. 

The family of [shift-register generators](https://en.wikipedia.org/wiki/Xorshift)
and their variations are considered to be high quality alternatives to the 
Mersenne Twister, and to be both simpler and faster. At the time of writing 
such generators are, however, not available from the base R package. 

Shift-register generators are based on linear transformations of the bit 
representation of integers. Three particular transformations are typically 
composed; the $\mathrm{Lshift}$ and $\mathrm{Rshift}$ operators and the 
bitwise $\mathrm{xor}$ operator. Let $z = [z_{31}, z_{30}, \ldots, z_0]$ 
with $z_i \in \{0, 1\}$ denote the bit 
representation of a 32-bit (unsigned) integer $z$ (ordered from most significant
bit to least significant bit). That is, 

\[
z = z_{31} 2^{31} + z_{30} 2^{30} + \ldots + z_2 2^2 + z_1 2^{1} + z_0.
\]

Then the left shift operator is defined as 

\[
\mathrm{Lshift}(z) = [z_{30}, z_{29}, \ldots, z_0, 0],
\]

and the right shift operator is defined as 

\[
\mathrm{Rshift}(z) = [0, z_{31}, z_{30}, \ldots, z_1].
\]

The bitwise xor operator is defined as 

\[
\mathrm{xor}(z, z') = [\mathrm{xor}(z_{31},z_{31}') , 
\mathrm{xor}(z_{30}, z_{30}'), \ldots, \mathrm{xor}(z_0, z_0')]
\]

where $\mathrm{xor}(0, 0) = \mathrm{xor}(1, 1) = 0$ and 
$\mathrm{xor}(1, 0) = \mathrm{xor}(0, 1) = 1$. Thus a transformation could 
be of the form 

\[
\mathrm{xor}(z, \mathrm{Rshift}^2(z)) = [\mathrm{xor}(z_{31}, 0) , 
\mathrm{xor}(z_{30}, 0), \mathrm{xor}(z_{29}, z_{31}), \ldots, \mathrm{xor}(z_0, z_2)].
\]

One example of a shift-register based generator is Marsaglia's [xorwow](https://en.wikipedia.org/wiki/Xorshift#xorwow) 
algorithm, @Marsaglia:2003. In addition to the shift and xor operations, the output of 
this generator is perturbed by a sequence of integers with period $2^{32}$. 
The state space of the generator is 

\[
\{0, 1, \ldots, 2^{32} -1\}^{5}
\]

with 

\[
f(\mathbf{z}) = (z_1 + 362437 \ (\mathrm{mod}\ 2^{32}), f_1(z_5, z_2), z_2, z_3, z_4),
\]

and 

\[
h(\mathbf{z}) = 2^{-32} (z_1 + z_2).
\]

The number 362437 is Marsaglia's choice for generating what he calls a
Weyl sequence, but any odd number will do. The function $f_1$ is given as 

\begin{align*}
\overline{z} & = \mathrm{xor}(z, \mathrm{Rshift}^2(z)) \\
f_1(z, z') & = \mathrm{xor}(\overline{z}, \mathrm{xor}(z', \mathrm{xor}(\mathrm{Lshift}^4(z'), \mathrm{Lshift}(\overline{z})))).
\end{align*}

This may look intimidating, but all the operations are very elementary. 
Take the numbers $z = 123456$ and $z' = 87654321$, say. Then we can find their 
binary representations in R, in the bit order as above, as follows:

```{r inttobit}
z <- intToBits(123456) |> as.integer() |> rev()
zp <- intToBits(87654321) |> as.integer() |> rev()
cat("z  = ", z, "\n", "z' = ", zp, sep = "")
```

The intermediate value 
$\overline{z} = \mathrm{xor}(z, \mathrm{Rshift}^2(z))$ is computed as follows:

\[
\begin{array}{ll}
z & \texttt{00000000 00000001 11100010 01000000} \\
\mathrm{Rshift}^2(z) & \texttt{00000000 00000000 01111000 10010000} \\
\hline
\mathrm{xor} & \texttt{00000000 00000001 10011010 11010000}
\end{array}
\]

The value of $f_1(z, z')$ is then computed like this:

\[
\begin{array}{ll}
\mathrm{Lshift}^4(z') & \texttt{01010011 10010111 11111011 00010000} \\
\mathrm{Lshift}(\overline{z}) & \texttt{00000000 00000011 00110101 10100000} \\
\hline
\mathrm{xor} & \texttt{01010011 10010100 11001110 10110000} \\
z' & \texttt{00000101 00111001 01111111 10110001} \\
\hline
\mathrm{xor} & \texttt{01010110 10101101 10110001 00000001} \\
\overline{z} & \texttt{00000000 00000001 10011010 11010000} \\
\hline
 \mathrm{xor} & \texttt{01010110 10101100 00101011 11010001}
\end{array}
\]

We can convert back to a 32-bit integer using R,

```{r fzzp, dependson="inttobit"}
result <- "01010110101011000010101111010001" 
result <- strsplit(result, "")[[1]] |> rev() %>% as.integer()
packBits(result, "integer")
```

That is $f_1(123456, 87654321) = 1454123985$. 

The shift and xor operations are tedious to do by hand but extremely fast on 
modern computer architectures, and shift-register based generators are some 
of the fastest generators with good statistical properties.  

To make R use the xorwow generator we need to implement it as a user supplied 
generator. This requires writing the C code that implements the generator, 
compiling the code into a shared object file, loading it into 
R with the `dyn.load()` function, and finally calling `RNGkind("user")` 
to make R use this pseudorandom number generator. See `?Random.user`
for some details and an example. 

Using the Rcpp package, and `sourceCpp()`, in particular, is usually much preferred
over manual compiling and loading. However, in this case we need to make 
functions available to the internals of R rather than exporting functions to be 
callable from the R console. That is, nothing needs to be exported from C/C++. 
If nothing is exported, `sourceCpp()` will actually not load the shared object 
file, so we need to trick `sourceCpp()` to do so anyway. In the implementation
below we achieve this by simply exporting a direct interface to the xorwow generator.

```{Rcpp xorwow}
#include <Rcpp.h>
#include <R_ext/Random.h>

// The Random.h header file contains the function declarations 
// for the functions that R rely on internally for a user defined 
// generator, and it also defines the type Int32 as an unsigned int.

static Int32 z[5];     // The state vector 
static double res;
static int nseed = 5;  // Length of the state vector

// Implementation of xorwow from Marsaglia's "Xorshift RNGs" 
// modified so as to return a double in [0, 1). The '>>' and '<<' 
// operators in C are bitwise right and left shift operators, and 
// the caret, '^', is the xor operator. 
double * user_unif_rand()
{ 
  Int32 t = z[4];
  Int32 s = z[1];
  z[0] += 362437;
  z[4] = z[3];
  z[3] = z[2];
  z[2] = s;
  // Right shift t by 2, then bitwise xor between t and its shift
  t ^= t >> 2;  
  // Left shift t by 1 and s by 4, xor them, xor with s and xor with t
  t ^= s ^ (s << 4) ^ (t << 1);
  z[1] = t;
  res = (z[0] + t) * 2.32830643653869e-10;
  return &res;
}

// A seed initializer using Marsaglia's congruential PRNG
void  user_unif_init(Int32 seed_in) { 
  z[0] = seed_in;
  z[1] = 69069 * z[0] + 1;
  z[2] = 69069 * z[1] + 1;
  z[3] = 69069 * z[2] + 1;
  z[4] = 69069 * z[3] + 1;
}

// Two functions to make '.Random.seed' in R reflect the state vector
int * user_unif_nseed() { return &nseed; }
int * user_unif_seedloc() { return (int *) &z; }

// Wrapper to make 'user_unif_rand()' callable from R
double xor_runif() {
  return *user_unif_rand();
}

// This module exports two functions to be directly available from R.
// Note: if nothing is exported, `sourceCpp()` will not load the shared 
// object file generated by the compilation of the code, and 
// 'user_unif_rand()' will not become available to the internals of R. 
RCPP_MODULE(xorwow) {
  Rcpp::function(
    "xor_set.seed", 
    &user_unif_init, 
    "Seeds Marsaglia's xorwow"
  );
  Rcpp::function(
    "xor_runif", 
    &xor_runif, 
    "A uniform from Marsaglia's xorwow"
  );
}
```

We first test the direct interface to the xorwow algorithm.

```{r xor_direct, dependson="xorwow"}
xor_set.seed(3573076633)
xor_runif()
```

Then we set R's pseudorandom number generator to be our user supplied 
generator. 

```{r userRNG, dependson="xorwow", cache=FALSE}
default_prng <- RNGkind("user")
```

All R's standard random number generators will after the call `RNGkind("user")`
rely on the user provided generator, in this case the xorwow generator. 
Note that R does an "initial scrambling" of the argument given to `set.seed` 
before it is passed on to our user defined initializer. This 
scrambling turns 24102019 used below into 3573076633 used above. 

```{r userRNGtest, dependson="xorwow"}
set.seed(24102019)
.Random.seed[-1]  # The state vector as seeded
runif(1)          # As above since same unscrambled seed is used
.Random.seed[-1]  # The state vector after one update
```

The code above shows the state vector of the xorwow algorithm when seeded 
by the `user_unif_init()` function, and it also shows the 
update to the state vector after a single iteration of the xorwow algorithm. 

Though the xorwow algorithm is fast and simple, a benchmark study (not shown) 
reveals that using xorwow instead of the Mersenne Twister does not impact the 
runtime in a notable way when using, e.g., `runif()`. The generator is simply 
not the bottleneck. As the implementation 
of xorwow above is experimental and has not been thoroughly tested, we will 
not rely on it and thus reset the random number generator to its default value. 

```{r resetRNG, cache=FALSE}
# Resetting the generator to the default
RNGkind(default_prng[1])
```

### Pseudorandom number packages {#rng-packages}

Instead of implementing our own generator, as in the section above, we 
can benefit from the recent developments in pseudorandom number generators 
by turning to R packages such as the [dqrng](https://CRAN.R-project.org/package=dqrng)
package. It implements pcg64 from the [PCG family](http://www.pcg-random.org) of
generators as well as [Xoroshiro128+ and Xoshiro256+](http://prng.di.unimi.it)
that are shift-register algorithms. Xoroshiro128+ is the default and other
generators can be chosen using `dqRNGkind`. The usage of generators from dqrng 
is similar to the usage of base R generators.

```{r dqrng-test}
dqrng::dqset.seed(24102019)
dqrng::dqrunif(1)
```

Using the generators from dqrng does not interfere with the base R generators
as the state vectors are completely separated. 

In addition to uniform pseudorandom variables generated by `dqrunif()` the 
dqrng package can generate exponential (`dqrexp()`) and Gaussian (`dqrnorm()`) 
random variables as well as uniform discrete distributions (`dqsample()` and 
`dqsample.int()`). All based on the fast pseudorandom integer generators that 
the package includes. In addition, the package has a C++ interface that makes it 
possible to use its generators in compiled code as well. 

We benchmark some of the functions from the package against the base R alternatives. 

```{r dqrng_bench, echo=2}
oldopt <- options(digits = 2)
bench::mark(
  runif(1e6),
  dqrng::dqrunif(1e6),
  check = FALSE,
  relative = TRUE
)
options(digits = oldopt$digits)
```

As the benchmark above shows, `runif()` is about a factor $1.5$ slower 
than `dqrunif()` when generating one million variables. Thus for uniformly 
distributed random numbers there is not much of a runtime benefit to using 
the dqrng package. The other generators provided by the package show 
greater improvements over the base R generators. Generating a small subsample 
from a large set is particularly fast.

```{r dqrng_bench_norm, echo=FALSE, eval=FALSE, dependson="dqrng_bench"}
bench::mark(
  rnorm(1e4, 1, 0.5),
  dqrng::dqrnorm(1e4, 1, 0.5),
  check = FALSE,
  relative = TRUE
)
```

```{r dqrng_bench_exp, echo=FALSE, eval=FALSE, dependson="dqrng_bench"}
bench::mark(
  rexp(1e7, 2),
  dqrng::dqrexp(1e7, 2),
  check = FALSE,
  relative = TRUE
)
```

```{r dqrng_bench_sample, echo=2, dependson="dqrng_bench"}
oldopt <- options(digits = 2)
bench::mark(
  sample.int(1e6, 1000),
  dqrng::dqsample.int(1e6, 1000),
  check = FALSE,
  relative = TRUE
)
options(digits = oldopt$digits)
```

The benchmark shows that `sample.int()` is close to a factor $30$ slower 
than `dqsample.int()` when sampling one thousand integers (without replacement) 
from a million integers. We return to the use of `dqsample.int()` 
in Section \@ref(full-Rcpp).

## Transformation techniques {#transform}

A widely used technique for simulating from a target distribution of interest is to 
represent the distribution as a transformation of another, possibly simpler, 
distribution. Mathematically, a transformation is a map $T : \mathcal{Z} \to \mathbb{R}$, 
and if we can simulate the random variable $Z \in \mathcal{Z}$, 
then we can also simulate $X = T(Z).$ The following 
theorem gives one standard way of transforming uniformly distributed variables 
to any target distribution on the real line via the inverse distribution 
function. 

::: {.theorem #inverse-distribution-function}
If $F^{-1} : (0,1) \mapsto \mathbb{R}$
is the (generalized) inverse of a distribution function and $U$ is uniformly distributed 
on $(0, 1)$ then the distribution of

\[
  F^{-1}(U)
\]

has distribution function $F$.
:::


For a proof of Theorem \@ref(thm:inverse-distribution-function) see, e.g., 
Lemma 8.7 in [@legall2022measure]. It is easiest to use this theorem if 
we have an analytic formula for the inverse distribution function as 
in the following example. 

::: {.example #exponential-inverse .boxed}
The exponential distribution with rate parameter $\lambda > 0$ has distribution 
function $F(x) = 1 - e^{-\lambda x}$ for $x \geq 0$. To find its inverse we
solve the equation $F(x) = u$ for $u \in (0, 1)$ and get

\[
  F^{-1}(u) = - \frac{1}{\lambda} \log (1 - u). 
\]

The function `r_exp()` below is a direct implementation of simulation of 
exponential variables by transforming uniform variables by the inverse 
distribution function. Figure \@ref(fig:sim-exponential) confirms that 
the implementation simulates from the correct exponential distribution. 
:::
 
(ref:expsim) Left: Histogram of 1000 simulated data points from the exponential distribution with rate parameter $\lambda = 2$ and the corresponding density. Right: The empirical (black) and theoretical (blue) distribution functions. 

```{r sim-exponential, fig.cap="(ref:expsim)", fig.show='hold', out.width="49%"}
r_exp <- function(n, lambda) {
  u <- runif(n)
  - log(1 - u) / lambda
}

# Test
x <- r_exp(1000, 2)
hist(x, breaks = seq(0, 8, length.out = 30), prob = TRUE, xlim = c(0, 4))
curve(dexp(x, 2), 0, 4, col = "blue", lwd = 2, add = TRUE)
curve(pexp(x, 2), 0, 2, col = "blue", lwd = 2)
plot(ecdf(x), xlim = c(0, 2), add = TRUE, lwd = 2)
```

A curious detail is that if $U$ is uniformly distributed on $(0, 1)$, then $1 - U$ 
is uniformly distributed on $(0, 1)$. It is therefore possible to slightly 
simplify the implementation by replacing $F^{-1}$ by $- \log(u) / \lambda$, 
see Exercise \@ref(exr:exponential-sim).

Even if we do not have a simple analytic expression for the inverse distribution
function we might have an accurate approximation that is fast to evaluate, 
which can then be used for simulation. An interesting example is the Gaussian
distribution. The call `RNGkind()` in the previous section revealed that the
[default in
R](https://github.com/wch/r-source/blob/af7f52f70101960861e5d995d3a4bec010bc89e6/src/nmath/snorm.c#L265)
for generating observations from $\mathcal{N}(0,1)$ is inversion. That is, Theorem
\@ref(thm:inverse-distribution-function) is used to transform uniform random
variables with the inverse distribution function $\Phi^{-1}$. This function is,
however, non-standard, and R implements a [technical
approximation](https://github.com/wch/r-source/blob/af7f52f70101960861e5d995d3a4bec010bc89e6/src/nmath/qnorm.c#L52)
of $\Phi^{-1}$ via rational functions.

::: {.example #t-distribution .boxed}
Another good example of a distribution that we can simulate by transformation 
is the $t$-distribution. 
Let $Z = (Y, W) \in \mathbb{R} \times (0, \infty)$ with $Y \sim \mathcal{N}(0, 1)$ and 
$W \sim \chi^2_k$ independent. 

Define $T : \mathbb{R} \times (0, \infty) \to \mathbb{R}$ by 
$T(y,w) = y / \sqrt{w/k},$ then 

\[
  X = T(Y, W) = \frac{Y}{\sqrt{W/k}}
\]

has a $t$-distribution with $k$ degress of freedom. Exercise \@ref(exr:t-distribution-sim)
asks you to implement simulation from the $t$-distribution via the transformation 
above. This requires an algorithm 
that simulates from the Gaussian distribution
and an algorithm that simulates from the $\chi^2_k$ distribution (which is 
also a gamma distribution). 

The R function `rt()` implements simulation from a $t$-distribution by the
above transformation and by generating $W$ from a 
gamma distribution with shape parameter $k / 2$ and scale parameter $2$.
The simulation from a gamma distribution via rejection sampling is 
dealt with in Section \@ref(gammasim).
:::

## Exercises {#univariate:ex}

::: {.exercise #exponential-sim}
Implement simulation from the exponential distribution as in Example 
\@ref(exm:exponential-inverse) but replacing $1 - U$ by $U$. Test your 
implementation and benchmark it against the implementation in 
Example \@ref(exm:exponential-inverse), against `rexp()` and 
against `dqrng::dqrexp()` from the dqrng package.
:::

::: {.exercise #t-distribution-sim}
Implement simulation from the $t$-distribution as in Example 
\@ref(exm:t-distribution). Test your 
implementation and benchmark it against against `rt()`.
:::

::: {.exercise #laplace-sim}
Recall that the Laplace distribution has density 

\[
  f(x) = \frac{1}{2} e^{-|x|}
\]

for $x \in \mathbb{R}$. Implement simulation from the Laplace distribution 
by transforming a uniform random variable by the inverse distribution function.
Test the implementation
:::

::: {.exercise #laplace-sim-2}
If $X$ and $Y$ are independent and exponentially distributed with mean one, 
then $X - Y$ has a Laplace distribution (why?). Use this to implement 
simulation from the Laplace distribution. Benchmark this implementation 
together with the implementation from Exercise \@ref(exr:laplace-sim). 
:::
