---
output: html_document
editor_options:
  chunk_output_type: console
---

# Numerical optimization {#numopt}

In Chapter\ \@ref(opt-problem), we framed maximum-likelihood and penalized
estimation as the minimization of $H = -\ell$ or $H = -\ell + J$, where $J$ is a
penalty that does not depend on the data.
\index{negative log-likelihood!minimization}\index{penalty function} In this
chapter, we introduce algorithms for actually computing those minimizers.
Throughout we assume $H : \Theta \to \mathbb{R}$ is twice continuously
differentiable on an open set $\Theta \subseteq \mathbb{R}^p$.

When $H$ is convex, every stationary point is a global minimizer
(Section\ \@ref(opt-conditions)) and the algorithms considered in this section
produce a certified optimum. When it is not, they produce a local minimizer or
merely a stationary point. That does not, however, mean that the solutions are
meaningless. @Kiefer:1978, for instance, showed that the Gaussian-mixture
likelihood of Section\ \@ref(Gauss-mix-ex) is unbounded above, yet $\nabla
H(\theta) = 0$ admits a consistent sequence of solutions as $N \to \infty$.
Convexity is convenient but not essential.

Throughout the rest of this chapter, two recurrent themes will emerge. First,
our goal is typically not the minimizer of $H$ for a particular dataset---what
matters is the statistical estimate it represents. Driving an optimizer to
machine precision is rarely worth the computational effort when the statistical
uncertainty of the estimate is orders of magnitude larger, and our budget is
then better spent exploring many models and starting values and quantifying
parameter uncertainty. Second, the runtime of every algorithm in this chapter is
dominated by the cost of evaluating $H$ and its derivatives. For
exponential-family models, that cost collapses to a one-time
sufficient-statistic computation (Section\ \@ref(parametrization)), making each
iteration $O(p)$ rather than $O(Np)$. We will exploit this structure repeatedly.

## Iterative algorithms

Optimization algorithms are typically *iterative*: from an initial value
$\theta_0 \in \Theta$ they generate a sequence $\theta_1, \theta_2, \ldots$ via
a step map $\Phi : \Theta \to \Theta$ with

\[
\theta_n = \Phi(\theta_{n-1}).
\]

We hope that $\theta_n \to \theta^\star$ for some minimizer
$\theta^\star \in \argmin_\theta H(\theta)$. When this happens, continuity of
$\Phi$ forces $\Phi(\theta^\star) = \theta^\star$, so the limit is a *fixed
point* of the step map. We thus specify $\Phi$ so that its fixed points are
exactly the points we want to converge to, typically the stationary points of
$H$, and then verify that the iterates do converge.

Less than full convergence is often the best we can guarantee. The global
minimizer may not exist or may not be unique, in which case the limit is defined
only up to that ambiguity. The iterates may converge only to a *local* minimizer
rather than the global one. And $\theta_n$ may not converge at all even if
$H(\theta_n)$ converges to a (local) minimum, as the iterates can wander along a
flat valley while the function value stabilizes.

Most of the algorithms in this chapter are unified via a common feature: each
step moves $\theta_n$ in a *descent direction* $\rho_n \in \mathbb{R}^p$ scaled
by a *step size* $t_n > 0$:

\[
\theta_{n+1} = \theta_n + t_n \rho_n.
\]

The direction $\rho_n$ is called a descent direction at $\theta_n$ when

\[
\rho_n^\top \nabla H(\theta_n) < 0
\]

whenever $\theta_n$ is not a stationary point. The first-order Taylor expansion

\[
H(\theta_n + t \rho_n) = H(\theta_n) + t \rho_n^\top \nabla H(\theta_n) + o(t)
\]

then guarantees $H(\theta_n + t \rho_n) < H(\theta_n)$ for sufficiently small
$t > 0$. The algorithms in this chapter differ in how they choose $\rho_n$ and
$t_n$. The simplest choice is gradient descent (Section\ \@ref(grad-descent)),
which moves opposite the gradient, taking $\rho_n = -\nabla H(\theta_n)$.
Newton's method (Section\ \@ref(Newton)) uses the curvature of $H$ (the Hessian)
to correct the gradient direction, resulting in
$\rho_n = -\nabla^2 H(\theta_n)^{-1} \nabla H(\theta_n)$, provided the Hessian
is positive definite. Quasi-Newton methods replace the Hessian inverse with a
cheaper surrogate. Step sizes can be fixed, chosen by line search, or obtained
by minimizing exactly along the ray $t \mapsto H(\theta_n + t \rho_n)$.

A sequence generated this way is called a *descent sequence* if

\[
H(\theta_0) \geq H(\theta_1) \geq H(\theta_2) \geq \ldots,
\]

and a *strict* descent sequence if the inequalities are sharp at every
non-stationary $\theta_n$. The sequence $H(\theta_n)$ converges whenever $H$ is
bounded below on the level set
$\mathrm{lev}(\theta_0) = \{\theta \in \Theta \mid H(\theta) \leq H(\theta_0)\}$,
which holds in particular when that level set is compact. But descent alone does
not imply convergence to a stationary point---an algorithm can take ever-smaller
steps and stall short of any minimizer. What we need is *sufficient* descent at
each step, controlled by a quantitative lower bound on
$H(\theta_n) - H(\theta_{n+1})$ in terms of the gradient. The rest of the
chapter is largely devoted to establishing such bounds and deriving convergence
rates from them.

## Gradient descent {#grad-descent}

Gradient descent is the quintessential first-order optimization algorithm. At
each iteration we take $\rho_n = -\nabla H(\theta_n)$, the steepest descent
direction, pick a step size $t_n > 0$, and update our current best guess of
$\theta$ (Algorithm\ 1).

```{algorithm, gd-basic}
\begin{algorithm}
  \caption{Gradient descent}
  \begin{algorithmic}
    \Require step size $t > 0$
    \Repeat
        \State $\theta \gets \theta - t \nabla H(\theta)$
    \Until{convergence}
    \Return $\theta$
  \end{algorithmic}
\end{algorithm}
```

Geometrically, the step replaces $\theta_n$ with the point
$\theta_n - t \nabla H(\theta_n)$ on the parameter axis: a move opposite to the
local slope, by an amount proportional to its magnitude.
Figure\ \@ref(fig:gd-step-1d) shows a single step on a one-dimensional convex
$H$.

```{r}
#| label: gd-step-1d
#| echo: false
#| fig-width: 5
#| fig-height: 3
#| fig-cap: "One step of gradient descent. The tangent at $\\theta_n$ has slope
#|   $\\nabla H(\\theta_n)$; the update moves $\\theta_n$ by $-t \\nabla H(\\theta_n)$
#|   along the parameter axis, towards lower $H$."
#| warning: false

f <- function(x) 0.4 * (x - 2)^2 + 0.5
f_prime <- function(x) 0.8 * (x - 2)

theta_n <- -0.5
t <- 0.6
theta_n1 <- theta_n - t * f_prime(theta_n)

x_grid <- seq(-1.5, 4, length.out = 200)
curve_df <- data.frame(x = x_grid, y = f(x_grid))

tan_x <- c(theta_n - 0.9, theta_n + 0.9)
tan_df <- data.frame(
  x = tan_x,
  y = f(theta_n) + f_prime(theta_n) * (tan_x - theta_n)
)

pts <- data.frame(
  x = c(theta_n, theta_n1),
  y = c(f(theta_n), f(theta_n1)),
  label = c("theta[n]", "theta[n+1]")
)

ggplot(curve_df, aes(x, y)) +
  geom_line() +
  geom_line(data = tan_df, linetype = "dashed", color = "steelblue") +
  geom_segment(
    data = pts,
    aes(x = x, xend = x, y = 0, yend = y),
    linetype = "dotted"
  ) +
  geom_segment(
    aes(x = theta_n, xend = theta_n1, y = -0.15, yend = -0.15),
    arrow = arrow(length = unit(0.15, "cm"), type = "closed")
  ) +
  annotate(
    "text",
    x = (theta_n + theta_n1) / 2,
    y = -0.4,
    label = "-t * nabla * H(theta[n])",
    parse = TRUE
  ) +
  geom_point(data = pts) +
  geom_text(
    data = pts,
    aes(label = label),
    parse = TRUE,
    nudge_y = 0.3,
    nudge_x = c(0.15, 0.15)
  ) +
  labs(x = expression(theta), y = expression(H(theta))) +
  coord_cartesian(ylim = c(-0.6, NA)) +
  theme_schematic()
```

But how do we set $t$? If we set it too large, then we risk overshooting the
target, possibly even diverging. If we set it too small, then we might never
reach the optimum within our time budget. The answer hinges on the *smoothness*
of $H$. Recall from Section\ \@ref(smoothness) that $H$ is *$L$-smooth* on
$\Theta$ when $\nabla^2 H(\theta) \preceq L I$ for all $\theta \in \Theta$, and
that the descent lemma (Lemma\ \@ref(lem:descent-lemma)) then guarantees

\[
H(\theta') \leq H(\theta) + \nabla H(\theta)^\top (\theta' - \theta) + \frac{L}{2} \|\theta' - \theta\|_2^2
\]

for all $\theta, \theta' \in \Theta$. Setting
$\theta' = \theta - t \nabla H(\theta)$ in this inequality yields

\begin{align}
H(\theta - t \nabla H(\theta)) & \leq H(\theta) - t \left(1 - \frac{tL}{2}\right) \|\nabla H(\theta)\|_2^2.
(\#eq:gd-descent-bound)
\end{align}

The right-hand side is minimized over $t > 0$ at $t = 1/L$, where the bound
collapses to

\[
H(\theta - L^{-1} \nabla H(\theta)) \leq H(\theta) - \frac{1}{2L} \|\nabla H(\theta)\|_2^2.
\]

The gradient descent step we will analyze in this section is correspondingly

\begin{align}
\theta_{n+1} & = \theta_n - \frac{1}{L} \nabla H(\theta_n).  (\#eq:gd-step)
\end{align}

Every iteration decreases $H$ by at least
$(2L)^{-1} \|\nabla H(\theta_n)\|_2^2$, so $(\theta_n)$ is a strict descent
sequence unless it hits a stationary point. The corresponding step map

\[
\Phi(\theta) = \theta - \frac{1}{L} \nabla H(\theta)
\]

has $\theta$ as a fixed point if and only if $\nabla H(\theta) = 0$, so the
fixed points of $\Phi$ are exactly the stationary points of $H$. Knowing $L$ in
advance is rarely realistic in practice, and Section\ \@ref(stop-crit) and the
line-search material that follows show how a backtracking step size adapts the
step length without requiring it explicitly.

Figure\ \@ref(fig:gd-step-sizes) illustrates the three regimes on a simple step
near $1/L$ converges in a handful of iterations; a step approaching the step
near $1/L$ converges in a handful of iterations; a step approaching the
stability boundary $2/L$ (where the descent inequality
\@ref(eq:gd-descent-bound) becomes vacuous) oscillates across the steep
direction. For $t > 2/L$ the iterates diverge.

```{r}
#| label: gd-step-sizes
#| echo: false
#| fig-width: 7
#| fig-height: 3
#| fig-cap: "Gradient descent on $H(\\theta) = (\\theta_1^2 + 4\\theta_2^2)/2$ (so $L = 4$) for three step sizes, starting from $\\theta_0 = (-2.5, 1.5)$. Contours are level sets of $H$; dots are the iterates. A step much smaller than $1/L$ descends slowly; a step near $1/L$ converges efficiently; a step near the stability boundary $2/L = 0.5$ oscillates across the steep direction."

library(patchwork)

A <- diag(c(1, 4))
grad_H <- function(theta) drop(A %*% theta)

gd_path <- function(t, theta0, n_steps) {
  path <- matrix(0, n_steps + 1, 2)
  path[1, ] <- theta0
  for (i in 1:n_steps) {
    path[i + 1, ] <- path[i, ] - t * grad_H(path[i, ])
  }
  data.frame(theta1 = path[, 1], theta2 = path[, 2])
}

theta0 <- c(-2.5, 1.5)
n_steps <- 12
grid <- expand.grid(
  theta1 = seq(-3, 1, length.out = 120),
  theta2 = seq(-2, 2, length.out = 120)
)
grid$H <- 0.5 * (grid$theta1^2 + 4 * grid$theta2^2)

make_panel <- function(t, title) {
  path <- gd_path(t, theta0, n_steps)
  ggplot(grid, aes(theta1, theta2)) +
    geom_contour(aes(z = H), color = "grey70", bins = 8) +
    geom_path(data = path) +
    geom_point(data = path, size = 1) +
    labs(
      title = title,
      x = expression(theta[1]),
      y = expression(theta[2])
    )
}

make_panel(0.05, "t = 0.05") +
  make_panel(0.20, "t = 0.20") +
  make_panel(0.45, "t = 0.45") +
  plot_layout(axes = "collect")
```

### Convergence rates

Two convergence-rate results follow from the descent inequality
\@ref(eq:gd-descent-bound), one assuming only convexity (Theorem
\@ref(thm:gd-sublinear)) and one assuming strong convexity
(Theorem\ \@ref(thm:gd-linear)).

::: {.theorem #gd-sublinear}
Let $H : \mathbb{R}^p \to \mathbb{R}$ be convex and $L$-smooth, and suppose $H$
attains its minimum at some $\theta^\star$. The gradient descent iterates
\@ref(eq:gd-step) starting from $\theta_0 \in \mathbb{R}^p$ satisfy

\[
H(\theta_n) - H(\theta^\star) \leq \frac{L \|\theta_0 - \theta^\star\|_2^2}{2n}
\]

for every $n \geq 1$.
:::

::: {.proof .boxed}
Apply \@ref(eq:gd-descent-bound) at $\theta = \theta_i$ with $t = 1/L$ to get

\[
H(\theta_{i+1}) \leq H(\theta_i) - \frac{1}{2L} \|\nabla H(\theta_i)\|_2^2.
\]

By the first-order characterization of convexity (Section\ \@ref(convexity)),

$$
H(\theta^\star) \geq H(\theta_i) + \nabla H(\theta_i)^\top (\theta^\star - \theta_i),
$$

so

$$
H(\theta_i) - H(\theta^\star) \leq \nabla H(\theta_i)^\top (\theta_i - \theta^\star).
$$

Combining with the previous display,

\begin{align*}
  H(\theta_{i+1}) - H(\theta^\star) & \leq \nabla H(\theta_i)^\top (\theta_i - \theta^\star) - \frac{1}{2L} \|\nabla H(\theta_i)\|_2^2 \\
  & = \frac{L}{2} \left( \|\theta_i - \theta^\star\|_2^2 - \|\theta_i - \theta^\star - L^{-1} \nabla H(\theta_i)\|_2^2 \right) \\
  & = \frac{L}{2} \left( \|\theta_i - \theta^\star\|_2^2 - \|\theta_{i+1} - \theta^\star\|_2^2 \right),
\end{align*}

where the second line completes the square. Summing from $i = 0$ to $n - 1$
gives us

\[
\sum_{i=0}^{n-1} \left( H(\theta_{i+1}) - H(\theta^\star) \right) \leq \frac{L}{2} \|\theta_0 - \theta^\star\|_2^2.
\]

The descent property ensures $H(\theta_n) - H(\theta^\star)$ is decreasing in
$n$, so the left-hand side is at least $n (H(\theta_n) - H(\theta^\star))$.
:::

The suboptimality $H(\theta_n) - H(\theta^\star)$ thus decays like $1/n$. This
is *sublinear* convergence: the number of correct digits in $H(\theta_n)$ grows
like $\log n$. Doubling the accuracy requires doubling the iteration count, a
slow rate when many digits are needed, but one that places remarkably weak
demands on $H$ (just convexity and smoothness).

Adding strong convexity sharpens the picture dramatically. If $H$ is
$m$-strongly convex in addition to $L$-smooth, then
Lemma\ \@ref(lem:strong-convex-suboptimality) gives the two-sided bound

\[
\frac{m}{2} \|\theta - \theta^\star\|_2^2 \leq H(\theta) - H(\theta^\star) \leq \frac{1}{2m} \|\nabla H(\theta)\|_2^2.
\]

The right inequality says that suboptimality is controlled by gradient norm,
which combines with the per-step decrease of
$(2L)^{-1} \|\nabla H(\theta_n)\|_2^2$ to yield a *contraction* in
suboptimality.

::: {.theorem #gd-linear}
Let $H : \mathbb{R}^p \to \mathbb{R}$ be $m$-strongly convex and $L$-smooth with
$0 < m \leq L$, and let $\theta^\star$ be its unique minimizer. The gradient
descent iterates \@ref(eq:gd-step) satisfy

\[
H(\theta_{n+1}) - H(\theta^\star) \leq \left( 1 - \frac{m}{L} \right) \left( H(\theta_n) - H(\theta^\star) \right),
\]

and therefore

\[
H(\theta_n) - H(\theta^\star) \leq \left( 1 - \frac{1}{\kappa} \right)^n \left( H(\theta_0) - H(\theta^\star) \right),
\]

where $\kappa = L/m$ is the condition number from Section\ \@ref(smoothness).
:::

::: {.proof .boxed}
The strong-convexity bound rearranges to

$$
\|\nabla H(\theta_n)\|_2^2 \geq 2m (H(\theta_n) - H(\theta^\star)).
$$

Substituting into the per-step decrease,

\begin{align*}
  H(\theta_{n+1}) & \leq H(\theta_n) - \frac{1}{2L} \|\nabla H(\theta_n)\|_2^2 \\
  & \leq H(\theta_n) - \frac{m}{L} (H(\theta_n) - H(\theta^\star)).
\end{align*}

Subtracting $H(\theta^\star)$ from both sides gives the one-step contraction;
iterating gives the geometric rate.
:::

Convergence at a geometric rate is called *linear* convergence in the
numerical-optimization literature, because the number of correct digits in
$H(\theta_n) - H(\theta^\star)$ grows linearly in $n$. The contraction factor
$1 - 1/\kappa$ depends only on the condition number. Well-conditioned problems
($\kappa$ close to $1$) converge in ewiterations. Ill-conditioned problems
($\kappa$ large) crawl slowly towards convergence with the iterates tracing a
long zigzag along the floor of a narrow ravine. In Section\ \@ref(smoothness),
we provided the geometric interpretation: $\kappa$ is the aspect ratio of the
level sets of $H$. Newton's method (Section\ \@ref(Newton)) corrects for the
conditioning at the cost of solving a linear system per iteration.

Both theorems assume $\Theta = \mathbb{R}^p$. The arguments go through on any
open convex $\Theta$ provided the iterates remain in $\Theta$, but ensuring that
under arbitrary constraints requires projection or barrier methods rather than
the plain step\ \@ref(eq:gd-step).

### Measuring convergence in practice

The theoretical rates above are *worst case* bounds. The actual rate on any
given problem can be much better. The standard way to estimate it empirically is
to run the algorithm for $M$ iterations, such that
$\theta_M \approx \theta^\star$ at machine precision and then plot
$\log(H(\theta_n) - H(\theta_M))$ against $n$. Linear convergence with rate $r$
means

\[
\log(H(\theta_n) - H(\theta_M)) \approx n \log r + \text{const}
\]

for sufficiently large $n$, so the slope of a linear fit to the late iterations
estimates $\log r$. Sublinear convergence shows up as a decay slower than any
straight line---superlinear convergence as a steepening slope.

The same idea can be applied to $\theta_n$ or to $\|\nabla H(\theta_n)\|$, the
latter being especially convenient because the limit is known to be zero, so we
can monitor it without waiting for the algorithm to finish.

Per-iteration rates are useful for analyzing a single algorithm but not for
comparing different algorithms, which may spend wildly different amounts of time
per iteration. Throughout the chapter we will also report *per-second*
convergence rates by plotting against measured runtime rather than iteration
count. If one iteration takes $\delta$ seconds the per-second rate is
$r^{1/\delta}$; in practice we estimate $\delta$ as $t_M / M$ for $M$ large.

## Step size and stopping criteria

The convergence theorems of Section\ \@ref(grad-descent) used the fixed step
$t = 1/L$, which requires knowing the smoothness (Lipschitz) constant $L$. In
practice $L$ is rarely available analytically, since it may vary across the
parameter space, and a generic Lipschitz bound is often so pessimistic that
fixed-step iterations slow to almost a stop. Practical implementations therefore
often choose the step length adaptively at each iteration via a *line search*.
The general step is

\[
\theta_{n+1} = \theta_n + \gamma_n \rho_n
\]

for a descent direction $\rho_n$ and a step length $\gamma_n > 0$ that is
allowed to depend on $\theta_n$ and $\rho_n$. In this section we introduce
backtracking line search, which we use throughout the rest of the chapter, and
then turn to the question of when to terminate our algorithm.

### Line search {#line-search}

Given $\theta_n$ and a descent direction $\rho_n$, restrict $H$ to the ray
emanating from $\theta_n$ in the direction $\rho_n$:

\[
h(\gamma) = H(\theta_n + \gamma \rho_n), \qquad \gamma \in [0, \infty).
\]

Then $h'(\gamma) = \nabla H(\theta_n + \gamma \rho_n)^\top \rho_n$ and the
descent-direction property gives $h'(0) = \nabla H(\theta_n)^\top \rho_n < 0$,
so $h$ is decreasing at $0$. The *exact line search* strategy picks
$\gamma_n = \argmin_{\gamma > 0} h(\gamma)$, which delivers the maximal possible
descent along the ray. It is rarely worth the compute unless the inner
minimization admits a closed form.

Inexact line search settles for *sufficient* descent. Fix $c \in (0, 1)$. The
*Armijo condition*\index{Armijo condition}

\[
h(\gamma) \leq h(0) + c \gamma h'(0)
\]

requires $h$ to descend by at least the fraction $c$ of the linear prediction
along the tangent at $0$. Since $h'(0) < 0$ the right-hand side sits below
$h(0)$, and the condition reduces to one evaluation of $H$. For any
$c \in (0, 1)$ the Armijo condition holds on some interval $[0, \varepsilon)$,
so a small enough step always works. Figure\ \@ref(fig:armijo) shows the
condition geometrically: $h$ must lie at or below the lower dashed line
$h(0) + c \gamma h'(0)$ to be accepted, while the steeper tangent
$h(0) + \gamma h'(0)$ is ceiling. Choosing $c$ close to $0$ asks for very little
descent and admits most steps, whilst choosing $c$ close to $1$ asks fr a almost
full linear prediction of descent and rules out most steps.

```{r}
#| label: armijo
#| echo: false
#| fig-width: 4
#| fig-height: 3.2
#| fig-align: "center"
#| fig-cap: "The Armijo condition. Solid: the restriction $h(\\gamma) = H(\\theta + \\gamma \\rho)$ along a descent direction $\\rho$. Dashed orange: the tangent $h(0) + \\gamma h'(0)$, the linear prediction of descent at rate $h'(0)$. Dashed blue: the Armijo line $h(0) + c \\gamma h'(0)$ with $c = 0.35$. The shaded interval marks the values of $\\gamma$ for which the Armijo condition $h(\\gamma) \\leq h(0) + c \\gamma h'(0)$ holds."

h_fun <- function(g) 0.5 * (g - 1.3)^2 - 0.4
h0 <- h_fun(0)
hp0 <- -1.3
c_arm <- 0.35

g_grid <- seq(-0.2, 2.4, length.out = 300)
df_h <- data.frame(g = g_grid, y = h_fun(g_grid))
df_tan <- data.frame(g = g_grid, y = h0 + g_grid * hp0)
df_arm <- data.frame(g = g_grid, y = h0 + c_arm * g_grid * hp0)
acc_max <- max(g_grid[df_h$y - df_arm$y <= 0])

ggplot() +
  annotate(
    "rect",
    xmin = 0,
    xmax = acc_max,
    ymin = -Inf,
    ymax = Inf,
    fill = "grey88",
    alpha = 0.7
  ) +
  geom_line(data = df_h, aes(g, y), linewidth = 0.7) +
  geom_line(
    data = df_tan,
    aes(g, y),
    linetype = "dashed",
    color = "darkorange",
    linewidth = 0.6
  ) +
  geom_line(
    data = df_arm,
    aes(g, y),
    linetype = "dashed",
    color = "steelblue",
    linewidth = 0.6
  ) +
  annotate("point", x = 0, y = h0, size = 2) +
  annotate(
    "text",
    x = 0.05,
    y = h0 + 0.18,
    label = "h(0)",
    hjust = 0,
    size = 3.6
  ) +
  annotate(
    "text",
    x = 2.05,
    y = h_fun(2.05) + 0.20,
    label = "h(gamma)",
    parse = TRUE,
    hjust = 0.5,
    size = 3.6
  ) +
  annotate(
    "text",
    x = 1.65,
    y = h0 + 1.95 * hp0 - 0.15,
    label = "h(0) + gamma * h*minute*(0)",
    parse = TRUE,
    hjust = 0.5,
    vjust = 1,
    size = 3.3,
    color = "darkorange"
  ) +
  annotate(
    "text",
    x = 1.85,
    y = h0 + c_arm * 1.95 * hp0 - 0.18,
    label = "h(0) + c * gamma * h*minute*(0)",
    parse = TRUE,
    hjust = 0.5,
    vjust = 1,
    size = 3.3,
    color = "steelblue"
  ) +
  annotate(
    "text",
    x = acc_max / 2,
    y = -2.0,
    label = "Armijo-acceptable",
    size = 3.1,
    color = "grey30"
  ) +
  coord_cartesian(xlim = c(-0.3, 2.4), ylim = c(-2.8, 0.9), expand = FALSE) +
  labs(x = expression(gamma), y = NULL) +
  theme_schematic()
```

The Armijo criterion alone can yield arbitrarily small step sizes, which can
stall the algorithm venathough we are far from stationary point. The *curvature
condition*

\[
h'(\gamma) \geq \tilde{c}\, h'(0)
\]

for some $\tilde{c} \in (c, 1)$ forces the slope at $\gamma$ to be less negative
than at $0$, ruling out the smallest step sizes. Together the two requirements
form the *Wolfe conditions*.\index{Wolfe conditions} They can always be
satisfied when $h$ is bounded below, and under a mild angle condition on
$\rho_n$ they ensure $\|\nabla H(\theta_n)\|_2 \to 0$.

Most practical implementations replace the curvature condition with a geometric
search. *Backtracking line search*\index{backtracking line search} (Algorithm 2)
picks a starting step $\gamma_0 > 0$ and a shrinkage factor $d \in (0, 1)$, then
tests the sequence

\[
\gamma_0, \, d \gamma_0, \, d^2 \gamma_0, \, d^3 \gamma_0, \ldots
\]

stopping at the first $\gamma = d^k \gamma_0$ satisfying the Armijo condition.
The algorithm always tries $\gamma_0$ first and only shrinks when forced, so the
smallest step sizes are excluded by design rather than by an explicit curvature
bound.

```{algorithm, backtracking}
\begin{algorithm}
  \caption{Backtracking line search}
  \begin{algorithmic}
    \Require descent direction $\rho$, initial step $\gamma_0 > 0$, shrinkage $d \in (0,1)$, Armijo constant $c \in (0,1)$
    \State $\gamma \gets \gamma_0$
    \While{$H(\theta + \gamma \rho) > H(\theta) + c \gamma \nabla H(\theta)^\top \rho$}
        \State $\gamma \gets d \gamma$
    \EndWhile
    \Return $\gamma$
  \end{algorithmic}
\end{algorithm}
```

Three constants control the search: the initial step $\gamma_0$, the shrinkage
factor $d$, and the Armijo slope $c$. Typical defaults are $d \in [0.5, 0.9]$
and $c \in [10^{-4}, 0.1]$. For Newton's method (Section\ \@ref(Newton)) the
choice $\gamma_0 = 1$ is a natural choice. For plain gradient descent there is
no universal default and $\gamma_0$ has to be tuned to the scale of the problem.

### Stopping criteria {#stop-crit}

No algorithm can run forever and every implementation needs a rule for when to
terminate. But choosing an appropriate one is notoriously difficult. Next, we
will provide four common stopping criter a, each of which is easy to implement
and common in practice, but none of which actually guarantees convergence to a
minimum. We also note that each should be combined with a *maximum iteration
count* as a safeguard against infinite loops.

::: {.boxed}
*Maximum number of iterations.* Stop when

\[
n = M
\]

for a fixed maximum number of iterations $M$. The simplest criterion: it
guarantees termination but says nothing about proximity to a minimum. Useful
mainly as a safety net combined with one of the criteria below.
:::

The remaining three criteria depend on a *tolerance parameter* $\varepsilon > 0$
that plays a different role in each. They can be used individually or in
combination, but none of them (nor their combination) implies convergence. One
nevertheless says that an algorithm "has converged" when its stopping criterion
is met, with the understanding that this is shorthand.

::: {.boxed}
*Small relative change.* Stop when

\[
\|\theta_n - \theta_{n-1}\| \leq \varepsilon (\|\theta_n\| + \varepsilon).
\]

A step that barely moves $\theta_n$ suggests proximity to a fixed point of the
step map. The relative form makes the criterion invariant to a global rescaling
of the parameters, and the additive $\varepsilon$ keeps the right-hand side from
collapsing when $\|\theta_n\|$ is near zero. Its main practical merit is that it
does not require evaluating $H$, which makes it useful for algorithms where $H$
is expensive or unavailable, such as the EM algorithm of Chapter\ \@ref(em). The
deficit is that a small step has many causes besides convergence, and does not
imply that either $\|\theta_n - \theta^\star\|$ or
$H(\theta_n) - H(\theta^\star)$ is small.
:::

::: {.boxed}
*Small relative descent.* Stop when

\[
H(\theta_{n-1}) - H(\theta_n) \leq \varepsilon (|H(\theta_n)| + \varepsilon).
\]

Sensible only for descent algorithms. The relative form neutralizes a global
rescaling of $H$, and the additive $\varepsilon$ handles
$H(\theta_n) \approx 0$. A conventional choice is $\varepsilon \approx 10^{-8}$,
often the square root of machine epsilon, a folk default with weak theoretical
backing. The deficit mirrors that of the previous criterion: small descent can
occur in flat regions of $\Theta$ without proximity to a minimizer.
:::

::: {.boxed}
*Small gradient.* Stop when

\[
\|\nabla H(\theta_n)\| \leq \varepsilon.
\]

This criterion measures whether $\theta_n$ is close to being a stationary point,
not whether it is close to a stationary point. A small gradient norm is no
guarantee that $\|\theta_n - \theta^\star\|$ or $H(\theta_n) - H(\theta^\star)$
is small. The criterion requires $\nabla H$ at each iteration. The choice of
norm matters: when the gradient coordinates differ by orders of magnitude, a
weighted norm (or a rescaling of the parameters) is more informative than the
plain $\ell_2$ norm.
:::

The picture improves under strong convexity. If $H$ is $m$-strongly convex,
Lemma\ \@ref(lem:strong-convex-suboptimality) gives

\[
\frac{m}{2} \|\theta_n - \theta^\star\|_2^2 \leq H(\theta_n) - H(\theta^\star) \leq \frac{1}{2m} \|\nabla H(\theta_n)\|_2^2,
\]

so a small gradient norm does imply both proximity to $\theta^\star$ and small
suboptimality, with constants in $m$. Strong convexity is the exception rather
than the rule, but when it holds the small-gradient criterion comes with a
rigorous interpretation.

A more general option uses a *lower envelope*. Suppose a continuous
$\tilde{H} : \Theta \to \mathbb{R}$ satisfies
$H(\theta^\star) = \tilde{H}(\theta^\star)$ and
$H(\theta^\star) \geq \tilde{H}(\theta)$ for all $\theta \in \Theta$ (or for all
$\theta \in \mathrm{lev}(\theta_0)$ when the algorithm is a descent algorithm).
Then

\[
0 \leq H(\theta_n) - H(\theta^\star) \leq H(\theta_n) - \tilde{H}(\theta_n),
\]

and the right-hand side gives a computable upper bound on the suboptimality.
Convex duality theory supplies such an envelope for many convex optimization
problems; in those cases a *duality-gap* criterion can replace the *ad hoc*
tolerances above with a quantitative one. Treating duality properly is beyond
our scope, but it is worth knowing that better stopping rules exist when problem
structure allows.

### Implementing gradient descent

We implement gradient descent with backtracking below as the function
`grad_desc()`. For gradient descent, the sufficient descent condition amounts to
choosing the smallest $k \geq 0$ such that

\[
H(\theta_{n} - d^k \gamma_0 \nabla H(\theta_{n})) \leq H(\theta_n) -  cd^k \gamma_0 \|\nabla H(\theta_{n})\|_2^2.
\]

The `grad_desc()` function takes the starting point, $\theta_0$, the objective
function, $H$, and its gradient, $\nabla H$, as arguments. The four parameters
$d$, $c$, $\gamma_0$ and $\varepsilon$ that control the algorithm can also be
specified as additional arguments, but are given some reasonable default values.
The implementation uses the *squared* norm of the gradient as a stopping
criterion, but it also has a maximum number of iterations as a safeguard. If the
maximum number is reached, a warning is printed.

Finally, we include a callback argument (the `cb` argument). If a function is
passed to this argument, it will be evaluated in each iteration of the
algorithm. This gives us the possibility of logging or printing values of
variables during evaluation, which can be highly useful for understanding the
inner workings of the algorithm. Monitoring or logging intermediate values
during the evaluation of code is referred to as *tracing*.

```{r}
#| label: gd

grad_desc <- function(
  par,
  H,
  gr,
  d = 0.8,
  c = 0.1,
  gamma0 = 0.01,
  epsilon = 1e-4,
  maxit = 1000,
  cb = NULL
) {
  for (n in 1:maxit) {
    value <- H(par)
    grad <- gr(par)
    h_prime <- sum(grad^2)

    if (!is.null(cb)) {
      cb()
    }

    # Convergence criterion based on gradient norm
    if (h_prime <= epsilon) {
      break
    }

    gamma <- gamma0

    # Proposed descent step
    par1 <- par - gamma * grad

    # Backtracking while descent is insufficient
    while (H(par1) > value - c * gamma * h_prime) {
      gamma <- d * gamma
      par1 <- par - gamma * grad
    }

    par <- par1
  }

  if (n == maxit) {
    warning("Maximum number, ", maxit, ", of iterations reached")
  }

  par
}
```

We will use the Poisson regression example to illustrate the use of gradient
descent and other optimization algorithms, and we need to implement functions in
R for computing the negative log-likelihood and its gradient.

The implementation below uses a function factory to produce a list containing a
parameter vector, the negative log-likelihood function and its gradient. We
anticipate that for the Newton algorithm in Section \@ref(Newton) we also need
an implementation of the Hessian, which is thus included here as well.

We will exploit the `model.matrix()` function to construct the model matrix from
the data via a formula, and the sufficient statistic is also computed. The
implementations use linear algebra and vectorized computations relying on access
to the model matrix and the sufficient statistics in their enclosing
environment.

```{r}
#| label: Implement

poisson_model <- function(form, data, response) {
  X <- model.matrix(form, data)
  y <- data[[response]]

  # The computation of t(X) %*% y is more efficiently done by
  # crossprod(X,y). Use the function drop() to drop the dim attribute
  # and turn a matrix with one column into a vector
  t_map <- drop(crossprod(X, y))
  N <- nrow(X)
  p <- ncol(X)

  H <- function(beta) {
    drop(sum(exp(X %*% beta)) - beta %*% t_map) / N
  }

  grad <- function(beta) {
    (drop(crossprod(X, exp(X %*% beta))) - t_map) / N
  }

  Hessian <- function(beta) {
    crossprod(X, drop(exp(X %*% beta)) * X) / N
  }

  list(par = rep(0, p), H = H, grad = grad, Hessian = Hessian)
}
```

We choose to normalize the log-likelihood by the number of observations $N$ (the
number of rows in the model matrix). This does have a tiny computational cost,
but the resulting numerical values become less dependent upon $N$, which makes
it easier to choose sensible default values of various parameters for the
numerical optimization algorithms. Gradient descent is very slow for the large
Poisson model with individual store effects, so we consider only the simple
model with two parameters.

```{r}
#| label: veg-pois

veg_pois <- poisson_model(
  ~ log(normalSale),
  CSwR::vegetables,
  response = "sale"
)
```

```{r, dependson=c("veg-pois", "gd")}
#| label: gd-test

pois_gd <- grad_desc(veg_pois$par, veg_pois$H, veg_pois$grad)
```

The gradient descent implementation is tested by comparing the minimizer to the
estimated parameters as computed by `glm()`.

```{r, dependson=c("gd-test", "pois-model-null")}
#| label: gd-comp
#| results: "hold"

rbind(pois_glm = coefficients(pois_model_null), pois_gd)
```

We get the same result up to the first two decimals. The convergence criterion
on our gradient descent algorithm was quite loose ($\varepsilon = 10^{-4}$,
which means that the norm of the gradient is smaller than $10^{-2}$ when the
algorithm stops). This choice of $\varepsilon$ in combination with
$\gamma_0 = 0.01$ implies that the algorithm stops when the gradient is so small
that the changes are at most of norm $10^{-4}$.

Comparing the resulting values of the negative log-likelihood shows agreement up
to the first five decimals, but we notice that the negative log-likelihood for
the parameters fitted using `glm()` is just slightly smaller.

```{r, dependson=c("veg-pois", "pois-model-null", "gd-test"), echo=2:3}
#| label: gd-object
#| results: "hold"

old_options = options(digits = 15)
veg_pois$H(coefficients(pois_model_null))
veg_pois$H(pois_gd)
options(digits = old_options$digits)
```

To investigate what actually went on inside the gradient descent algorithm we
will use the callback argument to trace the internals of a call to
`grad_desc()`. The `tracer()` function from the [CSwR
package](https://github.com/nielsrhansen/CSwR/tree/master/CSwR_package) can be
used to construct a tracer object with a `tracer()` function that we can pass as
the callback argument. The tracer object and its `tracer()` function work by
storing information in the enclosing environment of `tracer()`. When used as the
callback argument to, e.g., `grad_desc()` the `tracer()` function will look up
variables in the evaluation environment of `grad_desc()`, store them and print
them if requested, and store runtime information as well.

When `grad_desc()` has returned, the trace information can be accessed via the
`summary()` method for the tracer object. The tracer objects and their
`tracer()` function should not be confused with the `trace()` function from the
R base package, but the tracer object's `tracer()` function can be passed as the
`tracer` argument to `trace()` to interactively inject tracing code into any R
function. Here, the tracer objects will only be used together with a callback
argument.

We use the tracer object with our gradient descent implementation and print
trace information every 50th iteration.

```{r, dependson=c("gd", "veg-pois")}
#| label: gd-trace

gd_tracer <- CSwR::tracer(c("value", "h_prime", "gamma"), Delta = 50)
pois_gd <- grad_desc(
  veg_pois$par,
  veg_pois$H,
  veg_pois$grad,
  cb = gd_tracer$tracer
)
```

We see that the gradient descent algorithm runs for a little more than 350
iterations, and we can observe how the value of the negative log-likelihood is
descending. We can also see that the step length $\gamma$ bounces between
$0.004096 = 0.8^4 \times 0.01$ and $0.00512 = 0.8^3 \times 0.01$, thus the
backtracking takes 3 to 4 iterations to find a step length with sufficient
descent.

The printed trace does not reveal the runtime information. The runtime is
measured for each iteration of the algorithm and the cumulative runtime is of
greater interest. This information can be computed and inspected after the
algorithm has converged, and it is returned by the summary method for tracer
objects.

```{r, dependson=c("gd-trace")}
#| label: this-summary

tail(summary(gd_tracer))
```

The trace information is stored in a list. The summary method transforms the
trace information into a data frame with one row per iteration. We can also
access individual entries of the list of trace information via subsetting.

```{r, dependson=c("gd-trace")}
#| label: gd-trace-sub

gd_tracer[377]
```

```{r, dependson=c("gd-trace", "veg-pois")}
#| label: gd-trace-plot
#| echo: false
#| fig-cap: "Gradient norm (left) and value of the negative log-likelihood (right) above the limit value $H(\\theta_{\\infty})$. The straight line is fitted to the data points except the first ten using least squares, and the rate is computed from this estimate and reported per ms."
#| out-width: "100%"
#| fig-width: 8
#| fig-height: 3
#| message: false

trace_gd <- summary(gd_tracer)

rate1 <- exp(coef(lm(
  log(h_prime) / 2 ~ I(1000 * .time),
  data = trace_gd,
  subset = 11:nrow(trace_gd)
)))[2]

p1 <- ggplot(data = trace_gd, aes(x = 1000 * .time, y = sqrt(h_prime))) +
  geom_point() +
  scale_y_log10("Gradient norm") +
  xlab("Time (ms)") +
  geom_smooth(
    method = "lm",
    se = FALSE,
    method.args = list(subset = 11:nrow(trace_gd))
  ) +
  geom_text(
    aes(x = 15, y = 10, label = paste("rate:", round(rate1, 3), "per ms")),
    data = data.frame()
  )

value_Inf <- veg_pois$H(coefficients(pois_model_null))
rate2 <- exp(coef(lm(
  log(value - value_Inf) ~ I(1000 * .time),
  data = trace_gd,
  subset = 11:nrow(trace_gd)
)))[2]

p2 <- ggplot(data = trace_gd, aes(x = 1000 * .time, y = value - value_Inf)) +
  geom_point() +
  scale_y_log10(expression(Value - Value[infinity])) +
  xlab("Time (ms)") +
  geom_smooth(
    method = "lm",
    se = FALSE,
    method.args = list(subset = 11:nrow(trace_gd))
  ) +
  geom_text(
    aes(x = 15, y = 10, label = paste("rate:", round(rate2, 3), "per ms")),
    data = data.frame()
  )

cc <- coord_cartesian(xlim = c(0, 20), ylim = c(1e-6, 1000))

library(patchwork)

(p1 + cc) + (p2 + cc)
```

```{r}
#| label: gd-trace-plot-diff
#| echo: false
#| eval: false

rate3 <- exp(coef(lm.fit(
  cbind(1, 11:nrow(trace_gd)),
  log(trace_gd$h_prime[11:nrow(trace_gd)])
))[2])

delta <- trace_gd$.time[nrow(trace_gd)] / nrow(trace_gd)

plot(diff(trace_gd$.time), ylim = c(0, 0.0003))
abline(h = delta, col = "red")
rate3^(1 / (1000 * delta))
```

## Newton's method {#Newton}

We can re-derive gradient descent from a different angle, and that derivation
will lead us straight to Newton's method. Take the second-order Taylor expansion
of $H$ around $\theta$,

\[
H(\theta + \rho) \approx H(\theta) + \nabla H(\theta)^\top \rho + \frac{1}{2} \rho^\top \nabla^2 H(\theta) \rho,
\]

and replace the Hessian $\nabla^2 H(\theta)$ with the conservative diagonal
$(1/t) I$. Minimizing the resulting quadratic over $\rho$ recovers
$\rho = -t \nabla H(\theta)$, the gradient descent direction. Substituting a
scalar for the Hessian is what makes gradient descent both cheap (no Hessian to
compute) and slow on ill-conditioned problems (it discards the geometry of the
level sets). Figure\ \@ref(fig:gd-taylor-surrogate) shows the surrogate model
for a one-dimensional objective, where the dashed quadratic surrogate uses
curvature $1/t$ in place of the true Hessian, and its minimizer is the gradient
descent step.

```{r}
#| label: gd-taylor-surrogate
#| echo: false
#| fig-width: 4
#| fig-height: 3
#| fig-cap: "Gradient descent as surrogate minimization. The dashed quadratic, obtained from the second-order Taylor expansion of $H$ around $\\theta$ with the Hessian replaced by $(1/t)I$, has minimizer $\\theta^+ = \\theta - t \\nabla H(\\theta)$ (open circle). The filled dots mark $\\theta$ and the next iterate $\\theta^+$ on $H$. Newton's method instead keeps the true Hessian."
#| warning: false

H <- function(theta) (theta - 2)^2 + 1
grad_H_1d <- function(theta) 2 * (theta - 2)

theta <- 0
t <- 0.2
theta_p <- theta - t * grad_H_1d(theta)

x_grid <- seq(theta - 0.5, theta + 1.7, length.out = 200)
surrogate <- function(x) {
  H(theta) + grad_H_1d(theta) * (x - theta) + (x - theta)^2 / (2 * t)
}

curve_df <- data.frame(
  x = x_grid,
  H_val = H(x_grid),
  surr_val = surrogate(x_grid)
)

iterate_pts <- data.frame(
  x = c(theta, theta_p),
  y = c(H(theta), H(theta_p)),
  label = c("theta", "theta^{`+`}")
)

surr_min <- data.frame(x = theta_p, y = surrogate(theta_p))

ggplot(curve_df, aes(x)) +
  geom_line(aes(y = H_val)) +
  geom_line(aes(y = surr_val), linetype = "dashed", col = "steelblue") +
  geom_segment(
    aes(
      x = theta_p,
      xend = theta_p,
      y = H(theta_p),
      yend = surrogate(theta_p)
    ),
    linetype = "dotted"
  ) +
  geom_point(data = iterate_pts, aes(x = x, y = y)) +
  geom_point(data = surr_min, aes(x = x, y = y), shape = 1) +
  geom_text(
    data = iterate_pts,
    aes(x = x, y = y, label = label),
    parse = TRUE,
    nudge_y = -0.35
  ) +
  labs(x = expression(theta), y = expression(H(theta))) +
  theme_schematic()
```

Newton's method does the opposite: it keeps the full Hessian and minimizes the
second-order Taylor model exactly. When $\nabla^2 H(\theta)$ is positive
definite, the minimizer of
$\rho \mapsto \nabla H(\theta)^\top \rho + \frac{1}{2} \rho^\top \nabla^2 H(\theta) \rho$
is

\[
\rho_{\mathrm{N}}(\theta) = -[\nabla^2 H(\theta)]^{-1} \nabla H(\theta),
\]

the *Newton direction*.\index{Newton direction}
Figure\ \@ref(fig:gd-vs-newton-surrogate) compares the two surrogates and their
trajectories on the ill-conditioned OLS design from Section\ \@ref(smoothness)
(Figure\ \@ref(fig:ols-condition-contour), right panel). The objective is
quadratic, so its Hessian is constant and the Newton surrogate equals $H$
exactly---its level sets coincide with the grey ellipses, and a single Newton
step lands at $\hat{\beta}$. The gradient descent surrogate replaces that
Hessian with $L \cdot I$ (where $L$ is the largest eigenvalue of
$X^\top X$)---its level sets are circles aligned with neither axis of the true
objective, and the resulting iterates zigzag across the narrow valley before
drifting along it towards $\hat{\beta}$. This is the condition-number penalty
paid by gradient descent in geometric form.

```{r}
#| label: gd-vs-newton-surrogate
#| echo: false
#| fig-width: 8
#| fig-height: 4.5
#| fig-cap: "Surrogate models and iterates for the OLS objective on the ill-conditioned design of Figure\\ \\@ref(fig:ols-condition-contour) (right panel). Both panels start from the filled dot $\\beta_0$. Solid grey contours are level sets of $H(\\beta) = \\tfrac{1}{2}\\lVert y - X\\beta \\rVert^2$; dashed blue contours are level sets of the surrogate built at $\\beta_0$, drawn at the same values. Left: gradient descent at step size $t = 1/L$. The surrogate is isotropic and misses the elongation of $H$, and the iterates trace a zigzag across the valley. Right: Newton's method. The surrogate coincides with $H$ (the dashed contours overlay the grey ones), and a single step reaches $\\hat{\\beta}$."

library(patchwork)

set.seed(24)
N_fig <- 100
z1 <- rnorm(N_fig)
z2 <- rnorm(N_fig)
X <- cbind(z1, 0.8 * z1 + sqrt(1 - 0.8^2) * z2)
beta_true <- c(1, 2)
y <- X %*% beta_true

XtX <- crossprod(X)
Xty <- crossprod(X, y)
beta_hat <- as.numeric(solve(XtX, Xty))
eig <- eigen(XtX, symmetric = TRUE)
L <- eig$values[1]
flat_dir <- eig$vectors[, 2]
steep_dir <- eig$vectors[, 1]
grad_f <- function(b) drop(XtX %*% b - Xty)

beta0 <- beta_hat + 1.8 * steep_dir + 0.6 * flat_dir
g0 <- grad_f(beta0)
f0 <- 0.5 * sum((beta0 - beta_hat) * (XtX %*% (beta0 - beta_hat)))
t_gd <- 1 / L

n_steps <- 30
gd_path <- matrix(0, n_steps + 1, 2)
gd_path[1, ] <- beta0
for (i in seq_len(n_steps)) {
  gd_path[i + 1, ] <- gd_path[i, ] - t_gd * grad_f(gd_path[i, ])
}
newton_path <- rbind(beta0, beta_hat)

halfwidth <- 2.5

grid <- expand.grid(
  b1 = seq(beta_hat[1] - halfwidth, beta_hat[1] + halfwidth, length.out = 200),
  b2 = seq(beta_hat[2] - halfwidth, beta_hat[2] + halfwidth, length.out = 200)
)

de1 <- grid$b1 - beta_hat[1]
de2 <- grid$b2 - beta_hat[2]
grid$H <- 0.5 *
  (XtX[1, 1] * de1^2 + 2 * XtX[1, 2] * de1 * de2 + XtX[2, 2] * de2^2)

d1 <- grid$b1 - beta0[1]
d2 <- grid$b2 - beta0[2]
grid$gd_surr <- f0 + g0[1] * d1 + g0[2] * d2 + (d1^2 + d2^2) / (2 * t_gd)
grid$newton_surr <- f0 +
  g0[1] * d1 +
  g0[2] * d2 +
  0.5 * (XtX[1, 1] * d1^2 + 2 * XtX[1, 2] * d1 * d2 + XtX[2, 2] * d2^2)

start_pt <- data.frame(b1 = beta0[1], b2 = beta0[2])
gd_df <- data.frame(b1 = gd_path[, 1], b2 = gd_path[, 2])
newton_df <- data.frame(b1 = newton_path[, 1], b2 = newton_path[, 2])
breaks <- c(10, 30, 80, 200, 400, 700)

panel <- function(z_col, path_df, title) {
  ggplot(grid, aes(b1, b2)) +
    geom_contour(aes(z = H), color = "grey70", breaks = breaks) +
    geom_contour(
      aes(z = .data[[z_col]]),
      color = "steelblue",
      linetype = "dashed",
      breaks = breaks
    ) +
    geom_path(data = path_df, linewidth = 0.4) +
    geom_point(data = path_df, size = 0.8) +
    geom_point(data = start_pt, size = 2) +
    # coord_equal(
    #   xlim = beta_hat[1] + c(-1, 1) * halfwidth,
    #   ylim = beta_hat[2] + c(-1, 1) * halfwidth
    # ) +
    labs(
      x = expression(beta[1]),
      y = expression(beta[2]),
      title = title
    )
}

panel("gd_surr", gd_df, "Gradient descent") +
  panel("newton_surr", newton_df, "Newton") +
  plot_layout(axes = "collect")
```

The Newton iteration is then
$\theta_{n+1} = \theta_n + \gamma_n \rho_{\mathrm{N}}(\theta_n)$. Pure Newton
takes $\gamma_n = 1$, which is the right choice near a minimizer with
positive-definite Hessian. Far from a minimizer, the model may be a poor guide
and the natural unit step can overshoot, so the practical version combines the
Newton direction with the backtracking line search of
Section\ \@ref(line-search). This is called *damped
Newton*.\index{Newton's method!damped}

The Newton direction is automatically a descent direction whenever
$\nabla^2 H(\theta_n)$ is positive definite:

\[
\rho_{\mathrm{N}}(\theta_n)^\top \nabla H(\theta_n) = -\nabla H(\theta_n)^\top [\nabla^2 H(\theta_n)]^{-1} \nabla H(\theta_n) < 0.
\]

When the Hessian is not positive definite, which can happen for non-convex $H$,
we must either modify the Hessian by adding $\lambda I$ for some $\lambda > 0$
or fall back to gradient descent for the offending iteration.

Newton's method has two structural advantages over plain gradient descent.
First, it is *affine invariant*: under a reparametrization $\theta = A\eta + b$
for invertible $A$, the Newton iterates in $\eta$ are exact images of those in
$\theta$, so the algorithm sees no difference between a well-conditioned and an
ill-conditioned parametrization of the same underlying problem. The
condition-number dependence that makes gradient descent slow
(Section\ \@ref(smoothness)) is no longer an issue. Second, near a strict local
minimizer where the Hessian is positive definite and Lipschitz continuous, pure
Newton converges *quadratically*: there is a constant $M$ such that

\[
\|\theta_{n+1} - \theta^\star\|_2 \leq M \|\theta_n - \theta^\star\|_2^2,
\]

so the number of correct digits roughly doubles per iteration. We state this
without proof, and refer the interested reader to @Boyd:2004; @Nocedal:2006. Far
from the minimizer, damped Newton enters a *damped phase* with $\gamma_n < 1$
and at least linear convergence. Once the iterates reach a neighborhood of
$\theta^\star$, they transition to the pure-Newton phase, where quadratic
convergence takes over.

The cost of these advantages is an increase in the computational demands of
every iteration. Each Newton step requires the Hessian $\nabla^2 H(\theta_n)$, a
dense $p \times p$ matrix that costs $O(N p^2)$ to assemble for typical
exponential-family models (Section\ \@ref(parametrization)), and the solution of
a linear system in that Hessian, costing $O(p^3)$ by Cholesky factorization. For
$p$ in the hundreds this is unproblematic; for $p$ in the millions it is
prohibitive. The quasi-Newton methods of the next section cut the cost by
replacing the Hessian inverse with a cheap surrogate.

We implement damped Newton below as the function `newton()`, an immediate
extension of `grad_desc()` from Section\ \@ref(stop-crit) with the Hessian added
to the argument list and the gradient swapped for the Newton direction.

```{r}
#| label: Newton

newton <- function(
  par,
  H,
  gr,
  hess,
  d = 0.8,
  c = 0.1,
  gamma0 = 1,
  epsilon = 1e-10,
  maxit = 50,
  cb = NULL
) {
  for (n in 1:maxit) {
    value <- H(par)
    grad <- gr(par)
    grad_norm_sq <- sum(grad^2)

    if (!is.null(cb)) {
      cb()
    }

    if (grad_norm_sq <= epsilon) {
      break
    }

    Hessian <- hess(par)
    rho <- -drop(solve(Hessian, grad))
    gamma <- gamma0
    par1 <- par + gamma * rho
    h_prime <- t(grad) %*% rho

    while (H(par1) > value + c * gamma * h_prime) {
      gamma <- d * gamma
      par1 <- par + gamma * rho
    }

    par <- par1
  }

  if (n == maxit) {
    warning("Maximum number, ", maxit, ", of iterations reached")
  }

  par
}
```

### Poisson regression

We reconsider the Poisson regression model `pois_model` from
Section\ \@ref(regression). The vegetables dataset has 352 stores, so a fit with
store-level intercepts plus a slope on $\log(\mathrm{normalSale})$ gives 353
parameters, a non-trivial size that exposes the differences between optimization
algorithms.

```{r, dependson=c("Implement")}
#| label: X-update

veg_pois_large <- poisson_model(
  ~ store + log(normalSale) - 1,
  CSwR::vegetables,
  response = "sale"
)
```

The `veg_pois_large` list contains the negative log-likelihood $H$, the gradient
$\nabla H$, and the Hessian $\nabla^2 H$ that the Newton iteration requires.

```{r, dependson=c("X-update", "Newton")}
#| label: Newton-test

newton_tracer <- CSwR::tracer(
  c("value", "gamma", "grad_norm_sq"),
  Delta = 0
)
pois_newton <- newton(
  veg_pois_large$par,
  veg_pois_large$H,
  veg_pois_large$grad,
  veg_pois_large$Hessian,
  cb = newton_tracer$tracer
)
```

We compare the resulting parameter estimates to those in `pois_model`, computed
using `glm.fit()`.

```{r, dependson=c("Newton-test", "pois-model")}
#| label: Newton-comp

range(pois_newton - coefficients(pois_model))
```

The differences are very small. The corresponding values of the negative
log-likelihood agree to high precision as well:

```{r, dependson=c("X-update", "Newton-test", "pois-model"), echo=2}
#| label: Newton-object

old_options = options(digits = 20)
rbind(
  pois_newton = veg_pois_large$H(pois_newton),
  pois_glm = veg_pois_large$H(coefficients(pois_model))
)
options(digits = old_options$digits)
```

The trace information from the Newton run shows the algorithm reaching the
stringent stopping criterion (squared gradient norm below $10^{-10}$) in only 10
iterations.

```{r, dependson=c("Newton-test")}
#| label: Newton-tracer-summary

summary(newton_tracer)
```

The whole run takes well under two seconds, and the per-iteration trace reveals
the two-phase pattern: the first iterations are damped ($\gamma_n < 1$) while
the iterates approach the basin of attraction; once inside, $\gamma_n = 1$ and
the squared gradient norm drops by many orders of magnitude per step. This is
the quadratic-convergence endgame in action.

R's built-in `glm.fit()` also uses a Newton iteration (without backtracking) and
is a natural benchmark.

```{r, dependson=c("X-update", "Newton-test")}
#| label: pois-glm.fit

env_pois <- environment(veg_pois_large$H)
bench::bench_time(
  glm.fit(env_pois$X, env_pois$y, family = poisson())
)
```

Using `glm.fit()` is about a factor five faster on this example than our Newton
algorithm, but we should always be very careful when comparing the runtimes of
two different optimization algorithms. In this case they have achieved about the
same numerical precision, and by the comparison above of the optimized value of
the objective function, the faster `glm.fit()` even obtained the smallest
negative log-likelihood of the two.

## Quasi-Newton methods

Newton's method's per-iteration cost, $O(N p^2)$ to assemble the Hessian plus
$O(p^3)$ for the linear solve, is fine for $p$ in the hundreds, as the Poisson
regression example just demonstrated, but punitive for larger problems.
*Quasi-Newton* methods keep the Newton update form

\[
\rho_n = -B_n \nabla H(\theta_n)
\]

but replace the inverse Hessian $[\nabla^2 H(\theta_n)]^{-1}$ with a cheaper
positive-definite surrogate $B_n$ updated incrementally from gradient
differences across iterations. Any positive-definite $B_n$ makes $\rho_n$ a
descent direction, since
$\rho_n^\top \nabla H(\theta_n) = -\nabla H(\theta_n)^\top B_n \nabla H(\theta_n) < 0$.

The dominant quasi-Newton algorithm in practice is *BFGS*\index{BFGS}
(Broyden--Fletcher--Goldfarb--Shanno), which maintains a rank-two update to
$B_n$ satisfying a *secant condition*: $B_n$ maps the most recent gradient
difference to the corresponding parameter difference, matching the true Hessian
inverse along that direction. The update costs $O(p^2)$ per iteration without
ever forming the Hessian. *L-BFGS*\index{L-BFGS} stores only the last few
gradient and parameter differences, typically between five and twenty, and
reconstructs $B_n$ on the fly. Its memory footprint is linear in $p$, and it is
the de facto standard for problems with $p$ in the thousands or millions. We
will not derive the BFGS update; @Nocedal:2006 has the details. R's `optim()`
provides BFGS as one of its methods:

```{r, dependson=c("X-update")}
#| label: BFGS

bench::bench_time(
  pois_bfgs <- optim(
    veg_pois_large$par,
    veg_pois_large$H,
    veg_pois_large$grad,
    method = "BFGS",
    control = list(maxit = 10000)
  )
)
```

```{r, dependson=c("BFGS", "pois-model")}
#| label: BFGS-results

range(pois_bfgs$par - coefficients(pois_model))
pois_bfgs[c("value", "counts")]
```

BFGS reaches an estimate matching `pois_model` to high precision in seconds,
without ever computing $\nabla^2 H$ explicitly. This is the default
recommendation for general-purpose maximum-likelihood estimation in R when the
gradient is available; both `optim()` and `nlminb()` use a quasi-Newton method
underneath.

### Nonlinear conjugate gradient

A historical predecessor to BFGS, the *nonlinear conjugate
gradient*\index{conjugate gradient} algorithm uses only gradient information but
couples each step to the previous descent direction so as to escape the zig-zag
pattern that slows plain gradient descent on ill-conditioned problems. In the
Fletcher-- Reeves version the descent direction is

\[
\rho_n = -\nabla H(\theta_n) + \frac{\|\nabla H(\theta_n)\|_2^2}{\|\nabla H(\theta_{n-1})\|_2^2} \rho_{n-1},
\]

initialized at $\rho_0 = -\nabla H(\theta_0)$; the Polak--Ribière and
Hestenes--Stiefel variants differ in the prefactor of $\rho_{n-1}$. Conjugate
gradient uses only $O(p)$ memory and avoids matrix storage altogether, but needs
many more iterations than BFGS to reach a given accuracy and is rarely the
default choice today. R's `optim()` exposes the algorithm via `method = "CG"`:

```{r, dependson=c("X-update")}
#| label: cg-test

bench::bench_time(
  pois_cg <- optim(
    veg_pois_large$par,
    veg_pois_large$H,
    veg_pois_large$grad,
    method = "CG",
    control = list(maxit = 10000)
  )$par
)
```

```{r, dependson="cg-test"}
#| label: cg-objective

veg_pois_large$H(pois_cg)
```

The CG run reaches the same minimizer as BFGS but spends substantially more
wall-clock time on this problem. This is one reason BFGS, or L-BFGS for larger
problems, is the better first choice.

## Constrained optimization {#pep-moth-descent}

Every algorithm in this chapter so far has assumed the parameter space is all of
$\mathbb{R}^p$, but many statistical models are not so accommodating. The
peppered-moth maximum-likelihood problem of Section\ \@ref(pep-moth) is
\index{peppered moth!barrier method}\index{barrier method} inherently
constrained: the allele frequencies $(p_C, p_I)$ must lie in the triangle
$\{p_C \geq 0,\, p_I \geq 0,\, p_C + p_I \leq 1\}$. The implementation in
Section\ \@ref(pep-moth) handled this by returning $+\infty$ at infeasible
parameters, and Nelder--Mead's derivative-free search handles that gracefully,
but the descent methods of Sections\ \@ref(grad-descent) and do not. A gradient
step does not respect inequality constraints; a backtracking line search treats
an infinite boundary as an arbitrarily steep wall but cannot see around it; and
the gradient itself is undefined at $+\infty$. The right algorithm for the
constrained problem is instead the *barrier method*, which is what we devote the
sequel of this section to.

Write the constrained problem in the standard form used by
Theorem\ \@ref(thm:kkt):

\[
\operatorname*{minimize}_\theta H(\theta) \quad \text{subject to} \quad
g_i(\theta) \leq 0, \quad i = 1, \ldots, r,
\]

with $H$ and the $g_i$ convex and differentiable. For peppered moths
$\theta = (p_C, p_I)$, $H(\theta) = -\ell(\theta)$, and the three inequalities
are

\[
g_1(\theta) = -p_C, \quad g_2(\theta) = -p_I, \quad g_3(\theta) = p_C + p_I - 1.
\]

The KKT conditions characterize the minimizer through stationarity, primal
feasibility, and complementary slackness, but they presume one already *knows*
which constraints are active at $\theta^\star$. A numerical algorithm cannot
start from that knowledge; it has to discover the active set as it goes.

The barrier method softens the hard inequalities into a smooth penalty that
diverges at the boundary. The standard choice is the *logarithmic barrier*

\begin{equation}
\phi(\theta) = -\sum_{i=1}^r \log(-g_i(\theta)),
(\#eq:log-barrier)
\end{equation}

defined on the strict interior
$\{\theta : g_i(\theta) < 0,\ i = 1, \ldots, r\}$. For peppered moths,
$\phi(\theta) = -\log p_C - \log p_I - \log(1 - p_C - p_I)$. Each term diverges
as the corresponding constraint becomes tight, so $\phi$ acts as a one-sided
wall pushing iterates away from the boundary. For each $t > 0$ consider the
unconstrained *centering problem*

\begin{equation}
\operatorname*{minimize}_\theta \, t H(\theta) + \phi(\theta).
(\#eq:centering)
\end{equation}

When $H$ is convex the centering objective is strictly convex on the interior,
and its minimizer $\theta^\star(t)$, when it exists, lies strictly inside the
feasible set. As $t$ grows the barrier matters less, iterates are allowed closer
to the boundary, and $\theta^\star(t)$ approaches the constrained optimum
$\theta^\star$. The map $t \mapsto \theta^\star(t)$ is the *central
path*\index{central path}, and the gap between objective values along it admits
a clean bound.

::: {.theorem #barrier-suboptimality name="Barrier-method suboptimality"}
Let $H$ and the $g_i$ be convex and differentiable, suppose the feasible set has
nonempty interior, and suppose the constrained problem has a minimizer
$\theta^\star$. For each $t > 0$ the centering problem \@ref(eq:centering) has a
minimizer $\theta^\star(t)$, and

\[
H(\theta^\star(t)) - H(\theta^\star) \leq r / t.
\]
:::

The proof, which we omith here, constructs an explicit dual-feasible point from
$\theta^\star(t)$ via the KKT conditions for the centering problem and
constructs the duality gap as $r/t$ (see @Boyd:2004, Section\ 11.2.2). The bound
is essentially tight, so $r/t$ is the right tolerance to target: each
multiplicative bump in $t$ shaves the suboptimality by the same factor, and
reaching gap $\epsilon$ requires only $t \geq r / \epsilon$.

The algorithm follows from the bound. Start from a strictly feasible
$\theta^{(0)}$ and a modest $t_0 > 0$; minimize the centering problem to find
$\theta^{(1)} = \theta^\star(t_0)$; multiply $t$ by some $\mu > 1$; re-minimize
starting from $\theta^{(1)}$; iterate until $r / t < \epsilon$. Each inner
subproblem is unconstrained and (for convex $H$) convex, so the descent
algorithms of Sections\ \@ref(grad-descent)--\@ref(Newton) handle it. Warm
starting from the previous central-path iterate keeps the inner solves cheap
because consecutive points on the central path are close. Typical choices are
$\mu \in [10, 100]$ with BFGS or damped Newton for the inner solve; @Boyd:2004
(Section\ 11.3) discusses the tradeoff between aggressive $\mu$ (few outer
iterations, each expensive) and gentle $\mu$ (many outer iterations, each
cheap).

Figure\ \@ref(fig:pep-central-path) plots the centering objective for the
peppered-moth problem at four values of $t$. When $t$ is small the barrier
dominates and the minimizer sits near the analytic center of the simplex; as $t$
grows the iterate slides along the central path toward the maximum-likelihood
estimate. Here the constrained optimum lies in the *interior* of the
simplex---no constraint is active at $\theta^\star$---but the MLE is close
enough to the $p_C = 0$ and $p_I = 0$ edges that the barrier visibly bends the
contours away from them.

```{r}
#| label: pep-central-path
#| echo: false
#| warning: false
#| fig-cap: "Contours of the centering objective $t H(\\theta) + \\phi(\\theta)$ for the peppered-moth problem at four values of $t$ (contour levels normalized within each panel). The dashed lines mark the simplex boundary, the white circle marks the constrained optimum $\\theta^\\star$, and the orange cross marks the minimizer of the centering problem at that $t$. As $t$ grows, the cross traces the central path from the analytic center of the simplex toward $\\theta^\\star$."
#| fig-width: 6
#| fig-height: 6.2

ny_fig <- c(85, 196, 341)
H_fig <- function(p) {
  pT <- 1 - p[1] - p[2]
  pD <- p[1]^2 + 2 * p[1] * p[2] + 2 * p[1] * pT
  pM <- p[2]^2 + 2 * p[2] * pT
  pL <- pT^2
  -(ny_fig[1] * log(pD) + ny_fig[2] * log(pM) + ny_fig[3] * log(pL))
}
phi_fig <- function(p) -log(p[1]) - log(p[2]) - log(1 - p[1] - p[2])

p1_grid <- seq(0.005, 0.995, length.out = 161)
p2_grid <- seq(0.005, 0.995, length.out = 161)
grid_df <- expand.grid(p1 = p1_grid, p2 = p2_grid)
grid_df$pT <- 1 - grid_df$p1 - grid_df$p2
grid_df <- subset(grid_df, pT > 0.001)
grid_df$H <- -(ny_fig[1] *
  log(
    grid_df$p1^2 + 2 * grid_df$p1 * grid_df$p2 + 2 * grid_df$p1 * grid_df$pT
  ) +
  ny_fig[2] * log(grid_df$p2^2 + 2 * grid_df$p2 * grid_df$pT) +
  ny_fig[3] * log(grid_df$pT^2))
grid_df$phi <- -log(grid_df$p1) - log(grid_df$p2) - log(grid_df$pT)

ts <- c(1e-4, 1e-3, 1e-2, 1)
t_label <- function(t) paste0("t == ", formatC(t, format = "g", digits = 2))

contour_df <- do.call(
  rbind,
  lapply(ts, function(t) {
    z <- t * grid_df$H + grid_df$phi
    data.frame(
      p1 = grid_df$p1,
      p2 = grid_df$p2,
      z = (z - min(z)) / (max(z) - min(z)),
      t = t
    )
  })
)
centering_df <- do.call(
  rbind,
  lapply(ts, function(t) {
    star <- optim(
      c(0.3, 0.3),
      function(p) t * H_fig(p) + phi_fig(p),
      method = "BFGS"
    )$par
    data.frame(p1 = star[1], p2 = star[2], t = t)
  })
)
opt <- optim(c(0.3, 0.3), H_fig, method = "BFGS")$par
opt_df <- data.frame(p1 = opt[1], p2 = opt[2], t = ts)

contour_df$panel <- factor(t_label(contour_df$t), levels = t_label(ts))
centering_df$panel <- factor(t_label(centering_df$t), levels = t_label(ts))
opt_df$panel <- factor(t_label(opt_df$t), levels = t_label(ts))

simplex_df <- data.frame(
  x = c(0, 1, 0),
  y = c(0, 0, 1),
  xend = c(1, 0, 0),
  yend = c(0, 1, 0)
)

ggplot() +
  geom_contour(
    data = contour_df,
    aes(p1, p2, z = z, color = after_stat(level)),
    breaks = seq(0, 0.4, length.out = 22),
    linewidth = 0.3
  ) +
  geom_segment(
    data = simplex_df,
    aes(x = x, y = y, xend = xend, yend = yend),
    linetype = "dashed",
    color = "grey40"
  ) +
  geom_point(
    data = opt_df,
    aes(p1, p2),
    shape = 21,
    fill = "white",
    color = "black",
    size = 2.4,
    stroke = 0.6
  ) +
  geom_point(
    data = centering_df,
    aes(p1, p2),
    shape = 4,
    color = "darkorange",
    size = 2.8,
    stroke = 1.3
  ) +
  facet_wrap(~panel, labeller = label_parsed) +
  scale_color_viridis_c(guide = "none") +
  coord_fixed(xlim = c(0, 1), ylim = c(0, 1), expand = FALSE) +
  labs(x = expression(p[C]), y = expression(p[I])) +
  theme_minimal()
```

For the peppered moths we re-implement the negative log-likelihood as a function
of the allele frequencies, dropping the boundary check from
Section\ \@ref(pep-moth) because the barrier will keep iterates in the interior:

```{r}
#| label: pep-neg-loglik

neg_loglik_pep <- function(p, ny) {
  p_T <- 1 - p[1] - p[2]
  p_dark <- p[1]^2 + 2 * p[1] * p[2] + 2 * p[1] * p_T
  p_mott <- p[2]^2 + 2 * p[2] * p_T
  p_light <- p_T^2
  -(ny[1] * log(p_dark) + ny[2] * log(p_mott) + ny[3] * log(p_light))
}
```

The log-barrier is similarly direct:

```{r}
#| label: pep-barrier

phi_pep <- function(p) {
  -log(p[1]) - log(p[2]) - log(1 - p[1] - p[2])
}
```

A straightforward implementation of the outer loop uses BFGS for each inner
solve via `optim()`:

```{r, dependson=c("pep-neg-loglik", "pep-barrier")}
#| label: pep-barrier-loop
#| warning: false

ny <- c(85, 196, 341)
r <- 3
mu <- 10
t <- 1
theta <- c(0.3, 0.3)

repeat {
  centering <- function(p) t * neg_loglik_pep(p, ny) + phi_pep(p)
  theta <- optim(theta, centering, method = "BFGS")$par

  if (r / t < 1e-8) {
    break
  }

  t <- mu * t
}
theta
```

R's `constrOptim()` packages this loop with an adaptive $\mu$ and its own
stopping rule. It expects affine inequality constraints in the form
$A\theta \succeq b$, supplied as a matrix `ui` and a vector `ci` with
`ui %*% theta >= ci`. Rewriting $g_1, g_2, g_3 \leq 0$ in that form gives

\[
A = \begin{pmatrix} 1 & 0 \\ 0 & 1 \\ -1 & -1 \end{pmatrix},
\quad b = \begin{pmatrix} \phantom{-}0 \\ \phantom{-}0 \\ -1 \end{pmatrix},
\]

so the same problem reduces to one call:

```{r, dependson="pep-neg-loglik"}
#| label: pep-constropt
#| warning: false

A <- rbind(c(1, 0), c(0, 1), c(-1, -1))
b <- c(0, 0, -1)

constrOptim(
  c(0.3, 0.3),
  neg_loglik_pep,
  NULL,
  ui = A,
  ci = b,
  ny = c(85, 196, 341)
)$par
```

Both runs return the maximum-likelihood estimate found by Nelder--Mead in
Section\ \@ref(pep-moth). The algorithms differ; the optimum does not.

This example also illustrates the second strategy outlined in
Section\ \@ref(constraints): keep the constraints and solve a sequence of smooth
unconstrained problems. The first strategy, reparametrizing through the softmax
to lift the simplex to all of $\mathbb{R}^K$, would also work here but it
destroys convexity of the multinomial log-likelihood and forfeits the KKT-based
suboptimality bound. A third approach: The EM algorithm (covered in
Chapter\ \@ref(em)), exploits the latent-genotype structure specific to the
peppered moths to sidestep the constraint problem entirely.

## Algorithms in practice

The previous sections introduced five algorithms: gradient descent, Newton's
method, BFGS, nonlinear conjugate gradient, and the barrier method. We provided
implementations in R that we one could inspect and modify. In an actual data
analysis, R features a set of classical, well-tested algorithms through
`optim()`, `constrOptim()`, and a handful other specialized interfaces, and
these functions serve most basic use cases. This short section tries to link the
algorithms to those interfaces, raises two common pitfalls, and revisits the
Poisson-regression example to show what knowing the structure of the problem can
sometimes gain us in terms of numerical performance.

`optim()` is R's workhorse for general-purpose unconstrained minimization. Its
`method` argument exposes Nelder--Mead (the default, derivative-free), `"BFGS"`,
`"L-BFGS-B"` (box-constrained quasi-Newton), `"CG"` (Polak--Ribière conjugate
gradient), and `"SANN"` (simulated annealing, not covered here). If you are able
to supply a function to compute the gradient through `gr`, then you should
typically opt for L-BFGS-B, which is a good off-the-shelf choice that also
allows box constraints to be supplied through `lower` and `upper`. `nlminb()`
and `nlm()` are alternatives that calls quasi-Newton routines and are sometimes
faster on expensive objectives. For linear inequality constraints in the form
$A\teta \succeq b$, we can use `constrOptim()`, which wraps `optim()` with the
barrier method of Section\ \@ref(pep-moth-descent). If you are interested in an
extensive guide to the optimization infrastructure R and other CRAN packages
have to offer, we recommend @Nash:2014 and the [CRAN task view of optimization
and mathematical programming](https://cran.r-project.org/view=Optimization).

For canonical-link generalized linear models, `glm()` and `glm.fit()` implement
iteratively reweighted least squares, which is exactly Newton's method on the
negative log-likelihood specialized to the exponential family
(Section\ \@ref(parametrization)). Bypassing `glm()` to call `optim()` directly
is rarely worth it: the specialized implementation is both faster and more
convenient to work with. The Poisson-regression code from
Sections\ \@ref(grad-descent) and \@ref(Newton) was written by hand precisely to
expose the descent algorithms. In practice, fitting the model comes down to a
single line via `glm(..., family = "poisson")`.

Here, we want to raise two common pitfalls. First, numerical gradients via
finite differences are subject to a tradeoff between truncation and roundoff
error. They work well in well-behaved regions of the optimization surface,but
can stall a quasi-Newton method near an optimum. Supplying an analytic gradient
is almost always worth the effort, both for speed and stability. Second, the
default convergence tolerances assume $H \sim O(1)$---if the objective is far
away from unity, the `parscale` and `fnscale` entries of `optim()`'s `control`
argument may help by rescaling the parameters and the objective to a more
typical range.

### Sparsity

The custom implementations of $H$ and its derivatives, like the descent
algorithms that consume them, treat $\mathbf{X}$ as an abstract matrix and
exploit sparsity essentially for free. The Poisson-regression implementations
above used a dense $\mathbf{X}$ only because that is what `model.matrix()`
faster. The Matrix package supplies the sparse class:

```{r}
#| label: Matrix
#| cache: false

library(Matrix)
```

```{r, dependson=c("pois-glm.fit", "X-update")}
#| label: sparse

env_pois$X <- Matrix(env_pois$X)
class(env_pois$X)
```

Without any other change, BFGS via `optim()` runs faster:

```{r, dependson=c("X-update", "sparse")}
#| label: BFGS-sparse

bench::bench_time(
  pois_bfgs_sparse <- optim(
    veg_pois_large$par,
    veg_pois_large$H,
    veg_pois_large$grad,
    method = "BFGS",
    control = list(maxit = 10000)
  )
)
```

For real applications one should avoid constructing the dense matrix in the
first place. `MatrixModels::model.Matrix()` builds a sparse model matrix
directly from a formula:

```{r, dependson=c("pois-glm.fit"), echo=-1}
#| label: sparse-model-matrix

library(MatrixModels)
env_pois$X <- MatrixModels::model.Matrix(
  ~ store + log(normalSale) - 1,
  data = CSwR::vegetables,
  sparse = TRUE
)
class(env_pois$X)
```

```{r}
#| label: time-model.matrix
#| echo: false
#| eval: false

library(MatrixModels)
bench::mark(
  sparse = X_sparse <- model.Matrix(
    ~ store + log(normalSale) - 1,
    data = CSwR::vegetables,
    sparse = TRUE
  ),
  dense = X <- model.matrix(
    ~ store + log(normalSale) - 1,
    data = CSwR::vegetables
  ),
  check = FALSE
) %>%
  autoplot()

range(as.matrix(X_sparse) - X)
```

Newton's method benefits the most. The bottleneck of the dense implementation is
the per-iteration Hessian assembly; with a sparse $\mathbf{X}$, that assembly
drops by orders of magnitude.

```{r, echo=2:4, dependson=c("X-update", "Newton", "sparse-model-matrix")}
#| label: time-sparse-model-matrix

pois_newton <- newton(
  # warm up to avoid lazy-loading overhead in the timing
  veg_pois_large$par,
  veg_pois_large$H,
  veg_pois_large$grad,
  veg_pois_large$Hessian
)

newton_tracer <- CSwR::tracer(c("value", "h_prime", "gamma"), Delta = 0)

pois_newton <- newton(
  veg_pois_large$par,
  veg_pois_large$H,
  veg_pois_large$grad,
  veg_pois_large$Hessian,
  cb = newton_tracer$tracer
)

summary(newton_tracer)
```

Pulling the timings together: with a dense $\mathbf{X}$, the conjugate gradient
algorithm took around 10 seconds on the 353-parameter Poisson problem, BFGS
roughly 2--3 seconds, and Newton's method between 0.3 and 1.7 seconds. Switching
to a sparse $\mathbf{X}$ cut BFGS's runtime by a factor of three and Newton's by
a factor of roughly 150, to about 0.009 seconds. The construction of the sparse
model matrix is no slower than the dense one. Memory tells a similar story:

```{r, dependson=c("pois-glm.fit", "sparse-model-matrix"), echo=c(2, 4)}
#| label: X-size
#| results: "hold"

cat("Sparse matrix memory usage:\n")
object.size(env_pois$X)

cat("Dense matrix memory usage:\n")
object.size(as.matrix(env_pois$X))
```

The dense representation uses about thirty times more memory. At the
353-parameter scale of this example, the absolute difference between 3\ MB and
100\ kB is unimportant. At the scale where it *would* matter, say going from
3\ GB to 100\ MB, the choice between dense and sparse can be the difference
between fitting the model on a laptop and not being able to load the design at
all.

The `glm4()` function in MatrixModels accepts sparse model matrices directly and
is the recommended choice when the design is large and sparse. Two common
situations produce sparsity: high-cardinality factor predictors, especially
their interactions, and basis expansions of quantitative predictors using
locally supported bases such as B-splines. In either case, switching to a sparse
model matrix is essentially free at the modeling level and pays substantial
dividends in compute and memory for the algorithms developed in this chapter.
dividends in compute and memory for the algorithms developed in this chapter.
