# R programming {#app:R}

This appendix on R programming is meant to be a quick review of the most important
programming concepts in R that are used in the book. It should be read together 
with or after the [Introduction](#intro). For a detailed and much more
extensive coverage of R as a programming language the reader is referred to 
the book [Advanced R](https://adv-r.hadley.nz/).

## Functions

### Vectorization

## Objects and methods

## Environments

### Function factories

## Performance

### Parallel computations


## Exercises

### R training exercises {-}

```{exercise, numerics}
Explain the result of evaluating the following R expression. 
```

```{r ex-numerics}
(0.1 + 0.1 + 0.1) > 0.3

```


```{exercise, filter-function}
Write a function that takes a numeric vector `x` and a threshold value `h` 
as arguments and returns the vector of all values in `x` greater than `h`.
Test the function on `seq(0, 1, 0.1)` with threshold 0.3. Have the example 
from Exercise \@ref(exr:numerics) in mind.

```
  
  
```{exercise, special-values}
Investigate how your function from Exercise \@ref(exr:filter-function) 
treats missing values (`NA`), infinite values
(`Inf` and `-Inf`) and the special value "Not a Number" (`NaN`). Rewrite your
function (if necessary) to exclude all or some of such values from `x`.

*Hint: The functions `is.na`, `is.nan` and `is.finite` are useful.* 

```


### Histograms with non-equidistant breaks {-}

The following three exercises will use a data set consisting of measurements from objects 
outside of our galaxy including infrared emissions. We will focus on the variable 
F12, which is the total 12 micron band flux density.


```{r infrared-read}
infrared <- read.table("data/infrared.txt", header = TRUE)
F12 <- infrared$F12
```


The purpose of this exercise is two-fold. First, you will get familiar with the 
data and see how different choices of visualizations using histograms can affect 
your interpretation of the data. Second, you will learn more about how to write 
functions in R and gain a better understanding of how they work.


```{exercise, plotHist}
Plot a histogram of `log(F12)` using the default value of the argument `breaks`. Experiment with alternative values of `breaks`. 

```


```{exercise, myBreaks}
Write your own function, called `myBreaks`, which takes two arguments, `x` (a vector) and `h` (a positive integer). Let `h` have default value `5`. The function should first sort
`x` into increasing order and then return the vector that: starts with the smallest entry in `x`;
contains every $h$th unique entry from the sorted `x`; ends with the largest entry in `x`.  

For example, if `h = 2` and `x = c(1, 3, 2, 5, 10, 11, 1, 1, 3)` the function should return `c(1, 3, 10, 11)`. To see this, first sort `x`, which gives the vector `c(1, 1, 1, 2, 3, 3, 5, 10, 11)`, whose unique 
values are `c(1, 2, 3, 5, 10, 11)`. Every second unique entry is `c(1, 3, 10)`, and then the largest entry `11` is concatenated.

*Hint: The functions `sort` and `unique` can be useful.* 
  
Use your function to construct *breakpoints* for the histogram for different values of `h`, and compare with the histograms obtained in Exercise \@ref(exr:plotHist).  

```


```{exercise, myBreaks-advanced}
If there are no ties in the data set, the function above will produce breakpoints 
with `h` observations in the interval between two consecutive breakpoints 
(except the last two perhaps). If there are ties, the function will by construction 
return unique breakpoints, but there may be 
more than `h` observations in some intervals. 

*The intention is now to rewrite `myBreaks` so that if possible each interval 
contains `h` observations.*

Modify the `myBreaks` function with this intention and so that is has the 
following properties:

* All breakpoints must be unique.
* The range of the breakpoints must cover the range of `x`.
* For two subsequent breakpoints, $a$ and $b$, there must be at least `h` observations in the interval $(a,b]$, provided `h < length(x)`. (With the exception that for the first two breakpoints, the interval is $[a,b]$.)

```

### Functions and functional programming {-}

The following four exercises build on having implemented a function that 
computes breakpoints for a histogram either as in 
Exercise \@ref(exr:myBreaks) or as in Exercise \@ref(exr:myBreaks-advanced).
The purpose of the exercises is to explore various features of the R language
related to functions and using R as a functional programming language. 

```{exercise, myHist}
Write a function called `myHist`, which takes a single argument `h` and plots a 
histogram of `log(F12)` using your function. Extend 
the implementation so that any additional argument specified when calling `myHist` 
is passed on to `hist`. Investigate and explain what happens when executing 
the following function calls.
```

```{r, eval = FALSE}
myHist()
myHist(h = 5, freq = TRUE)
myHist(h = 0)
```

Finally, what happens if you remove that data from the global environment and 
call `myHist` subsequently? 

```{exercise, myHist-env}
What is the environment of `myHist`? Change it to a new environment, and assign 
(using the function `assign`) the data to a 
variable with an appropriate name in that environment. Once this is done, 
check what now happens when calling `myHist` after
the data is removed from the global environment. 

```

```{exercise, myHist-funfact}
Write a function that takes an argument `x` (the data) and 
returns a function, where the returned function 
takes an argument `h` (just as `myHist`) and plots a histogram (just as `myHist`). 
Because the return value is a function, we may refer to the function 
as a [function factory](https://adv-r.hadley.nz/function-factories.html). 

What is the environment of the function created by the function factory? What 
is in the environment? 
Does it have any effect when calling the function whether the data is altered 
or removed from the global environment? 

```

```{exercise, }
Evaluate the following function call:

```

```{r, eval = FALSE}
tmp <- myHist(10, plot = FALSE)
```

What is the type and class of `tmp`? What happens when `plot(tmp, col = "red")` 
is executed? How can you find help on what `plot` does with an 
object of this class? Specifically, how do you find the documentation for the 
argument `col`, which is not an argument of `plot`?