# R programming {#app-R}

This appendix on R programming is a brief overview of important
programming constructs and concepts in R that are used in the book. 
For a detailed and much more
extensive coverage of R as a programming language the reader is referred to 
the book [Advanced R](https://adv-rapp-R.hadley.nz/).

Depending on your background, the appendix can serve different purposes.
If you already have experience with programming in R, this appendix can serve 
as a brush up on some basic aspects of the R language that are used throughout 
the book. If you have experience with programming in other languages than R, it
can serve as an introduction to typical R programming techniques,
where some may differ from what you know from other languages. If you have little 
prior experience with programming, this appendix can teach you the most important
things for getting started with the book, but you are encouraged to follow up 
with additional and more detailed material like [Advanced R](https://adv-r.hadley.nz/).

For everybody, this appendix covers specific topics relevant for reading the book. 
These topics include

* data types, comparisons and numerical precision
* vectorized computations
* functions, environments and function factories
* performance assessment and improvement
* S3 objects and methods

Several important topics, such as 
[S4](https://adv-r.hadley.nz/s4.html) and 
[R6](https://adv-r.hadley.nz/r6.html) objects, 
[expressions](https://adv-r.hadley.nz/expressions.html), and 
[data wrangling](https://r4ds.had.co.nz/wrangle-intro.html), are not covered in 
any detail. The book is on implementations of correct and efficient 
numerical algorithms used in statistics, and this is reflected in the 
topics covered in this appendix.

## Data structures {#app-data-structures}

The fundamental data structure in R is a vector. Even variables that look and 
behave like single numbers are vectors of length one. Vectors come in two flavors:
*atomic vectors* and *lists*.

An atomic vector is an indexed collection of data elements that are all 
of the same type, e.g.

* integers
* floating point numbers
* logical values
* character strings

A list is an indexed collection of elements without any type restrictions on the 
individual elements. An element in a list can, for instance, be a list itself.

### Atomic vectors

You can construct a vector in R by simply typing in its elements, e.g. 

```{r first_vector}
first_vector <- c(1, 2, 3)  # Note the 'c'
first_vector
```

The constructed vector contains the numbers 1, 2 and 3. We use the classical 
assignment operator `<-` throughout, while R supports using `=` if you prefer. 
The `c()` used on the right hand side of the assignment is short for *combine* 
(or concatenate), and it is also used if you combine two vectors into one.

```{r second_vector, dependson="first_vector"}
second_vector <- c(4, 5, 6)
c(first_vector, second_vector)
```

There are several convenient techniques in R for constructing vectors of various
regular nature, e.g. sequences. The following example shows how to construct a 
vector containing the integers from 1 to 10. The type of the vector is `integer`
indicating that the elements of the vector are stored as integers. 

```{r integer_vector}
integer_vector <- 1:10
integer_vector
typeof(integer_vector)  # `typeof` reveals the internal storage mode
```

We can access the individual elements as well as subsets of a vector 
by indices using square brackets.

```{r subscripting}
integer_vector[3]
integer_vector[first_vector]  # The first three elements
```

The function `seq` generalizes the colon operator, `1:10`, as a way to 
generate regular sequences. The following example shows how to generate 
a sequence, `double_vector`, from 0.1 to 1.0 with increments of size 0.1. 
The type of the resulting vector is `double`, which indicates that the elements 
of `double_vector` are stored as doubles. That is, the numbers are stored as 
floating point numbers using 64 bits of storage with a precision of 
just about 16 digits.

```{r double_vector}
double_vector <- seq(0.1, 1, by = 0.1)
double_vector
typeof(double_vector)
```

Integers are often stored as or coerced into doubles automatically. The supposedly 
integer vector, `first_vector`, is actually of type `double`. 

```{r num_int_example}
typeof(first_vector)
```

In R, numerical data of either type integer or double is collectively referred to 
as *numerics* and have mode (and class) `numeric`. This is confusing, in particular
because "numeric" is used also as pseudonym of the *type* double, but it is
rarely a practical problem. A vector of type integer or double is often just
said to by a `numeric`, and the function `is.numeric()` reflects this.

It is possible to insist that integers are actually stored as integers by 
appending `L` to each integer, e.g.

```{r num_int_example_int}
typeof(c(1L, 2L, 3L))
```

When sequences are generated using the colon operator with the endpoints being 
integers, as in `1:10`, the result will be a vector of type integer. This is 
an exception. Apparent integers are usually -- and silently -- converted into 
doubles if they are not explicitly marked as integers by `L`.

Vectors of any length can be created by the generic function `vector()`, or 
by type specific functions such as `numeric()` that creates vectors of type 
double. 

```{r}
vector("numeric", length = 10) 
numeric(10)
```

Both vectors above are of type double and of length 10 and initialized with all 
elements being 0. 

A *logical vector* is another example of a useful atomic vector. The default
type of a vector created by `vector()` is logical, with all elements being `FALSE`.

```{r}
vector(length = 10)
```

Logical vectors are encountered when we compare the elements of one vector
to another vector or to a number.

```{r logical_vector, dependson = "integer_vector"}
logical_vector <- integer_vector > 4
logical_vector
typeof(logical_vector)
```

While a logical vector has its own type and is stored efficiently as such, 
it behaves in many ways as a numeric vector with `FALSE` equivalent 
to 0 and `TRUE` equivalent to 1. If we want to compute the relative 
frequency of elements in `integer_vector` that are are (strictly) larger than 4, 
say, we can simply take the mean of `logical_vector`. 

```{r logcial_vector_mean, dependson = "logical_vector"}
mean(logical_vector)
```

The mean of a logical vector behaves as if the logical values are coerced into
zeros and ones before the mean is computed. 

A final example of an atomic vector is a *character vector*. In this example, 
the vector is combined from 6 individual strings to form a vector of length 
6. Combining strings into a vector does not paste the strings together -- it
forms a vector, whose elements are the individual strings.

```{r character_vector}
character_vector <- c("A", "vector", "of", "length", 6, ".")
character_vector
typeof(character_vector)
```

The type of the vector is character. Elements of a vector of type character 
are strings. Note how the numeric value `6` in the construction of the vector was 
automatically coerced into the string `"6"`. 

It is possible to paste together the strings in a character vector. 

```{r string_paste, dependson="character_vector"}
paste(character_vector, collapse = " ")  # Now a character vector of length 1!
```

It is likewise possible to split a string according to a pattern. For instance 
into its individual characters.

```{r string_split, dependson="character_vector"}
strsplit(character_vector[2], split = "")  # Split "vector" into characters
```

Various additional string operations are available -- see `?character` for 
more information. 

In summary, atomic vectors are the primitive data structures used in R, 
with elements being accessible via indices (random access).
Typical vectors contain numbers, logical values or strings. There is no 
declarations of data types -- they are inferred from data 
or computations. This is a flexible type system with many operations in R 
silently coercing elements in a vector from one type to another. 

### Comparisons and numerical precision

We can compare vectors using the equality operator `==`, which compares
two vectors element-by-element. The result of the following comparison might 
be a surprise at first sight.

```{r double_vector_comparison, dependson = "double_vector"}
double_vector[2:3]
double_vector[2:3] == c(0.2, 0.3)
```

The result of the comparison is a vector of length 2 containing the logical 
values `TRUE FALSE`, but `double_vector[2:3]` appears
to be equal to `c(0.2, 0.3)`. The difference shows up if we increase the number
of printed digits from the default (which is 7) to 20.

```{r double_vector_precision, dependson = "double_vector", echo=2:3}
options(digits = 20)  
c(0.2, 0.3)
double_vector[2:3]
options(digits = 7)  
```

The 0.3 produced by `seq` is computed as `0.1 + 0.1 + 0.1`, while 
the 0.3 in the vector `c(0.2, 0.3)` is converted directly into a double precision
number. The difference arises because neither 0.1 nor 0.3 are exactly representable
in the binary numeral system, and the arithmetic operations induce rounding 
errors. The function `numToBits` can reveal the exact difference in the three 
least significant bits.

```{r bits}
numToBits(0.3)[1:3] 
numToBits(0.1 + 0.1 + 0.1)[1:3] 
```

Differences in the least significant bits are tolerable when we do numerical 
computations but can be a nuisance for equality testing. When comparing vectors 
containing doubles we are therefore often interested in testing for approximate
equality instead using e.g. 

```{r approx_test, dependson="double_vector"}
all.equal(double_vector, c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.0))
```

The function `all.equal` has a tolerance argument controlling if numerical 
differences will be regarded as actual differences. Another way of comparing 
numerical vectors is by computing the range of their difference 

```{r range, dependson="double_vector"}
range(double_vector - c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.0))
```

This shows the largest positive and negative difference. Usually, 
the size of the differences should be assessed relative to the magnitudes of 
the numbers compared. For 
numbers approximately of magnitude 1, differences 
of the order $2^{-52} \simeq 2.22 \times 10^{-16}$ (the "machine epsilon") can 
safely be regarded as rounding errors. The default (relative) tolerance 
of `all.equal` is the more permissive number $\sqrt{2^{-53}} \simeq 1.5 \times 10^{-8}$. 
This number is an *ad hoc* but commonly used tolerance. See `?all.equal` for more details. 

### Lists

### Other data structures

Factors, dates, data frames

## Functions

When you write an R script and source it, the different R expressions in the 
script are evaluated sequentially. Intermediate results may be stored in variables
and used further down in the script. It may not be obvious, but all the expressions 
in the script are, in fact, function calls. Your script relies on functions 
implemented in either core R packages or in other installed packages. 

It is possible to use R as a scripting language without ever writing your own 
functions, but writing new R functions is how you extend the language, and 
it is how you modularize and reuse your code in an efficient way. As a first 
example, we implement the Gaussian kernel with bandwidth $h$ (see Section \@ref(kernel-density))
$$K_h(x) = \frac{1}{h \sqrt{2 \pi}} e^{- \frac{x^2}{2h^2}}.$$
We call the function `gauss()` to remind us that this is the Gaussian kernel.

```{r gauss-kernel}
gauss <- function(x, h = 1) {
  exp(- x^2 / (2 * h^2)) / (h * sqrt(2 * pi))
}
```

The *body* of the function is the R expression `exp(- x^2 / (2 * h^2)) / (h * sqrt(2 * pi))`. 
When the function is called with specific numeric values of its two *formal arguments* 
`x` and `h`, the body is evaluated with the formal arguments replaced by their 
values. The value of the (last) evaluated expression in the body is returned by 
the function. The bandwidth argument `h` is given a default value 1,
so if that argument is not specified when the function is called, it gets the 
value 1. 

The Gaussian kernel with bandwidth $h$ is the Gaussian density with standard 
deviation $h$. Thus we can compare our implementation with `dnorm()` in R 
that computes the Gaussian density.

```{r gauss-kernel-test, dependson="gauss-kernel"}
c(gauss(1), dnorm(1))
c(gauss(0.1, 0.1), dnorm(0.1, sd = 0.1))  # Bandwidth changed to 0.1
```

For those two cases the functions compute the same (at least, up to the printed 
precision). Note how the formal argument `sd` is given the value `0.1` in 
the second call of `dnorm()`. The argument `sd` is the third argument of `dnorm()`,
but we don't need to specify the second, as in `dnorm(0.1, 0, 0.1)`, to 
specify the third. We can do so by its name, in which case the second argument
in `dnorm()` gets its default value `0`. 

There are several alternative ways to implement the Gaussian kernel that 
illustrate how functions can be written in R.

```{r gauss-kernel-alt}
# A one-liner without curly brackets
gauss_one_liner <- function(x, h = 1) 
  exp(- x^2 / (2 * h^2)) / (h * sqrt(2 * pi)) 

# A stepwise implementation computing the exponent first 
gauss_step <- function(x, h = 1) {
  exponent <- (x / h)^2 / 2
  exp(- exponent) / (h * sqrt(2 * pi))
}

# A stepwise implementation with an explicit return statement
gauss_step_return <- function(x, h = 1) {
  exponent <- (x / h)^2 / 2
  value <- exp(- exponent) / (h * sqrt(2 * pi))
  return(value)
}
```

The following small test shows two cases where all implementations compute the 
same.

```{r gauss-kernel-alt-test, dependson=c("gauss-kernel", "gauss-kernel-alt")}
c(gauss(1), gauss_one_liner(1), gauss_step(1), gauss_step_return(1))
c(
  gauss(0.1, 0.1), 
  gauss_one_liner(0.1, 0.1), 
  gauss_step(0.1, 0.1), 
  gauss_step_return(0.1, 0.1)
)
```

A function should always be tested. A test is a comparison of the return value 
of the function with the expected value for some specific argument(s). The 
expected value can either be computed by hand or by another implementation -- 
as in the comparisons of `gauss()` and `dnorm()`. Such tests cannot prove 
that the implementation is correct, but discrepancies can help you to catch and 
correct bugs. Remember that when testing numerical computations that use floating 
point numbers we cannot expect exact equality. Thus tests should reveal 
if return values are within an acceptable tolerance of the expected results. 

Larger pieces of software -- such as an entire R package -- should include 
a number of tests of each function it implements. This is known as 
*unit testing* (each unit, that is, each function, is tested), and there are 
[packages supporting](https://r-pkgs.org/tests.html) the systematic development 
of unit tests for R package development. A comprehensive set of unit tests 
also helps when functions are rewritten to e.g. improve performance or extend 
functionality. If the rewritten function passes all tests, chances are that we 
didn't break anything by rewriting the function.

It is good practice to write functions that do one well defined computation
and to keep the body of a function relatively small. Then it is easier to reason 
about what the function does and it is easier to comprehensively test it. 
Complex behavior is achieved by composing small and well tested functions. 

To summarize:

* The body is enclosed by curly brackets `{}`, which may be left out if the 
body is only one line.
* The function returns the value of the last expression in the body except when 
the body contains an explicit return statement.
* Formal arguments can be given default values when the function is implemented,
and arguments can be passed to the function by position as well as by name. 
* Functions should do a single well defined computation and be well tested.

### Vectorization

In Section \@ref(app-data-structures) it was shown how comparison operators 
work in a vectorized way. In R, comparing a vector to another vector or a 
number leads to element-by-element comparisons with a logical vector as the result.
This is one example of how many operations and function evaluations in R are
natively vectorized, which means that when the function is evaluated with a 
vector argument the function body is effectively evaluated for each entry in 
the vector.

Our `gauss()` function is another example of a function that automatically 
works as a vectorized function.

```{r gauss-vector}
gauss(c(1, 0.1), c(1, 0.1))
```

It works as expected for the vector input because all the functions in the body 
of `gauss()` are vectorized, that is, the arithmetic operators are vectorized, 
the square root is vectorized and the exponential function is vectorized. 

It is good practice to write R programs that use vectorized computations whenever
possible. The alternative for-loop can be much slower. Several examples in the 
book illustrate the computational benefits of vectorized implementations. It may,
however, not always be obvious how to correctly implement a vectorized function.

Suppose we want to implement the following function 
$$\overline{f}_h(x) =  \frac{1}{n} \sum_{i=1}^n K_h(x - x_i)$$
for a data set $x_1, \ldots, x_n$ and $K_h$ the Gaussian kernel. This is the 
Gaussian kernel density estimator considered in Section \@ref(likelihood). A 
straight forward implementation is 

```{r gauss-estimator}
xs <- rnorm(10)  # A data set with 10 observations
f_bar <- function(x, h) 
  mean(gauss(x - xs, h))
```

This implementation works correctly when `x` and `h` are single numbers
but when e.g. `x` is a vector, the function returns a number (not a vector)
unrelated to $\overline{f}_h(0)$ and $\overline{f}_h(1)$. 

```{r gauss-estimator-x-test, dependson="gauss-estimator"}
c(f_bar(0, 1), f_bar(1, 1))
f_bar(c(0, 1), 1) # Computation is not vectorized
```

A quick fix is the following explicit vectorization

```{r}
f_bar_vec <- Vectorize(f_bar)
```

The R function `Vectorize()` is a 
[function operator](https://adv-r.hadley.nz/function-operators.html), which 
takes a function as argument and returns a function. In this case a vectorized
version of the input function. That is, `f_bar_vec()` can be applied to a vector,
which results in applying `f_bar()` to each element of the vector. We test that `f_bar_vec()`
works correctly -- both when the first and the second argument is given as 
a vector. 

```{r gauss-estimator-x-test2, dependson="gauss-estimator"}
f_bar_vec(c(0, 1), 1)
c(f_bar(0, 1), f_bar(1, 1)) # Same result as the vectorized computation
```

```{r gauss-estimator-h-test, dependson="gauss-estimator"}
c(f_bar(1, 1), f_bar(1, 0.1))
f_bar_vec(1, c(1, 0.1)) # Same result as the vectorized computation
```

The function `Vectorize()` basically works by looping over 
the elements in the vector argument(s) and applying the function to each element. 
For prototyping and quick implementations it can be very convenient,
but it is not a shortcut to efficient vectorized computations. 

`Vectorize()` is used in Section \@ref(likelihood) to implement $\overline{f}_h$ based on 
`dnorm()`. The purpose of that implementation is to be able to compute 
$\overline{f}_h(x)$ for arbitrary $x$ in a vectorized way, so that the function 
can be used together with `curve()` and in the likelihood computations of 
that section. The implementations in Section \@ref(dens-implement) differ by 
computing and returning
$$\overline{f}_h(\tilde{x}_1), \ldots, \overline{f}_h(\tilde{x}_m).$$ 
That is, those implementations return the evaluations of $\overline{f}_h$ in 
an equidistant grid $\tilde{x}_1, \ldots, \tilde{x}_m$ and not $\overline{f}_h$
itself.

### Environments

Something

Use `xs` above as example. 

### Function factories

Something

The random number streams from Chap. 4. Include implementation in package.

## Performance

Something

Mention parallel computations

### Tracing

### Rcpp

Use the mean_numeric and mean_complex examples.

## Objects and methods

Something

## Exercises {#app-ex}

### Functions {-}

```{exercise, numerics}
Explain the result of evaluating the following R expression. 
```

```{r ex-numerics}
(0.1 + 0.1 + 0.1) > 0.3

```


```{exercise, filter-function}
Write a function that takes a numeric vector `x` and a threshold value `h` 
as arguments and returns the vector of all values in `x` greater than `h`.
Test the function on `seq(0, 1, 0.1)` with threshold 0.3. Have the example 
from Exercise \@ref(exr:numerics) in mind.

```
  
  
```{exercise, special-values}
Investigate how your function from Exercise \@ref(exr:filter-function) 
treats missing values (`NA`), infinite values
(`Inf` and `-Inf`) and the special value "Not a Number" (`NaN`). Rewrite your
function (if necessary) to exclude all or some of such values from `x`.

*Hint: The functions `is.na`, `is.nan` and `is.finite` are useful.* 

```


### Histograms with non-equidistant breaks {-}

The following three exercises will use a data set consisting of measurements 
of infrared emissions from objects outside of our galax. We will focus on the variable 
F12, which is the total 12 micron band flux density.


```{r infrared-read}
infrared <- read.table("data/infrared.txt", header = TRUE)
F12 <- infrared$F12
```


The purpose of this exercise is two-fold. First, you will get familiar with the 
data and see how different choices of visualizations using histograms can affect 
your interpretation of the data. Second, you will learn more about how to write 
functions in R and gain a better understanding of how they work.


```{exercise, plotHist}
Plot a histogram of `log(F12)` using the default value of the argument `breaks`. Experiment with alternative values of `breaks`. 

```


```{exercise, my-breaks}
Write your own function, called `my_breaks`, which takes two arguments, `x` (a vector) and `h` (a positive integer). Let `h` have default value `5`. The function should first sort
`x` into increasing order and then return the vector that: starts with the smallest entry in `x`;
contains every $h$th unique entry from the sorted `x`; ends with the largest entry in `x`.  

For example, if `h = 2` and `x = c(1, 3, 2, 5, 10, 11, 1, 1, 3)` the function should return `c(1, 3, 10, 11)`. To see this, first sort `x`, which gives the vector `c(1, 1, 1, 2, 3, 3, 5, 10, 11)`, whose unique 
values are `c(1, 2, 3, 5, 10, 11)`. Every second unique entry is `c(1, 3, 10)`, and then the largest entry `11` is concatenated.

*Hint: The functions `sort` and `unique` can be useful.* 
  
Use your function to construct *breakpoints* for the histogram for different values of `h`, and compare with the histograms obtained in Exercise \@ref(exr:plotHist).  

```


```{exercise, my-breaks-advanced}
If there are no ties in the data set, the function above will produce breakpoints 
with `h` observations in the interval between two consecutive breakpoints 
(except the last two perhaps). If there are ties, the function will by construction 
return unique breakpoints, but there may be 
more than `h` observations in some intervals. 

*The intention is now to rewrite `my_breaks` so that if possible each interval 
contains `h` observations.*

Modify the `my_breaks` function with this intention and so that is has the 
following properties:

* All breakpoints must be unique.
* The range of the breakpoints must cover the range of `x`.
* For two subsequent breakpoints, $a$ and $b$, there must be at least `h` observations in the interval $(a,b],$ provided `h < length(x)`. (With the exception that for the first two breakpoints, the interval is $[a,b].$)

```

### Functions and objects {-}

The following exercises build on having implemented a function that 
computes breakpoints for a histogram either as in 
Exercise \@ref(exr:my-breaks) or as in Exercise \@ref(exr:my-breaks-advanced).

```{exercise, my-hist}
Write a function called `my_hist`, which takes a single argument `h` and plots a 
histogram of `log(F12)`. Extend 
the implementation so that any additional argument specified when calling `my_hist` 
is passed on to `hist`. Investigate and explain what happens when executing 
the following function calls.
```

```{r, eval = FALSE}
my_hist()
my_hist(h = 5, freq = TRUE)
my_hist(h = 0)
```

```{exercise, my-hist-mod}
Modify your `my_hist` function so that it returns an object of class `my_histogram`, 
which is not plotted. Write a print method for objects of this class, 
which prints just the number of cells. 

*Hint: It can be useful to know about the function `cat`.*

How can you assign a class label to the returned object so that it is printed using your new print method, but it is still plotted as a histogram when given as argument to `plot`?
  
```

```{exercise, my-hist-summary}
Write a `summary` method that returns a data frame with two columns containing the midpoints of the cells and the counts. 
```

```{exercise, my-hist-ggplot2}
Write a new `plot` method for objects of class `my_histogram` that uses `ggplot2` for plotting the histogram.
``` 

### Functions and environments {-}

The following exercises assume that you have implemented a `my_hist` function
as in Exercise \@ref(exr:my-hist).

```{exercise, my-hist-env}
What happens if you remove that data and call `my_hist` subsequently? 
What is the environment of `my_hist`? Change it to a new environment, and assign 
(using the function `assign`) the data to a 
variable with an appropriate name in that environment. Once this is done, 
check what now happens when calling `my_hist` after
the data is removed from the global environment. 

```

```{exercise, my-hist-funfact}
Write a function that takes an argument `x` (the data) and 
returns a function, where the returned function 
takes an argument `h` (just as `my_hist`) and plots a histogram (just as `my_hist`). 
Because the return value is a function, we may refer to the function 
as a [function factory](https://adv-r.hadley.nz/function-factories.html). 

What is the environment of the function created by the function factory? What 
is in the environment? 
Does it have any effect when calling the function whether the data is altered 
or removed from the global environment? 

```

```{exercise, }
Evaluate the following function call:

```

```{r, eval = FALSE}
tmp <- my_hist(10, plot = FALSE)
```

What is the type and class of `tmp`? What happens when `plot(tmp, col = "red")` 
is executed? How can you find help on what `plot` does with an 
object of this class? Specifically, how do you find the documentation for the 
argument `col`, which is not an argument of `plot`?