--- 
title: "Computational Statistics with R"
author: "Niels Richard Hansen"
date: "`r paste(Sys.Date(), ', Git version: ', system('git rev-parse --short HEAD', intern = TRUE), sep = '')`"
#  pdf_document: default
#  html_document:
#    df_print: paged
description: Lecture notes providing an introduction to computational statistics using the R programming language.
documentclass: book
papersize: a4
fontsize: 11pt
link-citations: yes
colorlinks: yes
urlcolor: green
bibliography: CSwR.bib
site: bookdown::bookdown_site
biblio-style: apalike
---

```{r extra, include=FALSE, cache=FALSE}
source("CSwR_extra.R")
library("igraph")
library("ggplot2")
library("magrittr")
library("CSwR")
library("zeallot")
knitr::opts_chunk$set(cache = TRUE, dev.args = list(bg = 'transparent'), 
                      fig.align = "center", fig.pos="h", cache.lazy = TRUE,
                      out.width = "70%")
theme_replace(plot.background = element_rect(fill = NA, color = NA))
```

```{r error_hook, cache=FALSE, echo=FALSE}
library(knitr)
hook_error = knit_hooks$get('error')
knit_hooks$set(error = function(x, options) {
  x <- strwrap(x)
  hook_error(x, options)
})
```


# Preface {-}

This book is being developed for a graduate level course in computational 
statistics. It is used as the primary material for such a course within the MSc
program in statistics at University of Copenhagen. 

The book assumes a solid mathematical background, where the reader is  
expected to have a reasonable command of mathematical analysis, linear 
algebra and mathematical statistics -- exemplified by maximum likelihood 
estimation of multivariate parameters and asymptotic properties of multivariate
estimators. The reader is also expected to have an understanding of what 
an algorithm is, how numerical computations differ from symbolic computations, 
and to be able to write small computer programs. 

The material covered is not supposed to be a comprehensive treatment of 
computational statistics. It is intended to be a pedagogical introduction to 
some core aspects of computational statistics that bridges the gap between 
theory and implementation. The presentation is driven by a selection of 
statistical examples and their computational challenges. The examples are 
tied together by practical and experimental approaches to solving these 
computational challenges. 

Contemporary research in computational statistics revolves around 
large scale computations, either because the amount of data is massive or 
because we want to apply ever more complicated and sophisticated models 
and methods for the analysis and visualization of data. Compared to these 
research challenges, the examples treated in this book are of a modest 
complexity. They serve as a means to learn the necessary computational 
craftsmanship that is needed when more complex problems are to be solved. 

The book is based on R for several reasons. First of all,
the target audience of statisticians is expected to be familiar with R, and 
they should learn how to use their programming language in an optimal way. 
This includes knowledge of the infrastructure offered by R and RStudio that 
supports good software development. In addition, this infrastructure combined 
with R Markdown and bookdown makes it a bliss to write a book that systematically 
integrates code and software development with the theory. Finally, it is possible 
to write performant R code by a proper use of R as a high-level programming 
language or by interfacing compiled code via the Rcpp package. Statisticians who 
program in R should master these skills.











