---
title: "Univariate random variables: Exercise solutions"
author: "Niels Richard Hansen"
date: "September 15, 2019"
output:
  html_document: 
    code_folding: show 
---


# 4.1

# 4.2


TURN THIS INTRO AN EXERCISE IN CHAP. 5.

We can then reimplement `vMsim` using `rng_stream`. For later usage we add the 
possibility of printing out some tracing information.

```{r vM-rejection-fast, dependson="RNG-stream"}
vMsim <- function(n, kappa) {
  count <- 0
  y <- numeric(n)
  y0 <- runif_stream(n, - pi, pi)
  u <- runif_stream(n)
  for(i in 1:n) {
    reject <- TRUE
    while(reject) {
      count <- count + 1
      z <- y0()
      reject <- u() > exp(kappa * (cos(z) - 1))
    }
    y[i] <- z
  }
 y
}
```

Then we can compare the run time of this new implementation to the 
run time of the first implementation. 

```{r vMsim-fast-runtime, dependson="vM-rejection-fast", echo=2}
tmp <- vMsim(100, kappa = 5)
system.time(vMsim(100000, kappa = 5))
```

As we see from the time estimate above, using a vectorized call of `runif` 
reduces the run time by a factor 4-5. It is possible to get a further factor 2-3 
run time improvement (not shown) by implementing the computations done by
`rng_stream` directly inside `vMsim`. However, we prioritize here to have 
modular code so that we can reuse `rng_stream` for other rejection samplers without 
repeating code. A pure R implementation based on a loop will never be able to 
compete with a C++ implementation anyway when the accept-reject step is such a
simple computation.