---
title: "Exercises 1.11—1.13 <br> solution"
author: "Niels Richard Hansen"
date: "September 3, 2020"
output:
  xaringan::moon_reader:
    css: ["default", "science.css"]
    nature:
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:10'
      navigation:
        scroll: false
---


---
## Problem 1.8


```{r}
rm(F12)
```



```{r, error = TRUE}
my_hist(30)
```


```{r}
environment(myHist)
environment(myHist) <- new.env()
assign("F12", infrared$F12, environment(myHist))
```

--

```{r}
ls()
```

--

```{r}
ls(environment(myHist))
```

---
## Problem 1.8

```{r, error=TRUE}
myHist(40)
```

---
## Problem 1.9 

```{r}
histFactory <- function(x) {
  function(h, ...) 
     hist(x, myBreaks(x, h), ...)
}
F12 <- infrared$F12
myHist2 <- histFactory(log(F12))
```

---
## Problem 1.9

```{r}
myHist2(40)
```

---
## Problem 1.9

```{r}
myHist2 <- histFactory(log(F12))
rm(F12)
ls()
ls(environment(myHist2))
```

---
## Problem 1.9

```{r, error=TRUE}
myHist2(40)
```

--

Perhaps surprisingly, a reference to `F12` was still around, and removing `F12` from 
the global environment resulted in an error.

--

```{r}
substitute(x, environment(myHist2))
```

The explantion is that the R expression `log(F12)` was never evaluated. (It
has nothing to do with the log-transformation.)

---
## Problem 1.9 tricky point

```{r}
histFactory <- function(x) {
  ## force evaluation of argument `x`
  force(x)  #<<
  function(h, ...) 
     hist(x, myBreaks(x, h), ...)
}
F12 <- infrared$F12
myHist2 <- histFactory(log(F12))
```

Due to *lazy evaluation* arguments are not evaluated until used. 
--


The value of `F12` is, without `force(x)` as above, not looked up until the first call of `myHist2`. 
--


If `F12` is removed before the first call of `myHist2`, the call will result in an error!

---
## Problem 1.9

```{r}
rm(F12)
myHist2(40)
```


---
## Summary

* A function can look up variables in its *enclosing environment*. 
--

* The enclosing environment is suitable for storing *local variables*. 
--

* The default enclosing environment is where the function is defined. 
--

* The global environment (the workspace) is not suitable for *local variables*, and dependence 
upon *global variables* should be avoided.
--

* The enclosing environment can be set and populated manually (using `new.env` and `assign`).
--

* A *function factory* like `histFactory` is a systematic way of creating enclosing environments 
for local variables.

---
## Problem 1.10

```{r}
tmp <- myHist(10, plot = FALSE)
typeof(tmp)
class(tmp)
```

---
## Problem 1.10, internal structure

```{r}
## 'str' gives an overview of the internal structure of an R object 
str(tmp) 
```

---
## Problem 1.10, plotting

```{r}
plot(tmp, col = "red") 
```

---
## Problem 1.10, getting help 

You can find documentation for `plot` using e.g. 

```{r}
?plot
```

However, this will be uninformative on how an object of class histogram is plotted. Try instead 

```{r}
?plot.histogram
```

This will give the documentation for the plot method for objects of class histogram. 

